\chapter{Hazardous Substances and COSHH Risk Assessment}
\label{chap:Hazardous Substances and COSHH Risk Assessment}

%---------------------------- SECTION ----------------------------
\section{Why this assessment matters in construction}

\paragraph{}
The construction industry routinely employs some of the most chemically aggressive substances encountered in any working environment. Cement dust, isocyanate-containing spray foams, epoxy resin systems, solvent-based adhesives, and a broad spectrum of coating materials are applied daily on sites ranging from basement waterproofing operations to high-rise façade restoration. Unlike many industrial sectors, construction combines large numbers of operatives, transient and often uncontrolled working environments, variable weather conditions, and a complex supply chain in which substances procured at short notice may arrive on site without adequate accompanying information. The combination of these factors creates a systemic risk of occupational ill-health that dwarfs the incidence of acute injuries in terms of both prevalence and long-term human cost. The Health and Safety Executive estimates that occupational lung disease and occupational skin disease account for thousands of new cases attributable to the construction industry each year, with many resulting in permanent disability and early retirement. The financial burden borne by employers, the NHS, and the social security system is considerable; the burden borne by affected individuals and their families is immeasurable.

\paragraph{}
Occupational ill-health from hazardous substances is characterised by an insidious latency period during which damage accumulates imperceptibly before clinical symptoms emerge. A plasterer who works with wet cement for a decade may not present with contact dermatitis until the tenth year of exposure; an insulation installer repeatedly exposed to isocyanate aerosols during spray polyurethane foam application may develop occupational asthma after months or years of apparently asymptomatic exposure. Once sensitisation to an isocyanate or an epoxy hardener has occurred, the threshold for triggering an asthmatic response or a severe eczematous reaction falls dramatically — exposure at concentrations orders of magnitude below the workplace exposure limit may be sufficient to provoke a life-threatening bronchoconstriction. This irreversibility distinguishes sensitisation from simple irritation and places a categorical duty on employers to prevent sensitisation occurring in the first place, rather than merely to manage symptoms once they appear. A systematic, well-documented Control of Substances Hazardous to Health (COSHH) assessment is the primary mechanism through which that prevention duty is discharged.

\paragraph{}
Skin disease and respiratory disease attributable to work are prescribed occupational diseases under Schedule 1 to the Social Security (Industrial Injuries)(Prescribed Diseases) Regulations 1985 and the Social Security Act 1992. Chromate-induced contact dermatitis caused by cement is prescribed disease D5; occupational asthma caused by isocyanates is prescribed disease D7. Prescription confers a legal right to industrial injuries disablement benefit and triggers parallel reporting duties under the Reporting of Injuries, Diseases and Dangerous Occurrences Regulations 2013 (RIDDOR), which requires employers to report diagnoses of work-related diseases to the HSE using form F2508A. Failure to report constitutes a criminal offence. The disciplinary and financial consequences of allowing preventable prescribed diseases to develop on a construction project are significant: civil claims for damages in cases of occupational asthma attributable to isocyanate exposure regularly result in awards exceeding six figures, and enforcement action by HSE inspectors following a diagnosis of a COSHH-linked prescribed disease may include improvement and prohibition notices, prosecution, and substantial fines assessed under the Sentencing Council Guideline for Health and Safety Offences.

\paragraph{}
The COSHH assessment is not a bureaucratic formality to be completed once and filed away; it is a living document that must respond to changes in working methods, product reformulation, workforce composition, and the emergence of new toxicological evidence. The introduction of reduced-chromate cement requirements under the Supply of Machinery Safety Regulations and the reclassification of substances under the Globally Harmonised System (GHS) as embedded in the Classification, Labelling and Packaging (CLP) Regulation are examples of regulatory and scientific developments that can invalidate a previously adequate COSHH assessment overnight. The assessor must maintain a proactive relationship with product Safety Data Sheets (SDS), with the HSE's EH40 workplace exposure limits document (4th edition, 2020), and with emerging guidance from bodies such as the British Occupational Hygiene Society (BOHS) and the Association of Occupational and Environmental Lung Physicians. Chapter~\ref{chap:Silica, Wood Dust and Hazardous Dust Risk Assessment} addressed the specific challenges of mineral and organic dust exposure in construction; this chapter extends that analysis to the full range of chemical, biological, and sensitising hazards present on a modern construction site.

\barquote{Every construction worker who develops occupational asthma from isocyanate exposure, or irreversible chromate sensitisation from cement contact, has in effect suffered a permanent occupational injury. The COSHH assessment is not optional risk mitigation — it is the legal and moral instrument through which those injuries are prevented. An assessment that is generic, out of date, or disconnected from actual site practice is no assessment at all.}

\example{Isocyanate Exposure During Waterproofing Application}{A roofing subcontractor is engaged to apply a two-component spray polyurethane foam (SPF) waterproofing system to the flat roof of a new-build residential block. The operatives are equipped with half-face respirators fitted with combined A1P3 filter cartridges. The principal contractor's site manager reviews the COSHH assessment submitted by the subcontractor and notes that the SDS for the isocyanate component (Component B, MDI-based) specifies that spray application requires a supplied-air breathing apparatus (SABA) or self-contained breathing apparatus (SCBA) operating in positive-pressure mode — not a filtering facepiece or cartridge respirator. The half-face RPE specified in the subcontractor's COSHH assessment is wholly inadequate for spray application because isocyanate aerosol concentrations during spraying routinely exceed the assigned protection factor of any cartridge-based device. The principal contractor issues a stop-work notice under CDM 2015 and requires the subcontractor to revise the COSHH assessment, procure suitable supplied-air equipment, and confirm that all operatives have received isocyanate-specific health surveillance including pre-placement lung function testing. Work recommences only when the revised assessment has been reviewed, operatives have been medically cleared, and a dedicated COSHH supervisor has been assigned to monitor exposure controls throughout the application phase.}

%---------------------------- SECTION ----------------------------
\section{Legal and professional framework}

The COSHH assessment sits within a dense statutory and quasi-statutory framework that encompasses primary legislation, statutory instruments, approved codes of practice, HSE guidance, and European-derived technical standards. The assessor who approaches a COSHH assessment in isolation from this framework risks producing a document that is legally inadequate, operationally ineffective, and evidentially deficient in the event of enforcement action or civil litigation. A thorough understanding of the legislative architecture is therefore an operational prerequisite, not merely academic background knowledge. The assessment must be capable of demonstrating, to a competent occupational hygienist or HSE inspector, that the responsible person has identified every relevant hazardous substance in use, evaluated the routes and magnitude of likely exposure, selected controls from the correct point in the hierarchy, and put in place appropriate monitoring and health surveillance arrangements proportionate to the nature and severity of the hazards involved.

The key statutory and guidance instruments governing COSHH assessments in construction are set out below. These instruments do not operate in isolation: the principal contractor's COSHH obligations under CDM 2015 interact with the employer's duties under the Management Regulations and with product-specific obligations arising from REACH and CLP. The assessor must be capable of navigating all of these instruments simultaneously and of recognising where duties overlap or reinforce one another. A failure to comply with any of the instruments listed below may constitute a criminal offence under the Health and Safety at Work etc.\ Act 1974 (HSWA), which provides the overarching enforcement backstop for the entire regulatory framework governing workplace hazardous substances.

\begin{itemize}

\bitem{Health and Safety at Work etc.\ Act 1974}{The primary enabling legislation for workplace health and safety in Great Britain. Section~2(1) imposes a general duty on every employer to ensure, so far as is reasonably practicable, the health, safety and welfare at work of all employees. Section~2(2)(a) specifies the duty to provide and maintain safe plant and systems of work; Section~2(2)(b) requires safe arrangements for the use, handling, storage, and transport of articles and substances. Section~3(1) extends the general duty to persons not in employment who may be affected by the employer's undertaking — critical in construction where members of the public, neighbouring occupiers, and workers from multiple employers share a site. Contravention of HSWA duties is a criminal offence triable summarily or on indictment, with unlimited fines following the Legal Aid, Sentencing and Punishment of Offenders Act 2012.}

\bitem{Management of Health and Safety at Work Regulations 1999}{Regulation~3 requires every employer to make a suitable and sufficient assessment of risks to the health and safety of employees and others. For COSHH substances, the COSHH assessment supplements but does not replace the general risk assessment under Regulation~3. Regulation~5 requires employers to implement preventive and protective measures; Regulation~7 mandates the appointment of competent persons to assist in implementing those measures. The ``suitable and sufficient'' standard is the same standard that applies to COSHH assessments and has been interpreted by the courts to require both completeness (covering all foreseeable risks) and proportionality (depth of assessment commensurate with the nature and severity of the hazard).}

\bitem{Control of Substances Hazardous to Health Regulations 2002 (COSHH)}{The central instrument governing chemical, biological, and sensitising hazard management. Regulation~6 requires a suitable and sufficient assessment of health risks before work with hazardous substances commences; the assessment must be reviewed if there is reason to suspect it is no longer valid or following significant change. Regulation~7 imposes a hierarchy of control: prevention of exposure is preferred; where prevention is not reasonably practicable, adequate control must be achieved through engineering controls, process design, and enclosure before recourse to PPE. Regulation~8 places a duty on employees to use control measures provided. Regulation~9 requires employers to maintain, examine, and test engineering controls and LEV systems at specified intervals (at least every 14 months for LEV). Regulation~10 requires monitoring of workplace exposure where assessment shows a risk to health or a statutory limit applies. Regulation~11 requires appropriate health surveillance where the assessment shows a risk of identifiable disease or adverse health effect; records must be kept for 40 years for carcinogens and 10 years for other hazardous substances. Regulation~12 requires employers to provide employees with suitable information, instruction, and training about the substances they work with, the risks involved, and the precautions to be taken.}

\bitem{EH40/2005 Workplace Exposure Limits (4th Edition, 2020)}{Approved by the HSC under Regulation~7(7) of COSHH 2002, EH40 sets statutory Workplace Exposure Limits (WELs) for approximately 500 substances. WELs are expressed as an 8-hour time-weighted average (TWA) — the long-term exposure standard (LTEL) — and as a 15-minute TWA — the short-term exposure standard (STEL). The 4th edition (2020) incorporates reductions to a number of priority limits aligned with European OELs under Directive 2017/164/EU and its amending directives. For isocyanates, the WEL (as NCO) is 0.02~mg/m\textsuperscript{3} TWA and 0.07~mg/m\textsuperscript{3} STEL; for toluene, the WEL is 50~ppm TWA / 100~ppm STEL; for xylene (all isomers), 50~ppm TWA / 100~ppm STEL; for methyl ethyl ketone (MEK, butan-2-one), 200~ppm TWA / 300~ppm STEL. WELs represent maximum permissible concentrations, not safe levels — the duty under Regulation~7 is to reduce exposure as far below the WEL as is reasonably practicable.}

\bitem{REACH Regulation (EC) No 1907/2006}{The Registration, Evaluation, Authorisation and Restriction of Chemicals regulation, retained in UK law post-Brexit as UK REACH under the Environment Act 2021, places obligations on manufacturers and importers to register substances and to communicate hazard information down the supply chain via Safety Data Sheets (SDS). Article~31 of UK REACH requires suppliers to provide an SDS in the format specified by Annex II for any substance or mixture classified as hazardous under CLP, persistent, bioaccumulative, and toxic (PBT), or very persistent and very bioaccumulative (vPvB). The SDS must contain 16 mandatory sections covering identification, hazard identification, composition, first-aid measures, fire-fighting measures, accidental release, handling and storage, exposure controls, physical and chemical properties, stability and reactivity, toxicological information, ecological information, disposal, transport, regulatory information, and other information. The employer has a duty under COSHH Regulation~12 to ensure that current SDS information is available to and understood by the operatives who use the substance.}

\bitem{HSG97 --- A Step by Step Guide to COSHH Assessment}{The HSE's primary practical guidance document for conducting COSHH assessments. It articulates the five-step methodology that forms the operational backbone of a compliant assessment: (1) identify the hazardous substances present; (2) consider who might be harmed and how; (3) evaluate the risks and decide on precautions; (4) record the findings and implement them; (5) review the assessment and update if necessary. HSG97 is not an Approved Code of Practice but its recommendations, if followed, are treated by the courts as strong evidence of compliance with Regulation~6. Departure from HSG97's methodology without substituting an equally effective approach creates significant evidential vulnerability in enforcement or litigation proceedings.}

\bitem{RIDDOR 2013 --- Reporting of Injuries, Diseases and Dangerous Occurrences Regulations}{Schedule~1 (diseases) requires employers to report to the HSE where a doctor confirms that an employee is suffering from a reportable work-related disease. For COSHH purposes, reportable diseases include: occupational asthma (including isocyanate-induced asthma); occupational dermatitis; any occupational cancer; any disease attributed to an occupational exposure to a biological agent; and tendinitis or tenosynovitis of the hand or forearm. Reports must be made via the HSE's online RIDDOR portal within 10 days of the employer receiving the medical diagnosis. Failure to report is a criminal offence. RIDDOR data feeds national statistics on work-related ill-health trends and informs the HSE's targeting of COSHH enforcement programmes.}

\bitem{Personal Protective Equipment at Work Regulations 2022}{Extending the original 1992 Regulations to limb (b) workers, the 2022 Regulations require employers to provide suitable PPE to all workers where risks cannot be adequately controlled by other means. In the COSHH context, PPE — including respiratory protective equipment (RPE), chemical protective gloves, coveralls, and eye protection — is the last resort in the hierarchy of controls. PPE must be suitable for the purpose (CE/UKCA marked), properly maintained, stored correctly, and used in accordance with manufacturer instructions. RPE used for isocyanates during spray application must have an assigned protection factor (APF) sufficient to reduce exposure below the WEL; for concentrations generated during spray polyurethane foam application, this requires a powered air-purifying respirator (PAPR) or a supplied-air device at minimum. Face-fit testing under HSE guidance (INDG479) is mandatory for all tight-fitting facepieces.}

\bitem{Hazardous Waste Regulations 2005 (England and Wales)}{Implement Directive 2008/98/EC and the Hazardous Waste List as codified in the European Waste Catalogue. Solvent-based products, paint waste, isocyanate-containing waste, and epoxy waste are classified as hazardous waste and cannot be disposed of as general construction waste. Producers of hazardous waste must register as hazardous waste producers with the Environment Agency (EA) if they produce more than 500~kg per year; consignment notes must accompany every movement of hazardous waste from point of production to point of disposal or treatment. The waste hierarchy under the Waste (England and Wales) Regulations 2011 requires prevention as the first priority, followed by preparation for re-use, recycling, recovery, and only as a last resort, disposal. Site COSHH management plans must include a waste disposal section addressing segregation, labelling, storage, and consignment of hazardous chemical waste.}

\bitem{Construction (Design and Management) Regulations 2015}{Under CDM 2015, the principal contractor is responsible for coordinating health and safety during the construction phase, which includes managing COSHH risks arising from work by all contractors on the project. Regulation~15 requires the principal contractor to plan, manage, monitor, and coordinate the construction phase. The construction phase plan (CPP) must address the management of substances hazardous to health, including arrangements for obtaining and reviewing SDS documents, arrangements for COSHH assessments from contractors, and arrangements for health surveillance where sensitisers or carcinogens are in use. The principal designer must consider foreseeable COSHH risks arising from design decisions — for example, specifying an isocyanate-cured sealant where an alternative lower-hazard product could be used — and communicate residual risks via the health and safety file.}

\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Conducting the assessment using the five-step method}

\subsection{Step 1: Identify the hazardous substances present}

\paragraph{}
The first step in a compliant COSHH assessment is a comprehensive inventory of all substances hazardous to health that are used, produced, or may be encountered incidentally on the site or in the workplace. The assessor must cast the net broadly: hazardous substances include not only the raw materials deliberately purchased and applied but also the by-products of work processes (silica dust liberated during concrete cutting — addressed in Chapter~\ref{chap:Silica, Wood Dust and Hazardous Dust Risk Assessment}), the decomposition products generated during hot-work operations (isocyanate vapours from the flame cutting of polyurethane foam insulation boards), and any biological agents present in the working environment (Weil's disease-causing \textit{Leptospira} in groundwater encountered during excavation). The primary source of information for each substance is the current Safety Data Sheet supplied by the manufacturer or importer. SDS documents must be in their most recent version — manufacturers frequently update SDS in response to new toxicological data, regulatory reclassification, or changes in product formulation, and an out-of-date SDS may describe outdated WELs, incorrect PPE specifications, or superseded first-aid protocols. The assessor should note the SDS version date and check periodically for updates, particularly for substances recently classified under the GHS system or substances subject to ongoing HSE enforcement campaigns, such as isocyanates and sensitising metalworking fluids.

\paragraph{}
Cement deserves particular attention during the identification step. Ordinary Portland Cement (OPC) and blended cements contain calcium oxide (quicklime, CaO) and calcium hydroxide (slaked lime, Ca(OH)\textsubscript{2}) generated upon hydration. Wet cement has a pH of approximately 12--13 and is capable of causing full-thickness alkali burns with prolonged skin contact — a hazard often underestimated because the initial burning sensation is masked by the progressive destruction of nerve tissue. The most serious occupational health concern associated with cement, however, is the hexavalent chromium (Cr(VI), chromate) content naturally present in raw limestone and concentrated by the cement-manufacturing process. Chromate is a well-characterised skin sensitiser and is responsible for a significant proportion of occupational contact dermatitis in the construction industry, classified as prescribed disease D5 (chrome ulceration of the skin) under Schedule 1 of the Social Security (Industrial Injuries)(Prescribed Diseases) Regulations 1985. Since May 2005, manufacturers have been required to add ferrous sulphate to Portland cement as a reducing agent, converting Cr(VI) to the less hazardous trivalent Cr(III) form; the supply of cement with a soluble Cr(VI) content exceeding 2~mg/kg (dry weight) is prohibited under the Supply of Machinery Safety Regulations and successor instruments. The COSHH assessor must verify that cement supplied to site carries current chromate reduction certification and that the shelf life of bags containing ferrous sulphate has not been exceeded.

\paragraph{}
Solvents used in construction — for thinning coatings, cleaning tools, and degreasing substrates — include a range of substances with both acute and chronic toxicity profiles. Methyl ethyl ketone (MEK, butan-2-one) is widely used as a primer for PVC pipe jointing and as a constituent of contact adhesive thinners; its principal acute hazard is central nervous system (CNS) narcosis at elevated concentrations, and it is a dermal irritant and degreaser at lower levels. Toluene, used in some solvent-based adhesives and rubber-based contact cements, is classified as a reproductive toxicant Category 2 under CLP (harmful to the unborn child) and carries significant neurotoxic and hepatotoxic potential at chronic low-level exposures. Isopropanol (IPA) and xylene are encountered in cleaning agents, surface preparation products, and industrial coatings; xylene is classified as an aspiration hazard and is associated with auditory toxicity as well as CNS depression. For all these substances, the COSHH assessment must document the specific WEL from EH40 (both TWA and STEL), the estimated or measured exposure concentration in the work environment, the control measures in place (local exhaust ventilation, enclosure, dilution ventilation, or RPE), and the basis for concluding that exposure is controlled to a level as far below the WEL as is reasonably practicable.

\paragraph{}
Epoxy resin systems, used extensively in construction for structural adhesive bonding, protective coatings, crack injection, and substrate repair, present a complex two-component hazard profile. The base resin component, typically based on bisphenol A diglycidyl ether (BADGE), is a moderate skin sensitiser and irritant in its uncured state; it becomes harmless once fully cured. The hardener component, which may contain aliphatic amines, cycloaliphatic amines, polyamidoamine, or modified polyamine curing agents, is in many formulations more aggressively sensitising than the resin itself and may also be corrosive, with a pH significantly above 11. Amines are well-recognised causes of allergic contact dermatitis and may also be respiratory sensitisers when aerosolised during spray application. The assessor must consider both components separately, identify all hazardous constituents listed in the SDS (which for a complex epoxy hardener may include several individual substances each with its own classification and WEL), and assess the combined hazard of the mixed product as it is actually applied. Substitution with less hazardous hardener chemistries — such as aliphatic amines rather than aromatic amines, or pre-reacted adducts with lower free amine content — should always be considered as part of the hierarchy of control, in accordance with Regulation~7(1) of COSHH.

\subsection{Step 2: Consider who might be harmed and how}

\paragraph{}
The second step requires the assessor to identify all persons who may be exposed to the hazardous substance and to characterise the exposure routes and patterns relevant to their work activities. On a construction site, the population at risk extends well beyond the operative directly applying the substance. In a basement car park where a two-part epoxy coating is being applied by roller, the coating operative is exposed via skin contact and inhalation; adjacent operatives carrying out second-fix mechanical installations in the same enclosed space are exposed via inhalation to solvent vapours and amine aerosols even though they are not directly involved in the coating application; the supervisor walking through the area to monitor progress receives intermittent inhalation exposure; and groundworkers re-entering the space after the working day may be exposed to residual vapours from a partially cured coating. The assessment must consider all of these exposure scenarios, not merely the primary applicator's exposure profile.

\paragraph{}
Routes of exposure must be considered systematically for each substance. Inhalation is the dominant route for volatile solvents, isocyanate vapours, amine aerosols, and biological agents, but it is not the only route. Skin absorption is significant for many solvents (including toluene, xylene, and IPA — all of which carry HSE skin-notation in EH40, indicating that dermal absorption may contribute materially to systemic body burden), and is the primary route of concern for cement chromate sensitisation. Ingestion risk arises from inadequate hand hygiene before eating, drinking, or smoking and is particularly relevant for lead-based paint removal and biocide-treated timber. Injection risk, while uncommon, can occur when high-pressure spray equipment is used for coating application. The assessor should also note the concept of biological monitoring: for substances with a skin notation or for substances where the relationship between ambient air concentration and internal body burden is well established, urinary or blood biomarker monitoring may be required to verify that the total exposure from all routes is adequately controlled. Urinary toluene and urinary hippuric acid are appropriate biological monitoring indices for toluene exposure; urinary methylenedianiline (MDA) is the monitoring index for 4,4$'$-methylenediphenyl diisocyanate (MDI) exposure; blood lead remains the standard index for inorganic lead.

\subsection{Step 3: Evaluate the risks and decide on precautions}

\paragraph{}
Risk evaluation requires the assessor to bring together the hazard properties of the substance (as derived from the SDS, EH40, and specific toxicological guidance), the estimated or measured exposure concentration, and the adequacy of existing controls to reach a conclusion about the residual risk. Where quantitative exposure data are available from personal air monitoring by an occupational hygienist, these should be compared directly with the relevant WEL. Where monitoring data are not available, the assessor must make a reasoned qualitative or semi-quantitative estimate of exposure using recognised tools — Control Banding approaches (such as COSHH Essentials, now available as the HSE's web-based tool) provide a structured method for stratifying risk according to hazard group (A, B, C, D, E corresponding to broad hazard banding from irritants to carcinogens) and usage quantity and pattern. Control Banding assigns each scenario to one of four generic control approaches: good general ventilation (Band 1), local exhaust ventilation (Band 2), enclosed/contained process (Band 3), or specialist advice from an occupational hygienist (Band 4). Substances classified as sensitisers (Skin Sens 1 or Resp Sens 1 under CLP) are automatically assigned to Band 4 regardless of usage quantity, reflecting the irreversibility of sensitisation and the requirement that control must be effectively absolute — no sensitisation events are acceptable.

\paragraph{}
The distinction between sensitisation and irritation is operationally critical and must be explicitly addressed in the COSHH assessment. Irritation is a reversible, concentration-dependent, non-immunological response to a substance that causes tissue damage at the point of contact — for example, the alkali burns from wet cement or the mucous membrane irritation caused by solvent vapours at concentrations above the STEL. Irritant responses typically resolve once exposure ceases and the tissue heals. Sensitisation, by contrast, is an immunological response in which prior exposure (the primary sensitisation or induction phase) primes the immune system to mount a disproportionately intense reaction upon subsequent re-exposure (the elicitation phase). After sensitisation has occurred, the concentration required to trigger a clinical response may be orders of magnitude lower than the concentration required to induce sensitisation, and the response itself — whether a type IV delayed hypersensitivity reaction producing allergic contact dermatitis, or a type I immediate hypersensitivity reaction producing occupational asthma — may be severe and life-threatening. For practical COSHH assessment purposes, the key implication is that exposure to any concentration of a sensitiser carries risk and that the control objective is prevention of all significant exposure, not merely compliance with the WEL.

\subsection{Step 4: Record the findings and implement them}

\paragraph{}
The COSHH assessment must be recorded in writing when an employer has five or more employees. The assessment record should identify: the substances assessed; their classification and labelling under CLP; the relevant WELs from EH40; the assessed exposure routes; the estimated or measured exposure concentrations; the inherent risk rating prior to controls; the specific control measures selected; the residual risk rating after controls; any health surveillance requirements and the schedule for their implementation; any monitoring requirements; and the date of assessment and date for review. In practice, most construction principal contractors use a standardised COSHH assessment form that prompts the assessor through each of these elements. The form should be completed by a competent person — typically a Chartered Member of IOSH, a registered member of the Faculty of Occupational Medicine, or a Certificated Member of BOHS, or an assessor working under the supervision of such a person. Generic COSHH assessments sourced from product manufacturer websites or industry trade bodies may be used as a starting point but must be reviewed and adapted to the specific site conditions, work patterns, and workforce characteristics of each project.

\paragraph{}
Implementation of the selected controls must follow immediately upon completion of the assessment. Engineering controls — LEV systems, ventilated mixing enclosures, water suppression systems for dusty processes — must be commissioned, tested, and their performance verified before the hazardous work commences. Operatives must receive site-specific COSHH induction covering the substances they will encounter, the health risks, the control measures in place, and the correct use and maintenance of any PPE provided. Written COSHH assessments should be available at the point of use in the language(s) understood by the operatives, and SDS documents should be readily accessible on site — in a site COSHH folder held in the site welfare cabin, accessible in digital form via the contractor's safety management system, or posted in laminated format at the point of use. Health surveillance arrangements — pre-placement skin and lung function assessments for operatives working with sensitisers, and periodic surveillance during the employment — must be arranged and documented before sensitising substances are first used.

\subsection{Step 5: Review the assessment and update if necessary}

\paragraph{}
Regulation~6(3) of COSHH requires an employer to review the assessment if there is reason to suspect it may no longer be valid, or if there has been a significant change in the work to which the assessment relates. On a dynamic construction site, triggers for review are frequent: a contractor substitutes one solvent-based sealant product for another mid-project (different solvent composition, different WEL, possibly a different sensitiser profile); the ventilation system in a basement space is modified or removed to allow structural works to proceed; a new operative joins the team with a previously declared sensitisation to epoxy amines; or an incident investigation reveals that operatives have not been wearing the specified gloves because the specified gloves are incompatible with the particular tool grip required for the task. Each of these events constitutes a material change that may render the existing COSHH assessment invalid and requires a formal review, amendment, and reissue of the assessment documentation. The review date should be recorded on the face of the assessment form, and a reminder system should ensure that assessments are not allowed to become stale.

\example{Five-Step Application --- Two-Part Epoxy Coating in a Confined Basement Car Park}{A specialist coatings contractor is tasked with applying a two-component polyurethane-modified epoxy floor coating system (primer + two seal coats) to the concrete floor of a basement car park during the fit-out phase of a mixed-use development. The car park is approximately 1,200~m\textsuperscript{2} in plan, with a clear ceiling height of 2.4~m and two vehicular entry ramps; there is no permanent mechanical ventilation at the time of application. \textit{Step 1 (Identify substances):} The assessor reviews SDS documents for the epoxy primer (Component A: bisphenol A epoxy resin, Component B: polyamidoamine hardener) and the epoxy topcoat (Component A: bisphenol F/A blend, Component B: cycloaliphatic amine hardener containing isophorone diamine — Resp Sens 1B under CLP). Both hardener components contain substances classified as respiratory sensitisers. The SDS for the topcoat hardener specifies an WEL for isophorone diamine of 0.1~ppm TWA; no STEL is listed. \textit{Step 2 (Who is harmed and how):} Direct exposure to the coating operative (roller application, skin and inhalation routes); adjacent second-fix contractors in the same zone (inhalation); the site manager's daily walkthrough inspection (intermittent inhalation). \textit{Step 3 (Evaluate risks):} The confined space with no permanent ventilation represents a very high inhalation risk; the hardener components are classified as Resp Sens 1B and are therefore Band 4 (occupational hygienist advice required). The assessor commissions a ventilation survey which confirms that temporary forced ventilation delivering a minimum of 10 air changes per hour can be achieved using two 1,500~m\textsuperscript{3}/hr axial fans drawing air through the car park from ramp to ramp. \textit{Step 4 (Record and implement):} Temporary LEV fans are installed and performance-tested; coating operatives are issued with half-face respirators with A1P3 combination cartridges for roller application (sufficient APF given the lower aerosol generation from roller compared to spray); impermeable nitrile gloves (0.5~mm minimum thickness), chemical-resistant overalls, and safety spectacles are required throughout application; all operatives undergo pre-placement spirometry and skin examination; a 4-hourly re-entry interval (RIP) is posted at all car park access points to protect operatives from adjacent trades. \textit{Step 5 (Review):} The COSHH assessment is reviewed mid-project when the site manager proposes switching to a spray application method to accelerate progress; the revised assessment identifies that spray application of the same hardener product generates aerosol concentrations that exceed the APF of cartridge RPE, and the specification is changed to a full-face powered air-purifying respirator (PAPR) with P3 filter for the spray phase.}

%---------------------------- SECTION ----------------------------
\section{Applying the hierarchy of controls}

\paragraph{}
Regulation~7 of COSHH 2002 establishes a hierarchical approach to controlling exposure to hazardous substances that mirrors the broader risk control hierarchy embedded in the Management Regulations and BS ISO 45001. The hierarchy is not merely aspirational guidance — it is a statutory requirement, and any assessment that proceeds directly to reliance on PPE without adequately demonstrating why higher-order controls are not reasonably practicable will be regarded as non-compliant by an HSE inspector. The assessor must work systematically from the top of the hierarchy downwards, documenting at each level the consideration given and the reasons why a higher-order control has been ruled out where it has been. The following hierarchy applies in full to COSHH substance management on construction sites.

\begin{itemize}

\bitem{Elimination}{The most effective and preferred control at the apex of the hierarchy. Elimination means removing the hazardous substance from the work process entirely — for example, designing out the need for a solvent-based adhesive by redesigning the joint detail to use a mechanical fixing; specifying a pre-grouted masonry system that eliminates the need for on-site cement mixing; or selecting a cementitious grout supplied as a pre-mixed paste rather than a dry powder to eliminate inhalation exposure during mixing. On construction projects, elimination opportunities are most frequently available at the design stage, and CDM 2015 places a duty on the principal designer to consider hazardous substance risks during the design development process. Where elimination is not achieved at the design stage, the opportunity may still arise during procurement when equivalent performance products with different hazard profiles are being evaluated.}

\bitem{Substitution}{Where elimination is not reasonably practicable, substitution with a less hazardous substance or form of the same substance should be considered. Substitution does not mean replacing one hazardous substance with another of identical hazard — it means making a genuine improvement in the hazard profile. Practical construction industry examples include: replacing a toluene- or xylene-based contact adhesive with a water-based contact adhesive formulation for bonding insulation boards (eliminates the solvent narcosis and reproductive toxicity risk while achieving comparable bond strength); specifying a chromate-reduced cement (with verified current ferrous sulphate addition and within shelf life) as the baseline procurement standard to minimise dermal sensitisation risk; using pre-mixed mortar delivered in ready-to-use form rather than site-batched dry cement mixes to reduce inhalation and skin contact exposure; selecting an aliphatic isocyanate-based coating (lower vapour pressure, lower respiratory sensitisation risk) over an aromatic MDI- or TDI-based system for indoor spray applications. The SDS for both the current product and the proposed substitute must be reviewed to confirm that the substitution genuinely reduces overall health risk and does not inadvertently introduce a new hazard.}

\bitem{Engineering controls}{Where the substance cannot be eliminated or substituted, engineering controls are the next priority. Engineering controls are physical measures that prevent or reduce exposure without relying on human behaviour: local exhaust ventilation (LEV) captures contaminants at their source before they disperse into the breathing zone; enclosed or segregated processes isolate the hazardous operation from surrounding workers; wet suppression methods (damping down dust, wet-process rather than dry-grinding) reduce airborne dust concentrations; vacuum attachment to power tools removes dust at source in accordance with HSG150 and INDG463. LEV systems used for COSHH control must be designed and commissioned by a competent person (typically a member of BOHS Panel of Competent Persons in Thorough Examination and Testing of LEV), and must be thoroughly examined and tested at intervals not exceeding 14 months in accordance with Regulation~9 of COSHH and the COSHH ACoP. Records of examination and test must be kept for 5 years and must be available for inspection by HSE.}

\bitem{Administrative controls}{Where engineering controls alone cannot achieve adequate control, administrative measures supplement them by managing the duration, frequency, and circumstances of exposure. Job rotation to limit any individual operative's cumulative exposure; permit-to-work systems controlling access to areas where sensitising or carcinogenic substances are in use; timed re-entry intervals after solvent-based coating application; scheduling of hazardous tasks outside normal working hours to reduce the number of persons at risk; pre-task briefings to ensure all operatives understand the specific COSHH controls in place for each day's activities; and the appointment of a dedicated COSHH supervisor to monitor control implementation during high-risk tasks are all examples of administrative controls that may be required to supplement engineering controls for complex, multi-trade construction environments.}

\bitem{Personal protective equipment}{PPE is the control of last resort — used where residual risk remains after all higher-order controls have been applied, or where higher-order controls are transiently unavailable (e.g., emergency maintenance tasks). For COSHH applications, PPE selection must be based on the specific substance and exposure route: chemical-resistant gloves (nitrile, neoprene, or butyl rubber depending on the solvent — breakthrough time data from the glove manufacturer's chemical resistance chart must be consulted); eye and face protection (chemical goggles or full-face shield for corrosive or aerosol-generating tasks); coveralls (Type 5/6 disposable coveralls for powder handling or spray application); and RPE selected to provide an assigned protection factor sufficient to reduce the inhaled concentration below the relevant WEL, with an appropriate margin of safety. All RPE used for substances classified as respiratory sensitisers must be subject to quantitative or qualitative face-fit testing carried out by a competent person, recorded, and repeated if the operative's facial dimensions change (significant weight change, facial surgery, or dental work). Disposable FFP3 filtering facepieces are not appropriate for spray isocyanate work; a PAPR with P3 filter providing an APF of 20 or greater, or a supplied-air device, is required.}

\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Cadence of assessment and review triggers}

\paragraph{}
A COSHH assessment is not a one-time event but an ongoing process of risk management that must remain contemporaneous with actual site conditions. On a linear infrastructure project or a multi-phase building project, the hazardous substances in use change substantially as the project moves through enabling, groundworks, structural frame, façade, and fit-out phases; the workforce composition changes; subcontractors join and leave; and the working environment evolves from open external to enclosed internal conditions that can dramatically alter exposure concentrations. The principal contractor's COSHH management plan should include an explicit review cycle and a mechanism for receiving, reviewing, and recording COSHH assessments from all contractors introducing new substances to site. The following events trigger a mandatory review, in addition to the routine annual (or project-phase) review.

\begin{itemize}

\bitem{New substance introduced to site}{Whenever a contractor proposes to use a substance not previously assessed, a new COSHH assessment must be completed before use commences. This includes product reformulations by the existing supplier — even where the product name is unchanged, a change in SDS version may reflect a change in composition, classification, or recommended controls that renders the previous assessment invalid. The principal contractor's site induction process should require contractors to submit COSHH assessments for all hazardous substances as a condition of site access, and a designated COSHH co-ordinator should review all incoming assessments against the site-specific conditions before approving use.}

\bitem{Case of occupational ill-health arising}{Any confirmed or suspected case of work-related ill-health associated with a COSHH substance — including a reported skin rash, respiratory symptoms, occupational asthma diagnosis, or abnormal biological monitoring result — requires immediate review of the relevant COSHH assessment. The review must determine whether the control measures were adequate and properly implemented, whether the exposure scenario was correctly characterised in the original assessment, and what immediate remedial action is required to prevent further cases. Where a prescribed occupational disease is confirmed, RIDDOR notification must be completed within 10 days of the diagnosis being received by the employer.}

\bitem{Significant change in work method or location}{A change from roller application to spray application; relocation of a task from an open external environment to a confined internal space; an increase in the quantity of substance used; or a change in the duration of the task may all materially alter the exposure profile and require a revised COSHH assessment. The COSHH supervisor should have authority to suspend hazardous work pending review when such changes are proposed.}

\bitem{Engineering control failure or modification}{The failure of an LEV system; the removal of a temporary forced ventilation fan; a breakdown in wet suppression equipment; or a modification to an enclosed process that increases the potential for substance release all require immediate reassessment of the exposure scenario. Emergency PPE use should be implemented as a temporary measure pending restoration or modification of the engineering control.}

\bitem{New toxicological or regulatory information}{The publication of a revised EH40 WEL (as occurred with a number of substances in the 4th edition, 2020); the reclassification of a substance under CLP (for example, the addition of a respiratory sensitiser classification to a previously unclassified product constituent); the issue of an HSE enforcement notice in respect of a substance commonly used in the sector; or the emergence of new epidemiological evidence linking a substance to a previously unrecognised health effect all provide grounds for reviewing existing assessments. The assessor should maintain a watch on HSE alerts, COSHH guidance updates, and BOHS technical guidance bulletins.}

\bitem{Biological monitoring result outside reference range}{Where biological monitoring is in place — for example, urinary toluene monitoring for solvent workers, urinary MDA for MDI-exposed operatives, or urinary chromate for cement workers with persistent dermal exposure — a result exceeding the biological monitoring guidance value (BMGV) published in EH40 indicates that total exposure from all routes is excessive and that the COSHH assessment requires immediate review. The BMGV for toluene is 600~$\mu$mol/mol creatinine (end of shift, end of working week); for MDI-derived MDA, 1~$\mu$mol/mol creatinine. A result above the BMGV does not necessarily indicate that health harm has already occurred, but it does confirm that the current controls are not achieving the required level of exposure reduction and that prompt remedial action is needed.}

\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Common failure modes}

The following failure patterns represent the most frequently encountered deficiencies in COSHH management on construction sites, identified through HSE inspection programmes, enforcement actions, civil litigation case analyses, and occupational health audit findings. Each represents a systemic gap that an effective COSHH management system must explicitly address.

\begin{itemize}

\bitem{SDS not obtained, not reviewed, or outdated}{The single most common COSHH failure observed on site is the absence of a current SDS for a substance in use. Operatives may be using a product for which no SDS has been requested from the supplier; the SDS obtained at procurement may be several years old and may not reflect current CLP classification; or the SDS may be present in the site COSHH folder but in a language that the operative using the substance cannot read. Without a current, understood SDS, the COSHH assessment is built on incomplete information and may specify incorrect PPE, incorrect WELs, or inadequate first-aid responses. The principal contractor's procurement protocol should require that a current SDS (dated within the last three years, or where the substance has been reclassified, within the last 12 months) is obtained for every hazardous substance before delivery to site, and that the SDS is reviewed by the COSHH assessor before the substance is used.}

\bitem{No health surveillance for sensitisers}{Regulation~11 of COSHH requires appropriate health surveillance for work with substances that may cause sensitisation — including isocyanates, epoxy resins and hardeners, chromate-containing cement, and certain solvents with reproductive toxicity. In practice, health surveillance is routinely omitted from COSHH programmes for subcontractors on construction projects, particularly for short-duration specialist works. The consequence is that sensitisation events go undetected, operatives continue to be exposed after sensitisation has occurred (exacerbating their condition and potentially triggering severe asthmatic episodes), and employers miss the opportunity to prevent further cases through investigation and control improvement. Health surveillance records for carcinogens and substances causing sensitisation must be retained for 40 years under COSHH Regulation~11(8).}

\bitem{RPE not face-fit tested}{Many construction site COSHH assessments specify RPE as a control without specifying the type, the required APF, or the face-fit testing requirement. Half-face and full-face respirators are tight-fitting facepieces: if the seal between the facepiece and the wearer's face is inadequate — due to facial hair, facial scarring, or simply an incorrect size selection — the actual protection factor delivered may be a fraction of the APF assigned to the device type. Quantitative face-fit testing (using a PortaCount or equivalent instrument) or qualitative face-fit testing (using bitter/sweet aerosol challenge) is mandatory for all tight-fitting facepieces used for COSHH substances. Records of face-fit testing must be kept and must accompany the operatives who are tested. A COSHH assessment that specifies a tight-fitting RPE but makes no provision for face-fit testing is operationally inadequate and may constitute a breach of the PPE at Work Regulations 2022.}

\bitem{COSHH assessment not updated when product changes}{Construction procurement frequently substitutes one product for an apparently equivalent product mid-project — a different brand of sealant, a reformulated primer, a new adhesive from a different supplier. Each substitution may introduce a different hazard profile: a replacement sealant may contain a different isocyanate prepolymer with a different vapour pressure; a replacement primer may contain a higher proportion of a solvent with a lower WEL than the original product. Without a systematic trigger mechanism to require COSHH reassessment on product substitution, the site may be operating under an assessment that does not describe the substances actually in use. Product substitution must be treated as a formal change management event within the site COSHH management system, with the COSHH co-ordinator receiving notification of every substitution and confirming that the relevant assessment has been reviewed before the replacement product is used.}

\bitem{Biological monitoring not arranged for carcinogens}{For operatives regularly exposed to substances classified as carcinogens (Category 1A or 1B under CLP) — including benzene (present as an impurity in some petroleum-derived solvents), formaldehyde (released from certain adhesives and resins), crystalline silica (see Chapter~\ref{chap:Silica, Wood Dust and Hazardous Dust Risk Assessment}), and certain hardwood and softwood dusts — biological monitoring provides a direct measure of systemic uptake that supplements airborne monitoring. COSHH requires notification to the HSE of work involving certain carcinogens and biological agents (Regulation~14 and Schedule~4). Where biological monitoring is specified as a control under the COSHH assessment, failure to arrange and record the monitoring programme is a direct regulatory breach. Monitoring results outside the BMGV must trigger immediate COSHH review and, where a prescribed occupational disease is subsequently diagnosed, RIDDOR reporting.}

\bitem{Hazardous waste disposed of without consignment notes}{Isocyanate-containing waste (residual MDI components, used application equipment contaminated with isocyanate), solvent-contaminated rags and absorbents, epoxy waste before full cure, and paint waste classified under EWC codes 08~01~11* (waste paint containing organic solvents) and 08~04~09* (waste adhesives and sealants) must be managed as hazardous waste under the Hazardous Waste Regulations 2005. On construction sites, hazardous waste is routinely — and illegally — placed in general skip waste or crushed and mixed with general demolition arisings. Failure to segregate, label, and consign hazardous waste with appropriate Environment Agency consignment notes constitutes an offence under the Environmental Protection Act 1990 and the Hazardous Waste Regulations 2005, and may result in prosecution of both the waste producer and the principal contractor. The site waste management plan (SWMP) and the COSHH management plan must together address hazardous waste classification, segregation, storage in labelled, bunded receptacles, and removal by a licensed waste carrier with consignment notes.}

\end{itemize}

\example{Failure Pattern --- Operative Developing Occupational Asthma from Isocyanate Spray}{During the waterproofing phase of a multi-storey car park renovation project, a specialist subcontractor applies a spray-applied MDI-based polyurethane membrane to the concrete soffit and walls of each level. The subcontractor's COSHH assessment, submitted to the principal contractor at the start of the project, specifies half-face respirators with A1P3 filter cartridges as the RPE control. No health surveillance has been arranged; no face-fit testing records are held for the two operatives undertaking the spray application. After six weeks of application work, one operative begins experiencing nocturnal cough, wheeze, and dyspnoea. He self-medicates with an over-the-counter bronchodilator for two weeks before consulting his GP, who refers him to an occupational physician. Spirometry reveals a 20\% fall in FEV\textsubscript{1} from his pre-employment baseline (which had not been established, as no pre-placement health surveillance had been conducted). Serial peak flow monitoring confirms the pattern of work-related asthma. A diagnosis of MDI-induced occupational asthma is made and prescribed as disease D7. The employer is required to notify HSE under RIDDOR, a Prohibition Notice is served restricting the operative from any further MDI exposure for life, the subcontractor is subject to formal investigation, and a claim for personal injury damages is initiated. Investigation of the failure reveals that: the COSHH assessment did not reflect the correct RPE requirement for spray application; no health surveillance had been implemented; and the MDI concentration in the operative's breathing zone during spray application (estimated retrospectively by simulation) exceeded the WEL by a factor of at least 15 — well beyond the APF of the cartridge respirator specified.}

%---------------------------- CHAPTER SUMMARY ----------------------------
\newpage
\section{Chapter Summary}

\begin{table}[H]
\centering
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm}X}
\toprule
\textbf{Assessment Element} & \textbf{Operational Requirement} \\
\midrule
Legal/Professional Basis & COSHH Regulations 2002 (Regs~6, 7, 8, 9, 10, 11, 12); EH40/2005 4th edition (2020); REACH/UK REACH (EC) No~1907/2006; PPE at Work Regulations 2022; RIDDOR 2013 (Schedule~1 prescribed diseases); Hazardous Waste Regulations 2005; CDM 2015. \\
Method & Apply the full HSG97 five-step process: identify substances; characterise who is harmed and by what route; evaluate risks against WELs and sensitisation thresholds; record findings and implement controls; review on trigger events. Document inherent and residual risk. \\
Controls & Apply the full hierarchy under Regulation~7: elimination, substitution (water-based for solvent-based; reduced-chromate cement; pre-mixed mortar), engineering controls (LEV, enclosure, wet suppression — examined every 14 months), administrative controls, PPE as last resort (face-fit tested tight-fitting RPE; supplied-air for isocyanate spray). \\
Cadence & Pre-commencement assessment; review on product substitution, change of work method, case of occupational ill-health, LEV failure, new regulatory information, or biological monitoring result above BMGV. Annual review as a minimum on long-duration projects. \\
Assurance & Health surveillance under Regulation~11 for sensitisers and carcinogens (records 40 years); biological monitoring where skin notation or multi-route exposure applies; LEV thorough examination records (5 years); RIDDOR reporting of prescribed diseases; waste consignment notes for hazardous waste. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{itemize}

\bitem{Key takeaway 1 --- COSHH is a statutory requirement, not a best-practice option}{The COSHH Regulations 2002 impose specific legal duties on every employer whose workers may be exposed to hazardous substances. A suitable and sufficient assessment under Regulation~6 is a criminal law obligation. Failure to assess, or assessment that does not reflect actual site conditions and substance use, exposes employers to prosecution, unlimited fines, and civil liability. The COSHH assessment must be completed by or under the supervision of a competent person and must be underpinned by current SDS information and current EH40 WEL data.}

\bitem{Key takeaway 2 --- Sensitisation is irreversible and must be prevented absolutely}{Substances that cause skin or respiratory sensitisation — chromate in cement, MDI and TDI isocyanates, epoxy resin hardeners, and certain solvents — require a qualitatively different approach to risk management from simple irritants. Because sensitisation is irreversible and because the elicitation threshold following sensitisation may be vanishingly small, the control objective is the prevention of sensitisation in the first instance. This requires substitution with less sensitising products wherever reasonably practicable, engineering controls to the highest achievable standard, appropriate RPE with verified face-fit, and robust pre-placement and periodic health surveillance.}

\bitem{Key takeaway 3 --- The Safety Data Sheet is the assessment's primary information source}{The SDS under UK REACH must be obtained for every hazardous substance before its first use on site, reviewed by the COSHH assessor, and made available to operatives in a form they can understand. The 16 sections of the SDS contain the substance classification, WEL and BMGV data, recommended engineering controls and PPE, first-aid and emergency procedures, and waste disposal requirements. An assessment built on an outdated, incomplete, or unreviewed SDS is built on inadequate information and cannot be regarded as suitable and sufficient under Regulation~6.}

\bitem{Key takeaway 4 --- Health surveillance is a Regulation~11 obligation for sensitisers and carcinogens}{Where the COSHH assessment identifies a risk of sensitisation, occupational asthma, occupational dermatitis, or cancer-related disease, appropriate health surveillance must be provided. For isocyanate workers, this means pre-placement spirometry and skin examination, repeated at least annually. For epoxy and cement workers, periodic skin examination. For carcinogen-exposed workers, the surveillance regime and records (retained 40 years) must be agreed with an occupational physician. Health surveillance data enables early detection of adverse health effects, prompts immediate reassessment of controls, and provides evidence of compliance or non-compliance in the event of civil litigation.}

\bitem{Key takeaway 5 --- Waste disposal is integral to the COSHH assessment}{The COSHH assessment must address the complete lifecycle of hazardous substances on site, including their disposal as waste. Solvent waste, isocyanate-containing waste, epoxy waste, and paint waste are classified as hazardous waste under EWC and the Hazardous Waste Regulations 2005. Hazardous waste must be segregated, stored in bunded and labelled receptacles, and removed by a licensed waste carrier with an Environment Agency consignment note. Disposal of hazardous waste in general construction skips is an environmental offence and a health and safety failure that exposes workers on subsequent waste-handling tasks to unassessed chemical hazards.}

\end{itemize}

\barquote{In Chapter~\ref{chap:Electrical Safety Risk Assessment}, attention shifts from chemical and biological hazards to the unique risks posed by electrical systems in construction — both the permanent electrical installation under design and the temporary site supply that powers tools, plant, and welfare facilities throughout the construction phase. The assessment of electrical risk requires the same systematic, evidence-based approach that characterises a compliant COSHH assessment: identification of all potential sources of harm, evaluation of the likelihood and severity of harm, selection of controls from the established hierarchy, and verification through inspection, testing, and competence assurance that those controls are effective in practice.}
