\chapter{Silica, Wood Dust and Hazardous Dust Risk Assessment}
\label{chap:Silica, Wood Dust and Hazardous Dust Risk Assessment}

%---------------------------- SECTION ----------------------------
\section{Why this assessment matters in construction}

\paragraph{}
Construction work generates airborne dust as an unavoidable by-product of cutting, grinding, drilling, scabbling, chasing, sweeping, and demolition. Not all dust is equal in its hazard profile: some dusts cause acute irritation and reversible effects; others accumulate progressively in lung tissue, triggering occupational diseases that are irreversible, debilitating, and, in the worst cases, fatal. Respirable crystalline silica (RCS) and hardwood dust represent two of the most clinically significant occupational carcinogens and disease-causing agents encountered in the UK construction industry. The Health and Safety Executive has consistently identified construction as the sector with the highest burden of occupational lung disease in Great Britain, accounting for thousands of new cases of occupational asthma, asbestosis, silicosis, and chronic obstructive pulmonary disease (COPD) each year. The scale of the problem has prompted sustained regulatory attention, campaigns by the British Occupational Hygiene Society (BOHS), and the development of sector-specific guidance. Understanding the risk profile of construction dust is therefore not merely a compliance exercise: it is a precondition for preventing life-altering and life-shortening disease in the workforce.

\paragraph{}
Silica is a naturally occurring mineral found in many common construction materials: concrete, mortar, sandstone, granite, slate, and some engineering bricks all contain crystalline silica in varying concentrations. When these materials are mechanically disturbed --- particularly through dry disc cutting, angle grinding, scabbling, or abrasive blasting --- fine respirable particles, typically less than ten micrometres in aerodynamic diameter, are liberated into the working environment. These respirable particles penetrate beyond the ciliary defences of the upper respiratory tract and deposit in the alveolar region of the lung, where they trigger a persistent inflammatory and fibrotic response. Repeated or high-level exposure leads to silicosis in its classic nodular, accelerated, or progressive massive fibrosis (PMF) forms. Silicosis is a prescribed disease under Social Security (Industrial Injuries) legislation, reflecting the long-standing recognition by the state of its occupational aetiology. There is also strong epidemiological evidence linking RCS exposure to lung cancer, and the International Agency for Research on Cancer (IARC) classifies inhaled crystalline silica from occupational sources as a Group~1 human carcinogen.

\paragraph{}
Wood dust presents a distinct but equally serious hazard profile. Hardwood dusts --- particularly those from oak, beech, iroko, and teak --- are classified as Group~1 carcinogens by IARC, with a well-established causal link to sinonasal cancer, specifically adenocarcinoma of the nasal cavity and paranasal sinuses. The latency period can exceed 30 years, meaning that carpenters, joiners, and site operatives exposed during the 1980s and 1990s may only now be presenting with disease. Softwood dusts, whilst not classified as carcinogens, are capable of causing occupational asthma and sensitisation of the airways. The mechanism of sensitisation involves repeated low-level exposure triggering IgE-mediated or non-IgE-mediated immune responses that, once established, persist indefinitely: a sensitised individual cannot safely return to their previous exposure environment. Wood dust is therefore a significant driver of occupational asthma in the joinery, shopfitting, and timber-frame construction sectors, and the disease it causes is both preventable and irreversible once it has developed.

\paragraph{}
The construction sector's particular challenge is that hazardous dust exposure is seldom confined to a single trade or activity: a groundworks operative cutting kerb stones with an angle grinder, a joiner ripping timber on a bench saw, a plasterer dry-mixing bag cement, and a demolition operative breaking down a concrete slab may all be working within metres of one another on a busy site. The combined and cumulative nature of dust exposure creates a complex risk picture that a single-task assessment cannot fully capture. Principal contractors and employers must therefore approach dust risk assessment with a holistic site perspective, understanding which activities are being carried out simultaneously, which operatives may be in the exposure zone, and how the cumulative dose across a working shift relates to applicable Workplace Exposure Limits (WELs). This chapter provides a structured framework for conducting that assessment, grounded in the requirements of the Control of Substances Hazardous to Health Regulations 2002 (COSHH), EH40/2005 (4th edition 2020), and associated HSE guidance. It should be read alongside Chapter~\ref{chap:Asbestos Risk Assessment}, which addresses the separate but related regime governing asbestos, and Chapter~\ref{chap:Hazardous Substances and COSHH Risk Assessment}, which extends the COSHH framework to chemical hazardous substances more broadly.

\barquote{``Dust may be invisible to the naked eye but its effects on the lung accumulate silently over years. By the time a worker develops breathlessness or a scan reveals fibrosis, the damage is already irreversible. Prevention through rigorous assessment and control is the only defensible approach.'' --- HSE Construction Division guidance on occupational lung disease.}

\example{Silica Exposure During Floor Grinding}{A principal contractor engaged a specialist subcontractor to prepare a 400~m\textsuperscript{2} concrete floor slab for an epoxy resin system in a newly constructed warehouse unit. The works programme required grinding and scabbling to remove a laitance layer and provide a suitably keyed surface. The subcontractor deployed two angle grinders and one ride-on scarifier without on-tool extraction, relying instead on forced ventilation through open roller shutter doors. Personal air sampling subsequently conducted by an occupational hygienist during a comparable activity on a similar site revealed respirable dust concentrations exceeding 1.2~mg/m\textsuperscript{3} at the operative's breathing zone, representing a twelve-fold exceedance of the EH40 WEL for RCS of 0.1~mg/m\textsuperscript{3} as an eight-hour time-weighted average. Had on-tool H-class vacuum extraction been deployed in conjunction with shrouded grinding heads, published HSE data indicate reductions of up to 90 per cent in respirable dust concentrations would have been achievable. The scenario illustrates both the magnitude of the hazard in common construction activities and the technical feasibility of controlling it to within the WEL. The failure was not one of resource but of assessment: no site-specific COSHH assessment had been prepared, and the subcontractor's generic method statement had not been reviewed by a competent person before work commenced.}

%---------------------------- SECTION ----------------------------
\section{Legal and professional framework}

\paragraph{}
The legal framework governing dust exposure in construction is layered and interlocking. At its apex sits the Health and Safety at Work etc.\ Act 1974 (HSWA), which imposes the overarching duty on employers to ensure, so far as is reasonably practicable, the health, safety, and welfare of employees and those affected by their undertaking. This general duty is the foundation upon which all subordinate regulations are built. The Management of Health and Safety at Work Regulations 1999 (MHSWR) operationalise the duty through the requirement for suitable and sufficient risk assessment and the establishment of arrangements for planning, organising, and reviewing preventive and protective measures. These general duties are then given substance in relation to dust and chemical hazards by the specific regulatory regime of the Control of Substances Hazardous to Health Regulations 2002 (COSHH), which provides the most detailed and operationally relevant legal obligations governing dust risk management in the construction environment.

\paragraph{}
Supporting the primary regulatory framework is a comprehensive suite of technical standards, Approved Codes of Practice (ACOPs), HSE guidance documents, and British Standards that define best practice and inform the competent assessor. Key documents include EH40/2005 (4th edition 2020), which sets out the legally enforceable Workplace Exposure Limits for over 500 substances; MDHS101, which provides guidance on general principles of dust protection; HSG97, which guides practitioners through the COSHH assessment process; and BS~EN~481, which establishes the aerodynamic size fraction definitions for airborne particle measurement. The British Occupational Hygiene Society (BOHS) has published P402 and P403 guidance on the assessment and management of construction dust, and the Construction Dust Partnership --- now incorporated within broader BOHS construction occupational health initiatives --- has produced practical toolkits to support site-level implementation.

\begin{itemize}

    \bitem{Health and Safety at Work etc.\ Act 1974 (HSWA)}{The primary enabling legislation for occupational health and safety in Great Britain. Section~2 imposes a duty on every employer to ensure, so far as is reasonably practicable, the health, safety, and welfare of employees; section~3 extends this duty to persons not in employment who may be affected by the undertaking. The Act underpins all subordinate dust management obligations and provides the enforcement framework through which the HSE and local authorities act. Prosecution under the Act can result in unlimited fines and, in cases of gross breach, imprisonment. All subsequent regulations derive their authority from the Act.}

    \bitem{Management of Health and Safety at Work Regulations 1999 (MHSWR)}{Regulation~3 requires every employer to conduct a suitable and sufficient assessment of the risks to health and safety of employees and others arising from the conduct of the undertaking. For dust-generating activities, this means identifying the substances present, the likelihood and severity of exposure, and the measures required to prevent or adequately control that exposure. Regulation~5 requires arrangements for planning, organisation, control, monitoring, and review; Regulation~7 requires the appointment of competent persons to assist in implementing health and safety measures. The MHSWR assessment does not replace the COSHH assessment --- both are required --- but the MHSWR provides the overarching management framework within which the COSHH assessment sits.}

    \bitem{Control of Substances Hazardous to Health Regulations 2002 (COSHH)}{The primary regulatory instrument for dust risk management in construction. Regulation~6 requires employers to carry out a suitable and sufficient assessment of the risk created by work likely to expose employees to a substance hazardous to health; the assessment must identify the nature and degree of exposure and the steps required to meet the requirements of the Regulations. Regulation~7 requires prevention of exposure where reasonably practicable, and adequate control where prevention is not possible; the control hierarchy must be applied in order. Regulation~8 requires that employers ensure control measures are properly used and applied. Regulation~9 requires that engineering controls, including local exhaust ventilation (LEV) systems, are maintained in efficient working order and good repair, with thorough examination and test of LEV at least every 14 months. Regulation~10 requires monitoring of exposure at the workplace where necessary to ensure compliance with WELs. Regulation~11 requires health surveillance where there is a reasonable likelihood that exposure will give rise to an identifiable disease or adverse health effect.}

    \bitem{EH40/2005 --- Workplace Exposure Limits (4th edition 2020)}{The HSE publication that gives legal effect to Workplace Exposure Limits (WELs) as approved by the Health and Safety Executive under COSHH. WELs are expressed as eight-hour time-weighted averages (TWA) and, for some substances, short-term exposure limits (STEL) over a 15-minute reference period. For respirable crystalline silica (quartz), the WEL was reduced from 0.3~mg/m\textsuperscript{3} to 0.1~mg/m\textsuperscript{3} TWA with effect from 17 January 2020, reflecting accumulating evidence of health effects at lower concentrations. For hardwood dust (including oak and beech), the WEL is 1~mg/m\textsuperscript{3} inhalable dust TWA; for softwood, 5~mg/m\textsuperscript{3} inhalable dust TWA. Inhalable dust (all particles capable of entering the nose and mouth) has a general WEL of 10~mg/m\textsuperscript{3}; respirable dust (particles penetrating to the alveolar region) has a general WEL of 4~mg/m\textsuperscript{3}. Exceeding a WEL is not automatically a criminal offence, but it is strong evidence that adequate control has not been achieved.}

    \bitem{HSG97 --- A Step by Step Guide to COSHH Assessment}{The HSE's principal practical guidance document for those conducting COSHH assessments. It provides a five-step methodology applicable to all hazardous substance assessments, including dust, and advises on how to gather information about substances, work activities, and exposure routes; how to evaluate the risk; and how to record, implement, and review the assessment. HSG97 is particularly useful for construction managers who may not have formal occupational hygiene training but are nonetheless responsible for conducting or commissioning COSHH assessments for site activities. It stresses that the assessment must be specific to the work being carried out, not a generic document that fails to engage with actual site conditions.}

    \bitem{MDHS101 --- Dust: General Principles of Protection}{An HSE Methods for the Determination of Hazardous Substances document that provides foundational guidance on understanding the nature of dust hazards, the principles of dust control, and the selection of appropriate control measures. It explains the significance of particle size fractions (inhalable, thoracic, and respirable) and their relationship to health effects; it covers the application of engineering controls including LEV; and it addresses the limitations of RPE as a control measure. MDHS101 is essential background reading for competent assessors and is frequently referenced in HSE enforcement actions and civil litigation as the standard against which the adequacy of control measures is judged.}

    \bitem{BS~EN~481:1993 --- Workplace Atmospheres: Size Fraction Definitions for Measurement of Airborne Particles}{A British and European Standard that defines the inhalable, thoracic, and respirable fractions of airborne particulate matter in terms of sampling conventions and aerodynamic diameter. The respirable fraction, which is most relevant to silica and long-term lung disease, corresponds to particles with an aerodynamic diameter of approximately 10~\textmu{}m or less. The standard provides the technical basis for the measurement methodology used in personal air sampling campaigns and is the reference framework for interpreting results against the respirable-fraction WELs in EH40. Correct selection of sampling hardware --- including cyclone pre-separators and gravimetric filters --- depends on compliance with the sampling conventions defined in BS~EN~481.}

    \bitem{Reporting of Injuries, Diseases and Dangerous Occurrences Regulations 2013 (RIDDOR)}{Schedule~3 of RIDDOR lists reportable occupational diseases, including occupational asthma (where the disease is caused by a sensitising agent encountered at work) and certain forms of pulmonary disease. Whilst silicosis itself is not a RIDDOR-reportable disease per se, the HSE expects employers to report cases of occupational asthma caused by wood dust or other workplace sensitisers where a medical practitioner diagnoses a worker as having contracted the condition. Failure to report notifiable conditions is a criminal offence; more significantly, the obligation to report focuses management attention on the seriousness of the hazard and creates a documented record that can be used to drive improvement. RIDDOR records should be cross-referenced with the health surveillance programme and the outcomes of ongoing exposure monitoring.}

    \bitem{Personal Protective Equipment at Work Regulations 2022 (as amended)}{These Regulations, which extended the scope of the 1992 PPE at Work Regulations to cover workers who are not employees (including certain contractors and agency workers), impose obligations on employers to ensure that PPE is suitable for the risk, that it is CE or UKCA marked and meets relevant standards, that employees are trained in its use, and that it is properly maintained. For dust RPE, the relevant standards include BS~EN~149 (FFP disposable filtering facepieces), BS~EN~140 (half masks), BS~EN~143 (P3 particulate filters), and BS~EN~136 (full-face masks). Critically, the Regulations require that face-fit testing is conducted --- in accordance with the requirements of BS~EN~13274 --- to confirm the adequacy of the seal between the RPE facepiece and the wearer's face. Fit testing is not optional: an RPE device that has not been fit-tested cannot be assumed to provide the manufacturer's stated protection.}

\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Conducting the assessment using the five-step method}

\paragraph{}
The COSHH assessment for construction dust activities follows the same five-step methodology described in HSG97 and applicable to all COSHH assessments. However, the particular nature of dust hazards --- variable generation rates, task-dependency, site-to-site variability, and the cumulative nature of respiratory disease --- requires that the assessor applies this framework with a degree of technical rigour beyond that needed for simpler substance assessments. Each step is described below with reference to the specific considerations relevant to silica, wood dust, and construction dust more broadly.

\subsection{Step 1: Identify the hazardous substances present}

\paragraph{}
The first step requires the assessor to establish which substances are present in the work environment and which of those are hazardous to health. For dust assessments in construction, this begins with a survey of materials to be used, disturbed, or removed during the works. Concrete, mortar, calcium silicate blocks, granite, sandstone, slate, and many engineering bricks contain crystalline silica in concentrations that can give rise to respirable dust at levels above the WEL when mechanically disturbed. The assessor should obtain silica content data from manufacturers' safety data sheets (SDS) where available, or refer to published data in HSE guidance such as the Construction Dust Partnership toolkits. Hardwood species commonly encountered in construction --- oak (Quercus spp.), beech (Fagus sylvatica), iroko (Milicia excelsa), sapele (Entandrophragma cylindricum), and teak (Tectona grandis) --- all carry Group~1 carcinogen classification for their dust. The assessor should identify which species are in use on the site, since different joinery packages may specify different timbers. Mixed dust environments --- where RCS-bearing materials and wood are both present --- require a combined assessment that recognises the potential for additive or synergistic health effects. The identification step should also consider secondary dust sources: sweeping dry debris, emptying vacuum bags, handling dusty materials, and using compressed air for cleaning (a practice that should be prohibited) all generate airborne dust at potentially hazardous concentrations.

\subsection{Step 2: Determine who might be harmed and how}

\paragraph{}
The second step requires the assessor to identify which workers --- and potentially which non-workers, including members of the public, adjacent tradespeople, and site visitors --- may be exposed to the identified hazardous dusts. In the construction context this requires a thorough understanding of site layout, task scheduling, and the degree to which dust-generating activities are isolated from other workers. A concrete cutting operative using an angle grinder without LEV may expose not only themselves but also labourers, trades coordinators, and other operatives working within a radius of several metres; in an enclosed building, recirculated air may carry fine respirable particles throughout the floor plate. The assessor should consider the following exposure pathways: direct inhalation at the point of dust generation; secondary inhalation from settled dust resuspended by foot traffic or airflow; and, less commonly, dermal contact and ingestion from hand contamination. Particular attention must be given to young or new workers who may not have received adequate training or supervision; workers with pre-existing respiratory conditions who may be at elevated risk; and workers on short-duration but high-intensity tasks where the concentration during the task, even if brief, may produce a significant contribution to the TWA exposure. The assessor must also consider the cumulative exposure across the working shift and, for workers who carry out high-exposure tasks regularly, the cumulative lifetime dose.

\subsection{Step 3: Evaluate the risks}

\paragraph{}
Risk evaluation for dust exposure requires the assessor to estimate whether exposure to the identified substances could plausibly exceed the relevant WEL and, where that possibility exists, to characterise the severity and likelihood of harm with sufficient precision to justify the controls selected. Where personal air sampling data are available from similar activities conducted under similar conditions --- as documented in the HSE's COSHH essentials database, the BOHS P403 toolkit, or published peer-reviewed literature --- these should be used to anchor the assessment. Where sampling data are not available, the assessor should apply a precautionary approach: dust-generating activities involving silica-bearing materials or hardwood, particularly when carried out in enclosed or poorly ventilated spaces using dry methods, should be assumed to present a significant risk of WEL exceedance until monitoring evidence demonstrates otherwise. The EH40 WEL for RCS of 0.1~mg/m\textsuperscript{3} is extremely challenging to achieve through ventilation alone without on-tool extraction: studies commissioned by the HSE and the Construction Dust Partnership have consistently shown that dry cutting concrete or stone with an angle grinder in even moderately ventilated conditions generates respirable dust concentrations at the operative's breathing zone ranging from 1 to over 10~mg/m\textsuperscript{3}, representing exceedances of the WEL by factors of 10 to 100. The risk evaluation must also consider task duration: a short burst of cutting of five minutes may produce a significant contribution to the eight-hour TWA for a worker who then carries out other non-dust-generating tasks for the remainder of the shift; prolonged cutting over four or more hours may represent an exceedance even with some control measures in place.

\subsection{Step 4: Determine appropriate controls and precautions}

\paragraph{}
The fourth step requires the assessor to specify the control measures that will be implemented to prevent exposure or reduce it to a level that is as low as reasonably practicable (ALARP) and in any event below the applicable WEL. COSHH Regulation~7 establishes a mandatory hierarchy of controls: prevention first, then engineering controls, then organisational and administrative measures, and finally respiratory protective equipment (RPE) as a last resort or supplement. For construction dust, the practical application of this hierarchy typically involves some combination of wet methods (water suppression), on-tool local exhaust ventilation, work scheduling to minimise exposure duration and the number of workers in the exposure zone, and RPE as a supplementary control where residual exposure remains. The selection of RPE must be based on the assigned protection factor (APF) system: an FFP3 disposable filtering facepiece has an APF of 20 (meaning it reduces the wearer's exposure by a factor of 20 relative to the ambient concentration), making it suitable for use where concentrations are no more than 20 times the WEL; a P3 half-mask respirator (combination of half-mask facepiece and P3 particulate filter) has an APF of 20; a full-face respirator with P3 filters has an APF of 40. The assessor must ensure that the selected RPE is capable of protecting the wearer at the actual concentrations expected, allowing for uncertainty in the exposure estimate.

\subsection{Step 5: Record, implement, and review the assessment}

\paragraph{}
The fifth step requires that the assessment is recorded in a form that is legible, comprehensible, and accessible to those who need to act upon it; that the specified controls are actually implemented on site; and that the assessment is reviewed at appropriate intervals and whenever there is reason to believe it may no longer be valid. COSHH requires that the record of assessment is retained and that it identifies the hazardous substances, the nature and degree of exposure, the existing controls, the residual risk, and the conclusions of the assessment. For construction projects, the assessment must be task-specific and site-specific: a generic ``dust'' assessment that lists PPE and ventilation without specifying the tasks, materials, methods, and monitoring requirements is not a suitable and sufficient assessment within the meaning of COSHH. The assessment should cross-reference the site health and safety plan, the construction phase plan, and any relevant method statements. Review should be triggered by changes in the work activity, the introduction of new materials or tools, the outcome of exposure monitoring, the results of health surveillance, the occurrence of a reportable incident or near miss, and at minimum annually on longer-duration projects.

\example{Five-Step Application: Concrete Cutting for Drainage Channel}{A main contractor appointed a groundworks subcontractor to cut a drainage channel through a reinforced concrete floor slab within an existing industrial unit that was to be refurbished and re-occupied. The works involved approximately 12 linear metres of cutting using a 350~mm floor saw, followed by breaking out with a demolition hammer and removal of debris. Step~1 of the COSHH assessment identified the concrete slab as containing approximately 30\% crystalline silica by mass, generating both respirable and inhalable dust fractions during cutting. Step~2 identified the floor saw operative, a labourer assisting with debris removal, and a site manager who would be in the work area periodically as those at risk. Step~3 evaluated the risk as high: published HSE data for equivalent activities using a floor saw without water suppression suggested respirable dust concentrations in the range of 0.5--4~mg/m\textsuperscript{3}, representing between five and forty times the RCS WEL. Step~4 specified the following control hierarchy: water-cooled blade on the floor saw (wet cutting eliminating approximately 85\% of dust generation); on-tool extraction on the demolition hammer (M-class vacuum); respiratory protective equipment (FFP3 disposable mask for all persons in the work area during cutting and breaking); exclusion zone of 3~metres demarcated with barriers and signage; no dry sweeping (wet mopping or vacuum with M-class or H-class appliance only). Step~5 required that the assessment be documented on the subcontractor's COSHH assessment form, countersigned by the site manager, briefed to operatives during induction, and reviewed on completion of the first day of cutting in light of the actual conditions observed.}

%---------------------------- SECTION ----------------------------
\section{Applying the hierarchy of controls}

\paragraph{}
The COSHH control hierarchy requires that each level be considered in sequence before moving to the next: only where a higher-level control is not reasonably practicable may the assessor proceed to a lower-level measure. In practice, for construction dust activities, pure elimination is rarely achievable since the dust-generating activity is intrinsic to the construction process; the hierarchy therefore usually begins at substitution and engineering controls. However, the mandatory nature of the hierarchy means that the assessor must genuinely engage with the question of whether higher-level controls are possible before accepting the status quo. The following itemised breakdown describes the application of each tier to construction dust management, with particular reference to the specific dust types addressed in this chapter.

\begin{itemize}

    \bitem{Elimination}{Elimination of the dust hazard requires eliminating the dust-generating activity entirely. In construction, this is sometimes achievable through pre-fabrication: ordering pre-cut materials to length and specification from the supplier, using factory-machined joinery components that do not require on-site ripping or planing, or using pre-mixed wet mortar in lieu of on-site batching from dry cement and sand. Where a granite worktop or stone cladding panel can be cut to precise dimensions in a factory environment --- where fixed LEV systems and wet cutting are standard --- rather than on site, the elimination of the on-site cutting activity removes the exposure entirely. Similarly, specifying solid concrete blocks in dimensions that do not require cutting, or using pre-formed ducting that does not require chasing into masonry, can eliminate significant dust sources. Elimination should always be the first question put to the designer, the estimator, and the site team: ``Is there a way of achieving the same construction outcome without generating this dust?'' Where the answer is yes, and the cost or programme implications are not disproportionate, elimination should be pursued.}

    \bitem{Substitution}{Where the dust-generating activity cannot be eliminated, substitution involves replacing either the material or the method with a less hazardous alternative. Material substitution may include using calcium silicate boards in lieu of MDF (which generates high concentrations of respirable and inhalable dust during cutting), specifying pre-primed and pre-finished joinery to reduce on-site sanding, or using lower-silica-content concrete mixes where engineering requirements permit. Method substitution is particularly significant for RCS exposure: replacing dry disc cutting of concrete or stone with water-fed diamond sawing, replacing scabbling with floor preparation systems that incorporate integral water delivery, or replacing dry blasting with wet-abrasive blasting or vacuum-recovery blasting systems, can each produce very substantial reductions in airborne dust generation. The Partnership for Health and Asthma Management (PHAM) in construction has advocated strongly for method substitution as a primary control, noting that many high-risk activities in the construction sector have technically feasible wet alternatives that are not routinely specified because the hazard has not been adequately assessed.}

    \bitem{Engineering controls}{Engineering controls are the primary means of achieving adequate dust control in the construction environment once elimination and substitution have been applied as far as reasonably practicable. The principal engineering control is local exhaust ventilation (LEV): a system designed to capture dust at or close to the point of generation and transport it to a collection and filtration unit before it enters the worker's breathing zone. On-tool LEV --- in which the extraction shroud or hood is integrated with the cutting, grinding, or sanding tool --- is the most effective configuration, since it minimises the distance between the dust source and the capture point. For angle grinders, disc cutters, and floor saws used on concrete or stone, on-tool extraction using an H-class (high hazard) vacuum compliant with BS~EN~60335-2-69 is the standard engineering control: H-class vacuums are rated to retain at least 99.995\% of particles at 0.3~\textmu{}m (equivalent to HEPA H14 filter efficiency) and are therefore suitable for use with RCS and hardwood dust. M-class vacuums (99\% filtration efficiency at 0.3~\textmu{}m) may be acceptable for lower-hazard wood dust activities but should not be used for RCS without careful evaluation of the residual exposure. General dilution ventilation --- opening windows, using fans, or operating within open structures --- is not an adequate substitute for on-tool LEV where WELs for RCS or hardwood dust are at risk of being exceeded. Water suppression systems, whilst not technically LEV, function as an engineering control by suppressing dust at the point of generation: water-fed blades on disc cutters and floor saws, wet drilling attachments, and continuous water delivery to scabbling heads can reduce airborne dust concentrations by 80--90\% compared with dry methods. Engineering controls must be subject to the maintenance and examination requirements of COSHH Regulation~9, including periodic thorough examination and test of LEV systems by a competent person.}

    \bitem{Administrative controls}{Administrative controls operate at the level of how work is organised, scheduled, and supervised rather than at the level of the physical environment or tools. For construction dust, the principal administrative controls are: time limitation, whereby operatives are rotated between high-exposure and low-exposure tasks to reduce the duration of exposure within the eight-hour reference period and thereby reduce the time-weighted average; zoning, whereby dust-generating activities are physically segregated from other work areas, reducing the number of workers at risk and limiting the spread of contaminated air; scheduling, whereby high-dust activities are programmed during periods when the minimum number of workers are present on site, such as before the main workforce arrives or during breaks; training and instruction, whereby operatives receive adequate information about the hazards they face, the controls they are required to use, and the correct way to don, wear, and maintain RPE; and supervision, whereby site management actively monitors compliance with control measures and intervenes promptly where controls are not being used as specified. Administrative controls are secondary to engineering controls within the COSHH hierarchy: they depend on human behaviour and therefore introduce variability and the possibility of failure that engineering controls, by their nature, avoid. They are necessary and important components of a comprehensive control strategy but should not be relied upon as the primary means of controlling exposure where engineering controls are reasonably practicable.}

    \bitem{Personal Protective Equipment (PPE) and Respiratory Protective Equipment (RPE)}{RPE is the last resort in the control hierarchy for dust exposure: it does not reduce the concentration of dust in the work environment; it protects only the individual wearer; its effectiveness depends entirely on correct selection, fit testing, donning, wear discipline, and maintenance; and it places the burden of protection on the worker rather than on the system. COSHH Regulation~7 makes clear that RPE must not be used as the primary means of achieving adequate control where engineering controls or other higher-level measures are reasonably practicable. That said, RPE is a critical supplementary control and, in some circumstances --- particularly residual exposure after engineering controls, short-duration high-concentration tasks, and emergency maintenance situations --- it may be the most practical protection available. For RCS exposure, the minimum RPE standard recommended by HSE is FFP3 disposable filtering facepiece (assigned protection factor 20) or a P3 half-mask respirator (APF 20) for moderate exceedances of the WEL; where the concentration is likely to exceed 2~mg/m\textsuperscript{3} (20 times the WEL), a full-face respirator with P3 filters (APF 40) is required. For hardwood dust, the same RPE standards apply given the carcinogen status of the dust. All RPE must be subject to face-fit testing in accordance with BS~EN~13274 before first use and following any significant change in facial configuration (dental work, weight change, injury) or change of RPE model; this applies to both qualitative fit testing (using saccharin or bitrex challenge agents) and quantitative fit testing (using a fit test machine measuring particle counts inside and outside the facepiece). Disposable FFP3 masks must be discarded after each use; reusable half-masks and full-face masks must be cleaned and inspected, and filters replaced in accordance with the manufacturer's instructions and the findings of exposure monitoring.}

\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Cadence of assessment and review triggers}

\paragraph{}
A COSHH assessment for construction dust is not a one-time document: it must be a living record that is reviewed and updated in response to changes in the work, the workplace, and the evidence of how controls are performing. COSHH Regulation~6 requires that an assessment is reviewed if there is reason to suspect it is no longer valid or if there has been a significant change in the work to which it relates. The HSE's published guidance makes clear that ``significant change'' should be interpreted broadly in the context of dust-generating activities, since even apparently modest changes to task duration, task location, or tool selection can produce material changes in exposure. On active construction projects, assessments should be reviewed at least at each significant programme milestone --- for example, when moving from groundworks to structure, from structure to fit-out, or when a new subcontractor mobilises. The following events constitute specific triggers that must initiate an immediate review of the dust COSHH assessment.

\begin{itemize}

    \bitem{Introduction of a new activity, material, or tool}{Any change that introduces a new dust-generating activity, a material of different composition or hazard classification, or a tool with different generation characteristics --- for example, switching from a floor saw to a demolition diamond blade, or beginning work with a new timber species --- must trigger a reassessment. The original assessment cannot be assumed to remain valid because the conditions that informed it have changed.}

    \bitem{Results of personal air sampling or static monitoring}{Where exposure monitoring reveals that concentrations are exceeding or approaching the applicable WEL, the assessment must be reviewed immediately and controls enhanced. An exceedance is evidence that the assessment's evaluation of risk was incorrect or that controls are performing below their assumed effectiveness; both require a response. Equally, where monitoring shows that concentrations are comfortably below the WEL with current controls in place, the assessment may be able to be simplified or the monitoring frequency reduced, but the assessment must be formally reviewed to capture this finding.}

    \bitem{Health surveillance outcomes}{Any adverse finding from the health surveillance programme --- including a decline in FEV\textsubscript{1} or FVC on spirometry, a positive result on the Medical Research Council dyspnoea questionnaire, or a new diagnosis of occupational asthma, rhinitis, or silicosis in any member of the workforce --- must trigger an immediate review of the COSHH assessment and an investigation into the cause of the health effect. COSHH Regulation~11 requires that health surveillance is provided where there is a reasonable likelihood that work-related ill-health will occur; the outcome of that surveillance is integral to the feedback loop of the assessment process.}

    \bitem{Incident, accident, or near-miss involving dust exposure}{Any incident in which dust controls have failed --- for example, a vacuum extraction unit malfunctioning and venting unfiltered exhaust, an RPE seal failure identified during fit testing, an operative observed working without specified controls --- must be investigated and the findings fed back into the assessment. Near-miss reporting is particularly valuable for identifying systemic weaknesses in the control regime before they result in actual harm.}

    \bitem{Change in the applicable legal or technical standards}{Where the WEL for a substance is revised --- as occurred for RCS in January 2020 --- or where new guidance is issued by the HSE, BOHS, or relevant standards bodies, the assessment must be reviewed to ensure that it remains compliant with the updated requirements. Employers who continue to rely on pre-2020 assessments that were calibrated against the old RCS WEL of 0.3~mg/m\textsuperscript{3} are at serious risk of non-compliance, particularly for high-dust activities.}

    \bitem{Change in workforce or supervision}{Where new operatives are introduced to the work, including workers from labour agencies, subcontractors, or self-employed tradespeople, the assessment must be communicated to those individuals and their understanding verified. Where there is a change in site management or supervisory personnel responsible for monitoring and enforcing dust controls, the incoming personnel must be briefed on the requirements of the assessment before the work continues. Where an operative is identified during health surveillance as having a pre-existing respiratory condition that increases their susceptibility to dust-related harm, the assessment must consider whether additional controls or task exclusions are required to protect that individual.}

\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Common failure modes}

\paragraph{}
Analysis of enforcement actions by the HSE, civil litigation records, and the findings of the Construction Dust Partnership and BOHS inspections programmes reveals a consistent pattern of failure modes in dust risk management on UK construction sites. These failures are rarely attributable to ignorance of the hazard at the organisational level: the senior management of most principal contracting businesses is well aware that silica and wood dust are serious occupational health hazards. The failures occur at the interface between policy and practice --- in the gap between the document and the site --- and they are typically systemic and repeated rather than isolated and aberrant. Understanding these failure modes is essential for any competent assessor who seeks not just to produce a compliant assessment document but to achieve actual protection of workers' health.

\begin{itemize}

    \bitem{Dry cutting without LEV or water suppression}{The most commonly cited failure in HSE enforcement notices and improvement notices relating to construction dust is the use of angle grinders, disc cutters, and floor saws to cut concrete, stone, or masonry without any form of dust suppression. This is frequently observed even on sites where a COSHH assessment specifying wet cutting or on-tool extraction has been formally prepared: the assessment exists as a document but has not been operationalised into the day-to-day conduct of work. The root cause is typically inadequate supervision, absence of the correct tooling on site, or operative preference for the dry method as faster or more convenient. Enforcement action in these cases has included prohibition notices requiring cessation of work, improvement notices requiring the installation of controls within a specified timeframe, and prosecutions resulting in significant fines where the failure was prolonged, systemic, or knowingly condoned by management.}

    \bitem{Face-fit testing not conducted for RPE users}{RPE is only effective if the facepiece forms an adequate seal against the wearer's face. FFP3 disposable masks and half-mask respirators are frequently encountered on construction sites without any record of face-fit testing. In such cases, the device may be providing little or no protection despite appearing to be worn correctly: facial contours, beard growth, nasal bridge geometry, and cheekbone structure all affect the seal, and only a formal fit test using the methodology of BS~EN~13274 --- qualitative or quantitative --- can confirm adequacy. The HSE has made clear in its published guidance and enforcement communications that the supply of RPE to workers without face-fit testing is not compliance with COSHH: the employer is providing equipment that may be ineffective and recording it as a control measure, thereby providing false assurance of protection.}

    \bitem{Health surveillance not arranged}{COSHH Regulation~11 imposes a specific duty to arrange health surveillance for workers exposed to substances known to cause occupational disease where there is a reasonable likelihood of disease occurring. For RCS and hardwood dust, the duty to provide health surveillance is clear and has been well-established in HSE guidance for many years. In practice, health surveillance for construction workers is inconsistently implemented: on longer-term projects, spirometry and occupational health questionnaires may be conducted at induction but not repeated annually; on short-duration projects and for self-employed workers, it may not be provided at all. The consequence is that early-stage disease --- treatable and manageable if identified promptly --- progresses undetected until the worker presents with advanced symptoms, at which point irreversible damage has occurred. The cost of providing health surveillance through an occupational health provider is modest compared with the cost of compensating a worker who has developed silicosis or occupational asthma.}

    \bitem{RPE not worn correctly or continuously}{Even where RPE has been correctly selected, face-fit tested, and issued to the operative, observational studies on construction sites consistently record high rates of non-compliance with wear discipline. Common failures include: wearing the mask under the chin rather than over the nose and mouth; not conducting the positive and negative pressure pre-use checks to verify the seal before beginning work; removing the mask during the dust-generating activity to communicate verbally with colleagues; and wearing a mask with beard growth that prevents an adequate seal. Administrative and supervisory controls are required to address these behaviours, but the root cause is frequently inadequate training: operatives who understand why the mask is important and how their exposure translates into disease risk are significantly more likely to maintain correct wear discipline than those who have simply been told to wear the mask.}

    \bitem{Dust monitoring not conducted where controls are in use}{A common misconception in site management is that the presence of control measures --- on-tool extraction, water suppression, RPE --- implies compliance with the WEL without any need for measurement. COSHH Regulation~10 requires monitoring of exposure at the workplace where it is necessary to ensure maintenance of adequate control. For activities involving RCS and hardwood dust, where the WEL is challenging to achieve and the consequences of exceedance are severe, personal air sampling should be regarded as a standard part of the control verification process rather than an optional extra. Published HSE data show that on-tool extraction and water suppression systems do not uniformly achieve the same percentage reduction in exposure across all tool types, materials, and working conditions: measurement is the only reliable means of verifying that the assumed control effectiveness has been achieved in practice. Exposure monitoring should be conducted by a competent occupational hygienist using personal sampling equipment calibrated against the conventions of BS~EN~481.}

    \bitem{WEL exceedance not acted upon}{Where monitoring reveals a WEL exceedance, the correct response is to treat the result as evidence that adequate control has not been achieved within the meaning of COSHH Regulation~7 and to implement enhanced controls immediately. A failure mode frequently observed in practice is that monitoring results are filed without triggering any change in working practice: the result is noted, sometimes acknowledged in a meeting, but the underlying control deficiency is not remedied. In enforcement terms, an employer who has commissioned monitoring, received a result showing WEL exceedance, and failed to act upon it is in a significantly worse legal position than one who has not monitored at all: the knowledge of the exceedance creates an unambiguous duty to act, and inaction in the face of that knowledge is difficult to defend in either regulatory proceedings or civil litigation.}

\end{itemize}

\example{Failure Pattern: Dry Cutting of Paving Slabs on a Housing Development}{A subcontractor engaged to lay block paving on a residential development was observed by HSE inspectors cutting natural stone paving slabs using hand-held angle grinders without water suppression or on-tool extraction. The works had been ongoing for approximately three weeks. The COSHH assessment prepared by the subcontractor referenced ``water suppression or dust extraction'' as the specified control measure but the site manager confirmed that no water delivery equipment was available on site and no extraction unit had been mobilised. The operatives were wearing FFP2 (not FFP3) disposable masks, none of which had been face-fit tested. Spot-check air sampling conducted by the inspectors during the cutting activity recorded a respirable dust concentration of 2.7~mg/m\textsuperscript{3} at the operative's breathing zone, representing a 27-fold exceedance of the RCS WEL. A prohibition notice was served immediately, requiring cessation of all dry cutting until adequate controls were in place. An improvement notice was subsequently served requiring the subcontractor to: procure and mobilise water-fed cutting equipment or on-tool extraction; replace FFP2 masks with FFP3 devices; arrange face-fit testing for all operatives; and arrange lung function testing through an occupational health provider. The scenario encapsulates the most common pattern of dust control failure: an adequate assessment on paper, wholly absent from practice.}

%---------------------------- CHAPTER SUMMARY ----------------------------
\newpage
\section{Chapter Summary}

\begin{table}[H]
\centering
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm}X}
\toprule
\textbf{Assessment Element} & \textbf{Operational Requirement} \\
\midrule
Legal/Professional Basis & COSHH Regulations 2002 (Regulations 6, 7, 8, 9, 10 and 11); EH40/2005 (4th edition 2020) --- RCS WEL 0.1~mg/m\textsuperscript{3} TWA, hardwood dust WEL 1~mg/m\textsuperscript{3} TWA; MDHS101; HSG97; BS~EN~481; PPE at Work Regulations 2022; RIDDOR 2013; HSWA 1974; MHSWR 1999. \\
Method & Apply the full five-step COSHH assessment process: identify substances; determine who may be harmed; evaluate risks against WELs; specify and document the hierarchy of controls; record, implement, and review. Assessment must be task-specific, site-specific, and conducted or reviewed by a competent person. \\
Controls & Apply the COSHH control hierarchy in order: elimination (pre-cut materials, prefabrication); substitution (wet methods, lower-hazard materials); engineering controls (on-tool H-class or M-class LEV, water-fed blades, wet drilling, LEV shrouds); administrative controls (zoning, scheduling, rotation, training, supervision); RPE as a supplementary measure (FFP3 or P3 half-mask APF 20 for RCS and hardwood dust; full-face APF 40 where concentrations exceed 20$\times$ WEL); all RPE to be face-fit tested per BS~EN~13274. \\
Cadence & Review on every significant change of activity, material, or tool; on receipt of adverse monitoring or health surveillance results; following any incident or near-miss; on revision of applicable WELs or guidance; at minimum annually on longer-duration projects. Personal air sampling to be conducted by a competent occupational hygienist to verify control effectiveness for RCS and hardwood dust activities. \\
Assurance & Health surveillance programme (annual spirometry, PEFR, occupational health questionnaire) for all workers with regular dust exposure; COSHH Regulation~9 examination and test of LEV systems at least every 14 months; face-fit test records retained and reviewed; monitoring results actioned; audit of control compliance by site management; governance reporting to contractor health and safety leadership. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{itemize}

    \bitem{Key takeaway 1}{Respirable crystalline silica is a Group~1 human carcinogen and the EH40 WEL was reduced to 0.1~mg/m\textsuperscript{3} TWA in January 2020. Common construction activities --- dry concrete cutting, grinding, and scabbling --- routinely generate concentrations many times this level without adequate controls. On-tool H-class LEV and water suppression are the primary engineering controls and should be treated as non-negotiable requirements for all silica-bearing material cutting on UK construction sites.}

    \bitem{Key takeaway 2}{Hardwood dust (oak, beech, and other species) carries Group~1 carcinogen status for sinonasal adenocarcinoma with a WEL of 1~mg/m\textsuperscript{3}. Softwood dust is a significant cause of occupational asthma. Health surveillance including annual spirometry and lung function testing is a legal requirement under COSHH Regulation~11 for workers regularly exposed to wood dust; it must not be treated as optional or deferred on grounds of cost or programme pressure.}

    \bitem{Key takeaway 3}{The COSHH hierarchy of controls requires genuine engagement with elimination and substitution before relying on RPE. Pre-cutting materials to specification off-site, specifying wet methods, and using pre-fabricated components are not merely aspirational: where they are reasonably practicable, they are legally required as the preferred means of control. The assessment must document why higher-level controls were or were not pursued.}

    \bitem{Key takeaway 4}{RPE is only effective where it has been correctly selected for the exposure level (using assigned protection factors), face-fit tested in accordance with BS~EN~13274 for each individual wearer, correctly donned and worn continuously during exposure, and maintained in accordance with the manufacturer's instructions. An FFP3 mask that has not been fit-tested, is worn with beard growth, or is removed during the task provides little or no protection. Face-fit testing is a legal requirement, not a best-practice aspiration.}

    \bitem{Key takeaway 5}{The construction dust risk assessment must be a living document: it must be reviewed whenever conditions change, whenever monitoring reveals an exceedance, and whenever health surveillance identifies an adverse finding. The most consequential failures in dust risk management are not failures of initial assessment but failures of review and enforcement: assessments prepared at mobilisation and not revisited as conditions evolve, and control measures specified in writing but not implemented or monitored in practice. Effective governance requires that site management actively verifies compliance with control measures and that assessment records reflect the reality of what is happening on site.}

\end{itemize}

\barquote{In Chapter~\ref{chap:Hazardous Substances and COSHH Risk Assessment}, the COSHH assessment framework introduced in this chapter is extended beyond construction dust to encompass the full range of hazardous chemical substances encountered in the built environment sector: solvents, adhesives, paints, resins, isocyanate-containing products, and other substances that present inhalation, dermal, and systemic health risks requiring structured assessment, exposure monitoring, health surveillance, and layered control. The principles of hierarchy, documentation, and review established here apply with equal force to those hazards, and the competent assessor will find a consistent methodology applicable across all COSHH-regulated substances.}
