\chapter{Electrical Safety Risk Assessment}
\label{chap:Electrical Safety Risk Assessment}

%---------------------------- SECTION ----------------------------
\section{Why this assessment matters in construction}

\paragraph{}
Electrical hazards represent one of the most lethal categories of risk on United Kingdom construction sites, accounting for a disproportionate share of fatal and life-changing injuries relative to the number of workers directly engaged in electrical activities. Unlike many construction hazards that present visible, tangible dangers, electricity is invisible in its transmission, instantaneous in its effect, and frequently encountered in circumstances where site operatives have no formal electrical training whatsoever. A groundworker striking a buried service cable with a mechanical excavator, a labourer connecting a spur from a site distribution board without understanding phase-to-earth voltage, or a finishing tradesperson using a 230V domestic extension lead sourced from home in a damp internal environment --- each represents a scenario that recurs with tragic regularity on sites across England, Scotland, and Wales. The Health and Safety Executive's published fatality statistics consistently identify contact with electricity as a principal cause of construction site deaths, and the broader European injury database reinforces the picture that electrical accidents, when they occur, carry an exceptionally high severity ratio: the proportion of electrical contacts that result in fatality or permanent impairment is materially higher than for most other construction injury mechanisms.

\paragraph{}
The particular hazard profile of construction sites distinguishes them sharply from fixed industrial premises. Temporary supplies must be established, modified, and eventually removed as the project proceeds through its phases. Workers operating from generators, transformer centres, and site distribution boards (SDBs) frequently work in conditions of incomplete structural enclosure, exposed to weather, mud, and mechanical damage from plant and vehicles. Cable management is inherently more difficult when the building is not yet finished, penetrations through floors and walls remain open, and operatives from multiple trades are working in overlapping zones simultaneously. The 110-volt centre-tapped-to-earth (CTE) system mandated by site safety best practice for portable hand tools substantially reduces --- though does not eliminate --- the risk from contact with live conductors, since a fault between phase and earth produces a maximum voltage of 55\,V to earth rather than 230\,V. Nevertheless, even 55\,V can cause ventricular fibrillation in susceptible individuals, and the CTE system offers no protection when operatives misuse 230V equipment sourced outside the controlled supply, bypass protection devices, or work on live circuits without authorisation or demonstrated competence. The disciplined management of the temporary electrical installation is therefore a primary duty of the principal contractor and must be treated with the same rigour as other life-safety systems on the project.

\paragraph{}
Arc flash, a hazard less widely understood among non-electrical workers than simple electric shock, deserves specific attention in the construction context. An arc flash event occurs when a fault current travels through air between conductors or between a conductor and earth, creating a plasma arc capable of generating temperatures in excess of 20{,}000 degrees Celsius, an intense pressure wave, and radiant energy sufficient to cause severe burns at considerable distances from the arc point. On construction sites, arc flash risk arises principally when workers open, modify, or work near live switchgear, distribution boards, or transformer terminals without appropriate isolation and personal protective equipment. The transition from construction supply to permanent supply during the commissioning phase of a project frequently creates conditions where high-energy circuits are energised whilst the surrounding environment remains physically incomplete --- a combination that demands rigorous permit-to-work controls and the involvement of competent electrical engineers rather than general site personnel. Arc flash energy levels must be calculated by a specialist and the required arc-rated PPE specified before any work on energised high-voltage or high-fault-energy low-voltage equipment is permitted to proceed.

\paragraph{}
Overhead power lines represent a recurring cause of fatalities in construction and adjacent civil engineering and agricultural sectors. The Electricity Safety, Quality and Continuity Regulations 2002 place duties on those who distribute electricity, but the primary duty on the construction employer and principal contractor is to identify overhead line constraints at the planning stage, implement goal-post crossing restrictions, maintain safe working distances, and ensure plant operators are trained and briefed. HSE guidance document GS6, \textit{Avoiding Danger from Overhead Power Lines}, sets out a hierarchy of protective measures for working near overhead lines that is directly applicable to construction sites and whose requirements must be incorporated into the site-specific risk assessment. Fatal accidents involving mobile cranes, telehandlers, concrete pumps with extended booms, and tipper lorry bodies raised inadvertently near overhead lines continue to be investigated by the HSE every year, underscoring that this hazard is neither obscure nor highly technical --- it is a straightforward but persistently fatal interface between everyday construction activity and live electrical infrastructure. Preventing contact requires advance planning, physical engineered controls, and sustained management attention throughout the project, not merely a note on the method statement.

\barquote{``Electricity obeys physics, not intention. The operative who believes they understand electricity well enough to improvise is the operative most likely to die from it.'' --- Electrical Safety First, \textit{Safe Isolation Procedure Guidance}, 2019}

\example{Site Electrical Incident Scenario}{A principal contractor engaged in the construction of a medium-sized commercial office building was nearing the end of the first-fix electrical phase. A second-fix carpenter, working independently during a Saturday shift with minimal supervision, required additional lighting in a first-floor corridor. Finding the temporary lighting distribution inadequate, the operative sourced a domestic 13A extension reel from his van and connected it to a 230V socket outlet on an adjacent spur circuit that a first-fix electrician had wired and left with the terminals partially made off but not yet fully secured. The extension reel cable was partially uncoiled, reducing its current-carrying capacity through thermal build-up. The cable ran across a wet concrete floor. Within forty minutes, the cable insulation failed at a point of mechanical abrasion near a door threshold, producing a short circuit. The operative received a serious electric shock and sustained burns to his right hand and forearm. Investigation revealed: no permit system for non-electrical trades accessing electrical circuits; no lock-off/tag-out applied to the incomplete spur; no restriction on the use of 230V domestic equipment; and no site inspection covering the temporary supply that weekend. The assessment failure was not technical --- it was organisational and procedural. The construction phase plan documented a 110V CTE policy, but the site lacked sufficient 110V socket outlets, and no supervisor had challenged the use of 230V equipment in the weeks preceding the incident.}

%---------------------------- SECTION ----------------------------
\section{Legal and professional framework}

The regulation of electrical safety in the construction sector in Great Britain is underpinned by a combination of primary statute, statutory instruments, and non-statutory but highly authoritative guidance documents. The Health and Safety at Work etc.\ Act 1974 imposes the overarching general duties of care upon employers, the self-employed, and those in control of premises. Within that framework, electrical safety is specifically addressed by dedicated regulations, equipment standards, and inspection and testing requirements, each of which plays a defined role in the management of electrical risk on construction sites. Compliance with the specific regulations described below provides a strong evidential platform for demonstrating that duties under the primary statute have been discharged, though the duty to ensure safety extends further than mere technical compliance with any individual instrument: the courts and the HSE will look to the totality of the arrangements in place.

The legal and professional framework applicable to construction electrical safety is extensive and multi-layered. In addition to the principal statutory instruments listed below, the principal contractor has duties under the Construction (Design and Management) Regulations 2015 (CDM 2015) to ensure that the Construction Phase Plan addresses temporary electrical supplies and that designers consider permanent electrical infrastructure during the design phase. Relevant documents include those issued by the Institution of Engineering and Technology (IET), the British Standards Institution (BSI), the Health and Safety Executive's L-series Approved Codes of Practice, and HSG-series guidance documents. The following instruments and standards are of primary relevance to the electrical safety risk assessment:

\begin{itemize}
  \bitem{Electricity at Work Regulations 1989 (EaWR)}{The principal statutory instrument governing electrical safety at work in Great Britain. Regulation~4 requires that all electrical systems shall at all times be of such construction as to prevent, so far as is reasonably practicable, danger. Regulations 10--14 address insulation, protection, earthing, connections, and the requirement that no live working shall be undertaken unless it is unreasonable in all the circumstances to carry out the work dead and suitable precautions have been taken to prevent injury. The Regulations impose duties on ``duty holders'' --- employers, the self-employed, and employees --- and carry criminal liability for breach. Regulation~16 requires that no person engage in electrical work activities unless they possess the technical knowledge or experience to prevent danger, or are under appropriate supervision from someone who does.}

  \bitem{HSG141 --- Electrical Safety on Construction Sites (HSE)}{The Health and Safety Executive's principal guidance document specifically tailored to the construction context. HSG141 provides detailed advice on safe systems of supply, the 110V CTE system, site distribution boards, residual current devices, inspection and testing intervals, overhead lines, buried services, and the competence requirements for those carrying out electrical work on construction sites. Although not an Approved Code of Practice, HSG141 represents the HSE's considered view of how duty holders should achieve compliance with the EaWR in the construction setting, and departure from its recommendations requires demonstration of an equally effective alternative approach. The document sets out minimum inspection and test frequencies for temporary installations that must be incorporated into the project's electrical management plan.}

  \bitem{BS 7671:2018+A2:2022 --- Requirements for Electrical Installations (IET Wiring Regulations, 18th Edition)}{The national standard for electrical installation design, construction, inspection, and testing in the United Kingdom, published jointly by the IET and BSI. Although a British Standard rather than statutory law, BS~7671 is widely recognised by courts and the HSE as representing good practice, and installations departing from it will be scrutinised closely in any post-incident investigation or prosecution. Section~704 of BS~7671 contains special location requirements for construction and demolition site supplies, setting out requirements for fault protection, socket outlet voltage systems, transformer specifications, and minimum inspection and test frequency of the temporary installation.}

  \bitem{GS6 --- Avoiding Danger from Overhead Power Lines (HSE)}{HSE's guidance document addressing the hazard of overhead lines across all industrial and agricultural sectors. Provides a hierarchy of controls from diversion and de-energisation through goal-post crossing restrictions, exclusion zones, and working practice constraints, and specifies safe approach distances dependent upon the operating voltage of the line. Essential reference for any construction site where cranes, plant, scaffold, or personnel will operate in proximity to overhead lines, and must be cited and its requirements applied directly in the site-specific risk assessment.}

  \bitem{HSG47 --- Avoiding Danger from Underground Services (HSE)}{Whilst principally addressed to the excavation hazard (cross-referenced in Chapter~\ref{chap:Excavation and Groundworks Risk Assessment}), HSG47 is directly relevant to electrical safety because the majority of underground utility strikes involve electricity distribution cables operating at 11\,kV or 33\,kV. It covers service avoidance techniques including utility record acquisition, the use of CAT (cable avoidance tool) and Genny (signal generator), hand digging in proximity to identified services, and permit-to-dig systems. The electrical safety risk assessment must cross-reference the site's buried services management procedures and confirm that all groundwork activities are controlled under an HSG47-compliant permit-to-dig regime.}

  \bitem{Provision and Use of Work Equipment Regulations 1998 (PUWER)}{Apply to all electrical tools and equipment used on site. Require that work equipment is suitable for its purpose, maintained in safe condition, inspected at appropriate intervals, and used only by people who have received adequate training, information, and instructions. In the electrical context, PUWER underpins the requirements for portable appliance testing (PAT) and the inspection of portable tools and extension leads, as well as the maintenance of fixed electrical equipment including distribution boards, transformers, and generators. PUWER Regulation~6 requires that the employer ensure work equipment is inspected in order to identify any deterioration that could result in danger.}

  \bitem{Personal Protective Equipment at Work Regulations 2022}{Relevant to the provision and use of electrical PPE including insulated gloves to BS EN~60903, arc flash face shields with appropriate arc rating determined by calculation, arc-rated flame-resistant clothing for high-voltage switching and commissioning activities, and insulated tools to BS EN~60900. The 2022 amendments extend explicit PPE duties to workers in the gig economy and those working under umbrella arrangements, an increasingly significant consideration on construction sites with complex supply chain structures. PPE provision must be risk-assessment-led and matched to the energy levels and hazard types present.}

  \bitem{Electricity Safety, Quality and Continuity Regulations 2002 (ESQCR)}{Apply primarily to Distribution Network Operators (DNOs) rather than to construction employers directly, but are relevant when considering connection to the permanent supply network, notification of temporary supplies, and the interface between the site electrical system and the public distribution network. DNOs have obligations under ESQCR to respond to requests for overhead line diversions or temporary de-energisation, and the construction employer must understand the engagement process with the DNO and initiate contact at the earliest stage of project planning --- not days before cranes are due to be erected.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Conducting the assessment using the five-step method}

\subsection{Step 1 --- Identify Hazards}

\paragraph{}
The hazard identification phase for an electrical safety risk assessment must be comprehensive and must not be limited to the activities of qualified electricians. The full range of interactions between site workers and electrical energy must be catalogued: electricians carrying out first-fix and second-fix wiring; general operatives using 110V power tools; plant operators working near overhead lines; groundworkers excavating near buried cable routes; tradespeople using temporary lighting and battery chargers; site managers operating generators; and maintenance personnel accessing switchrooms and distribution boards during live phases of the project. Hazard identification should draw upon site drawings showing the temporary electrical supply layout; utility records identifying underground cable routes and their operating voltages; network operator records of overhead line positions and voltages including those on adjacent properties; the project's CDM pre-construction information pack, which should contain utility survey results and any constraints identified during the design stage; and inspection records from previous phases of the project. The assessment should explicitly identify and document: exposed or incomplete live conductors in partially wired areas; damaged cable insulation from mechanical abrasion or plant traffic; inadequate earthing of the temporary installation; water ingress to distribution boards, transformers, and socket outlets; mechanical damage to cables at floor penetrations and doorways; unauthorised or uncontrolled connections to the supply; use of 230V equipment in lieu of 110V CTE equipment; inadequate RCD protection; proximity to overhead lines and any identified constraints on crane or plant movement; work in energised switchrooms or near commissioning activities; and the risk of arc flash during energisation and commissioning of permanent high-voltage and low-voltage switchgear. The identification phase must include a physical site walk-round by a competent person rather than being conducted solely from the office using drawings.

\subsection{Step 2 --- Decide Who or What Is Affected}

\paragraph{}
The population at risk from electrical hazards on a construction site is considerably wider than might initially be assumed when focusing solely on the electrical trades. Qualified electricians engaged in first-fix and second-fix wiring are obviously at risk, but the assessment must also encompass: non-electrical tradespeople who use power tools and temporary lighting in their daily work; plant operators whose machines --- mobile cranes, telehandlers, concrete pump booms, tipper bodies --- may inadvertently contact overhead lines; groundworkers and tunnelling operatives working near underground electrical infrastructure with limited visibility of hazards below ground; labourers who may move, drag, coil, or stand upon cables; supervisory and managerial staff who enter plant rooms or switchrooms during their site inspections; visiting professionals including structural engineers, quantity surveyors, building control officers, and client representatives who may enter energised areas without electrical awareness; and members of the public and adjacent occupiers who could be affected by a major electrical incident such as a supply failure, fire, or cable strike near the site boundary. For buried cable strikes, the population at risk extends beyond the operative directly involved to include bystanders in the immediate vicinity, as the explosive release of energy from a high-voltage cable strike can cause injury over a substantial radius. Vulnerable individuals --- young workers or apprentices who may be unfamiliar with electrical hazards, workers whose first language is not English and who may not fully comprehend verbal safety instructions, and new starters who have not yet received site-specific electrical safety induction --- should be specifically identified and provided with proportionate additional controls including enhanced supervision.

\subsection{Step 3 --- Evaluate Likelihood and Consequence}

\paragraph{}
The evaluation of electrical risk must reflect both dimensions of the probability--consequence matrix, but the crucial characteristic of electrical hazards is that consequence is almost invariably at the extreme end of the severity scale. Contact with electrical energy carries a realistic prospect of cardiac arrest from ventricular fibrillation, severe deep-tissue burns from arc flash or sustained contact, fall from height following involuntary muscle contraction, and death. Accordingly, even hazards with relatively low inherent probability --- such as the chance that a specific plant operator will inadvertently contact an overhead line on a particular working day --- must be assigned an overall risk level that reflects the catastrophic and frequently irreversible consequence that would follow contact. A standard 5$\times$5 risk matrix should assign the maximum consequence score to all categories of live electrical contact, making the overall inherent risk rating Very High for uncontrolled electrical hazards. Likelihood should be assessed against specific site conditions: a site working within the statutory approach distance of a 33\,kV overhead line has a materially higher likelihood of a contact event than one where the nearest line is 100 metres distant. Similarly, a site where domestic 230V equipment is observed in widespread use has a higher probability of a fatal shock event than one where the 110V CTE system is comprehensively enforced and audited. The assessment must not allow the familiarity of a hazard to reduce the assigned consequence: operatives who have used 230V extension leads for years without incident may perceive the risk as negligible, but the risk matrix must be completed objectively based on the physics of electrical injury, not on historical near-miss frequency.

\subsection{Step 4 --- Select and Implement Controls}

\paragraph{}
Control selection for electrical hazards must follow the hierarchy of controls, addressed in detail in Section~\ref{sec:elec_hierarchy} of this chapter, with a strong and documented preference for engineering controls over procedural and PPE-based measures. The cornerstone engineering control for construction site portable tool supply is the 110V CTE system delivered through yellow BS EN~60309-2 socket outlets at the site distribution board. All portable hand tools must operate at 110V unless a specific dispensation has been granted, documented, and authorised --- 230V or 400V equipment must be permitted only where a 110V equivalent does not exist, subject to enhanced 30\,mA RCD protection and a formal individual assessment. Residual current devices must be provided at the SDB level and, as an additional protective layer, on individual circuits or through in-line plug-in RCDs where the supply serves higher-risk areas such as excavations, wet environments, or rooftop areas. Lock-off/tag-out (LOTO) procedures must be documented in a written safe system of work and physically enforced using proprietary multi-lock hasp devices, individually assigned padlocks, and danger tags, ensuring that the circuit isolation cannot be reversed whilst any person remains within the work zone. The permit-to-work system for electrical isolation must require that the authorised person physically verifies the dead condition using a two-pole voltage indicator tested against a known live source before and after testing the isolated circuit, following the safe isolation sequence prescribed by HSG141. Overhead line controls must be engineered as goal-post crossing restrictions with physical height limitation barriers and warning notices, supplemented by a plant movement exclusion zone marked on the site layout plan and signed on the ground. All portable electrical equipment must be subject to formal PAT testing at intervals determined by risk assessment --- HSG141 recommends a three-month interval for 110V tools and leads on construction sites, with visual inspection by the user before each use.

\subsection{Step 5 --- Record and Review}

\paragraph{}
The written record of the electrical safety risk assessment must capture the inherent risk rating, the specific controls applied, the resultant residual risk rating, and the named individuals responsible for implementing and monitoring each control measure. Given the dynamic nature of construction site electrical installations --- which are progressively modified as the project advances through phases, as new trades arrive with their specific equipment requirements, and as permanent supplies are energised at the commissioning stage --- the electrical risk assessment must not be treated as a document prepared once at mobilisation and filed until project completion. A formal review mechanism must be built into the project's risk management programme, with mandatory review points triggered by the events specified in Section~\ref{sec:elec_cadence} of this chapter. All inspection and test results for the temporary installation, including the site distribution board test records required by BS~7671 Section~704 and the inspection frequency stipulated in HSG141, must be retained as part of the project's electrical safety documentation and made available to the HSE on request. The electrical safety risk assessment should be signed by the principal contractor's project manager and countersigned by the principal contractor's competent electrical person, to confirm that the assessment has been technically reviewed by someone with the relevant knowledge and that the controls described are both appropriate to the hazards identified and physically achievable within the site's infrastructure.

\example{Five-Step Application}{A principal contractor mobilising for a 14-month mixed-use residential and commercial development on a brownfield site adjacent to an 11\,kV overhead distribution line and traversed by a network of redundant high-voltage underground industrial cables carries out the electrical safety risk assessment as follows. \textbf{Step~1 (Hazards):} Overhead 11\,kV line with a minimum ground clearance of 6.8\,m running across the eastern boundary within the operating radius of the tower crane; unmapped high-voltage underground cables identified from historical electricity board utility records and confirmed by trial holes; temporary supply from a 200\,kVA diesel generator feeding four SDBs at grid locations; 110V and 230V circuits for trade use; commissioning of permanent 11\,kV switch room and 400V LV distribution in project months 11--14. \textbf{Step~2 (Who Affected):} Groundworkers, plant operators (tower crane, two telehandlers, concrete pump), all trade electricians, all other trades using power tools, commissioning engineers from the specialist HV contractor, visiting client team, building control inspectors. \textbf{Step~3 (Evaluation):} Overhead 11\,kV line --- inherent risk Very High (contact highly likely to be fatal, crane operations within exclusion distance without controls); underground HV cables --- High (strike possible given uncertainty of cable routing in brownfield ground); portable tools at 110V --- Medium inherent (correct voltage but risk of damaged equipment); commissioning HV switchgear --- Very High without arc flash assessment and isolation regime. \textbf{Step~4 (Controls):} Goal-post restriction at 3.5\,m height on eastern site access route; plant exclusion zone 9\,m from line centre enforced by barriers and ground markings; overhead line permit for any crane lift approaching within 15\,m of the line centre; HSG47-compliant CAT/Genny survey and trial holes before all groundworks in the unmapped cable zone; 110V CTE enforced for all portable tools; PAT testing register for all SDB-connected equipment; LOTO system with multi-lock hasps and individual padlocks for all SDB and electrical circuit work; arc flash study commissioned from HV specialist in month 9; arc-rated PPE specified for commissioning team before energisation. \textbf{Step~5 (Record):} Assessment signed by project manager and electrical contractor QS; review points scheduled at months 3, 6, 9, and 11 (pre-commissioning). Residual risk rated Medium for portable tools with 110V CTE system in place; overhead line risk rated High pending confirmation of DNO diversion feasibility (diversion sought at pre-application stage).}

%---------------------------- SECTION ----------------------------
\section{Applying the hierarchy of controls}
\label{sec:elec_hierarchy}

\paragraph{}
The hierarchy of controls, as established by Schedule~1 to the Management of Health and Safety at Work Regulations 1999 and elaborated throughout HSE guidance, provides the evaluative framework for selecting and justifying electrical safety measures. Because the consequence of electrical incidents is so frequently catastrophic and irreversible, the default position for any electrical hazard on a construction site must be to exhaust the higher levels of the hierarchy before resorting to lower-level administrative controls or PPE. Live working --- the ultimate failure to apply the hierarchy --- should be regarded as an exceptional measure of last resort, permissible only where it is genuinely unreasonable in all the circumstances to carry out the work dead, and only under a formal live working permit issued and supervised by a competent authorised person. The assessment must document the evaluative process by which higher-order controls were considered and either implemented or rejected with clear justification.

\begin{itemize}
  \bitem{Elimination}{The most effective control is to eliminate the electrical hazard entirely. Overhead line diversions or de-energisation, arranged in advance with the Distribution Network Operator, remove the overhead line hazard at source rather than relying on plant operators to remain within exclusion zones throughout the project. Phased energisation of the permanent supply --- so that circuits are only energised at the point at which they are fully completed and required --- eliminates the hazard of inadvertent live working in incomplete areas of the building. Battery-powered cordless tools, where technology permits equivalent performance, eliminate dependency on mains-supplied power cables and remove the cable trip hazard, mechanical damage risk, and shock hazard simultaneously. Formal diversion and decommissioning of underground electrical cables before groundworks commence, where feasible and proportionate, removes the buried cable strike hazard entirely and is always preferable to managing around existing live cables.}

  \bitem{Substitution}{Where elimination is impracticable, substitution of a higher-risk energy source with a lower-risk equivalent is the preferred control. The mandatory use of 110V CTE equipment in place of 230V single-phase equipment for portable tools is the most significant and widely practised substitution control in United Kingdom construction, mandated by HSG141 and reflected in site rules across the industry. The CTE system ensures that the maximum earth fault voltage is 55\,V, substantially reducing --- though not eliminating --- the severity of a shock event. Similarly, the use of battery-powered or 110V LED temporary lighting systems on scaffolds, in excavations, and in confined spaces substitutes a lower-risk technology for higher-risk 230V mains-fed halogen or fluorescent lighting without compromising the illumination required for safe working. 12\,V or 24\,V supply systems may be considered for specific high-risk environments such as confined spaces or permanently wet excavations.}

  \bitem{Engineering controls}{Engineering controls place physical barriers or automatic protective devices between the worker and the electrical hazard. They include: residual current devices at the SDB level, at individual circuit breakers, and as plug-in in-line RCDs for additional protection; double-insulated Class~II construction of portable tools; mechanical interlocks on switchgear that prevent simultaneous access to live parts and working space; padlocked and restricted-access switchrooms and distribution board enclosures; physically enforced goal-post restrictions at overhead line crossings with robust height limitation barriers; armoured cable protection covers and purpose-designed cable bridges across vehicle and pedestrian routes; and arc flash guarding screens on energised switchgear during commissioning activities. Engineering controls are inherently superior to procedural controls because they do not depend upon worker behaviour, memory, or sustained attention for their effectiveness.}

  \bitem{Administrative controls}{Where engineering controls alone are insufficient, administrative measures must be layered on top. These include: the permit-to-work and LOTO system for all electrical isolation work; restricted access procedures for switchrooms and energised areas; site rules prohibiting the use of 230V domestic extension leads and tools not issued through the site stores; mandatory electrical awareness induction for all site personnel on first arrival regardless of trade; enhanced supervision of young workers, apprentices, and operatives new to construction on all activities near electrical equipment; restricted plant movement zones under and near overhead lines enforced by banksmen and physical barriers; the formal PAT testing regime with documented register; and the schedule of formal electrical inspection and test required by BS~7671 Section~704 and HSG141. Administrative controls must be subject to regular audit to confirm that actual site practice matches the documented procedures.}

  \bitem{PPE}{Personal protective equipment is the last line of defence against electrical hazard and must never be relied upon as the primary or sole control measure. Relevant electrical PPE includes: insulated gloves to BS EN~60903 tested at appropriate voltage class for live or prospective live working; arc flash face shields with appropriate arc rating (expressed in cal/cm\textsuperscript{2}) determined by arc flash energy calculation from a qualified electrical engineer; arc-rated flame-resistant (FR) coveralls and clothing for high-voltage switching, commissioning, and testing activities where the residual risk of arc flash cannot be fully eliminated; safety footwear with an electrical hazard rating; and insulated hand tools to BS EN~60900. Electrical PPE must be selected to match the specific hazard and energy level present, must be regularly inspected for damage and replaced when defective, and must be worn correctly and consistently throughout the work activity --- a requirement that demands active supervision and training rather than assumption that operatives will self-manage.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Cadence of assessment and review triggers}
\label{sec:elec_cadence}

\paragraph{}
An electrical safety risk assessment for a construction project cannot be treated as a static document completed at project mobilisation and filed until project completion. The dynamic nature of construction --- new phases commencing, new trades arriving with different equipment requirements, temporary supplies being extended and modified, and the progressive energisation of permanent circuits during the commissioning phase --- means that the electrical hazard profile of the site changes continuously throughout the project's life cycle. The responsible person, typically the principal contractor's site manager or a nominated competent electrical safety officer, must maintain the assessment as a live document, updated in response to changes in site conditions, and formally reviewed at each of the trigger events listed below. The review schedule must be documented in the Construction Phase Plan and communicated to all relevant parties including the electrical contractor, other principal subcontractors, and the CDM co-ordinator function.

\begin{itemize}
  \bitem{Phase change or major scope change}{Whenever the site progresses to a new construction phase --- from groundworks to superstructure, from structure to first-fix, from first-fix to second-fix, from fit-out to commissioning --- the electrical supply arrangements change and new hazards are introduced. A formal assessment review must be completed before or at the commencement of each new phase, with the updated assessment communicated in the phase-start toolbox talk.}

  \bitem{Commissioning of permanent supply}{The transition from temporary construction supply to energised permanent supply represents the highest-risk period in the site's electrical life cycle. Permanent high-voltage switchgear is energised, HV proving tests are carried out under the DNO's supervision, and construction workers may still be present in the building carrying out snagging and defects activities. A full assessment review supported by an arc flash assessment from a competent high-voltage electrical engineer is mandatory before commissioning activities commence. Zonal access restrictions must be physically implemented before energisation.}

  \bitem{Change of electrical contractor or key competent personnel}{When the electrical subcontractor changes, or when the individual competent person responsible for the temporary installation leaves the project, the incoming party must formally review, adopt, and countersign the existing assessment or produce a replacement. Continuity of knowledge about the site's electrical infrastructure cannot be assumed to transfer automatically through commercial handover processes.}

  \bitem{Any electrical incident, near miss, or dangerous occurrence}{Any electric shock, burn, cable strike, equipment flashover, or near miss with an overhead line must trigger an immediate assessment review. Dangerous occurrences specified in Schedule~1 to the Reporting of Injuries, Diseases and Dangerous Occurrences Regulations 2013 (RIDDOR) --- including electrical short circuits attended by fire or explosion and overhead line contacts by plant --- must be reported to the HSE within prescribed timescales and internally investigated, with the assessment updated to reflect the revised understanding of risk before work in the affected area resumes.}

  \bitem{Adverse weather or site flooding}{Water ingress to electrical equipment through heavy rainfall, flood water, condensation in unheated buildings, or compromised weatherproofing of distribution board enclosures substantially increases both the probability and severity of electrical incidents. Following any period of significant adverse weather or site flooding, a competent inspection of the temporary electrical installation must be carried out before the supply is reinstated, and the assessment reviewed in light of the inspection findings.}

  \bitem{Periodic inspection and test outcomes}{BS~7671 Section~704 and HSG141 require formal inspection and test of the temporary site installation at prescribed intervals --- typically three-monthly for the main distribution system, monthly for individual circuits, and before first use and at three-monthly intervals for portable equipment and extension leads. Any deficiency identified during formal inspection --- an RCD that fails to trip within the prescribed disconnection time, insulation resistance below the 1\,M$\Omega$ minimum, or damaged cable or equipment --- must trigger an assessment review and immediate remedial action, including isolation of the defective circuit or equipment pending repair or replacement.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Common failure modes}

\begin{itemize}
  \bitem{Failure to enforce the 110V CTE rule}{The single most common electrical safety failure on United Kingdom construction sites is the tolerance, whether active or passive, of 230V domestic extension leads and tools sourced by tradespeople from their own homes or domestic suppliers. This failure mode is exacerbated by an inadequate number of 110V socket outlets at site distribution boards, a reluctance by supervisors to challenge established behaviour patterns, and a cultural normalisation of the practice among experienced operatives who ``have always done it this way.'' The effective remedy requires three concurrent actions: adequate 110V supply infrastructure provided as a condition of mobilisation; a zero-tolerance site rule enforced with consistent disciplinary consequences; and daily enforcement checks by line management.}

  \bitem{Inadequate or bypassed RCD protection}{RCDs are frequently found to be inoperative in practice because the test button has not been operated regularly (HSG141 recommends the push-to-test operation be carried out before each use by the user), because the device has been bypassed using a bridge connector or rewired neutral connection to prevent nuisance tripping caused by an underlying wiring fault, or because the RCD is fitted at the SDB but the circuit is fed from an upstream source that bypasses it. A management system that treats RCD push-to-test as a daily pre-use inspection check, equivalent to a plant operator's daily walkaround, removes the dependency on periodic maintenance to detect this failure.}

  \bitem{Failure to isolate before working on electrical equipment}{Live working, whether carried out through ignorance of the requirement, time pressure from the programme, casual over-confidence accumulated over years of near-misses, or a misunderstanding of what constitutes electrical work, is consistently identified as the proximate cause of electrical fatalities and serious injuries on UK construction sites. The Electricity at Work Regulations 1989 Regulation~14 dead working principle must be given operational life through a robust permit-to-work and LOTO system that is both documented and physically enforced. The most dangerous scenario is the experienced operative who believes that isolation ``takes too long'' and who has worked near or on live circuits many times without incident --- until the occasion when the outcome is fatal.}

  \bitem{Poor cable management and mechanical damage}{Cables trailing across pedestrian access routes, hanging unsupported from scaffold tubes or projecting reinforcement, routed through doorways across sharp edges, and driven over repeatedly by site vehicles are ubiquitous on poorly managed construction sites. Mechanical damage to cable insulation creates immediate shock and fire risk. Cable management must be specified in the site electrical plan and enforced through regular inspection by a responsible person. HSG141 provides detailed guidance on cable routing, minimum support intervals, protection at floor penetrations, and the use of proprietary cable protection bridges across vehicle routes.}

  \bitem{Inadequate competence assessment for electrical work}{The Electricity at Work Regulations 1989 Regulation~16 requires that no person shall engage in work activities on or near electrical systems unless they possess the technical knowledge or experience to prevent danger, or are under the appropriate supervision of a person who does. In practice, competence is frequently interpreted too broadly, with operatives accepted as competent on the basis of their own uncorroborated self-declaration or a generic industry card that does not specifically evidence knowledge of the site's electrical systems. Competence for construction electrical work should be evidenced by formal qualification (City and Guilds 2382 for BS~7671; JIB or SJIB grading card; relevant NVQ), documented relevant experience, and a documented briefing on site-specific conditions including the temporary installation layout, SDB locations, overhead line constraints, and the permit-to-work system.}

  \bitem{Failure to obtain and act upon records of underground electrical services before excavation}{The buried cable strike hazard in groundworks is addressed comprehensively in Chapter~\ref{chap:Excavation and Groundworks Risk Assessment}, but it bears specific emphasis in the electrical context that a significant proportion of cable strikes occur not because the CAT and Genny procedure was applied and failed, but because it was not applied at all: utility records were not obtained before mobilisation, no signal was applied to the buried cable to assist detection, or the trial hole programme was omitted under programme pressure. The electrical safety risk assessment must impose a non-negotiable procedural gate requiring verified utility records, CAT/Genny survey, and trial hole confirmation before any mechanical excavation proceeds in areas where live underground electrical cables are identified or suspected.}
\end{itemize}

\example{Failure Pattern}{An investigation by the HSE into a fatal electrical incident on a large industrial construction project identified the following chain of interconnected failures. The site had a 110V CTE policy documented in the Construction Phase Plan, but the electrical subcontractor had provided only six 110V socket outlets across four site distribution boards for a project with over 80 operatives regularly using power tools. The shortfall in 110V supply infrastructure was identified during the initial site set-up but not escalated to the principal contractor's management as a compliance issue, because no individual had been assigned specific responsibility for verifying that the electrical supply infrastructure met the assessment's requirements before trades were permitted to mobilise. The inevitable consequence was that demand for 230V tool supply was met by the informal and progressive introduction of 230V multi-way extension blocks connected via 16A adaptors into 230V outlets provided for the temporary site offices. These 230V extension blocks proliferated through the building under construction over a period of several weeks, with no RCD protection beyond a 100\,mA device at the SDB level --- a tripping threshold entirely inadequate for personal protection against electric shock. A first-fix plumber working on the second floor connected a 230V drill to one of these extension blocks and received a fatal electric shock when the drill's field winding failed and fault current passed through his body to the wet concrete floor. The assessment failure was not one of technical knowledge but of implementation governance: the assessment existed on paper but had never been translated into verifiable physical compliance on the ground.}

%---------------------------- CHAPTER SUMMARY ----------------------------
\newpage
\section{Chapter Summary}

\begin{table}[H]
\centering
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm}X}
\toprule
\textbf{Assessment Element} & \textbf{Operational Requirement} \\
\midrule
Legal/Professional Basis & Electricity at Work Regulations 1989; BS~7671:2018+A2:2022 (IET Wiring Regulations, 18th Edition); HSG141 (Electrical Safety on Construction Sites); GS6 (Avoiding Danger from Overhead Power Lines); HSG47 (Avoiding Danger from Underground Services); PUWER 1998; CDM 2015. \\
Method & Apply the full five-step process and document inherent/residual risk. Electrical hazards must be assessed specifically for electricians, non-electrical tradespeople, plant operators, and groundworkers. The assessment must explicitly address overhead lines, buried cables, temporary supply infrastructure, and arc flash during commissioning. \\
Controls & 110V CTE system mandatory for all portable tools; 30\,mA RCDs on individual circuits; LOTO and permit-to-work for all electrical isolation; goal-post crossing restrictions and exclusion zones for overhead lines; CAT/Genny survey and trial holes for all groundworks near buried cables; PAT testing at 3-month intervals; arc flash assessment before HV commissioning; competence verification for all electrical personnel. \\
Cadence & Review at each construction phase change; mandatory review before commissioning of permanent supply; review after any electrical incident, near miss, or RIDDOR dangerous occurrence; review after adverse weather or site flooding; review following periodic inspection and test results. \\
Assurance & Use audit, incident learning, and governance reporting to verify effectiveness. Maintain BS~7671 Section~704 inspection and test records; PAT test registers; permit-to-work logs; RCD test records; and overhead line permit logs as part of the project safety file. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{itemize}
  \bitem{Key takeaway 1}{The 110V centre-tapped-to-earth system is the single most important engineering control for reducing portable tool electric shock risk on construction sites. Its effectiveness depends entirely upon comprehensive enforcement: adequate 110V socket outlet provision at all SDBs, zero tolerance of 230V domestic equipment, and consistent challenge of non-compliance by all levels of site management.}

  \bitem{Key takeaway 2}{Live working must be treated as an absolute last resort under the Electricity at Work Regulations 1989 Regulation~14. The permit-to-work and lock-off/tag-out system must be genuinely implemented and physically enforced using individual padlocks and danger tags --- not simply documented in a policy that is overridden under programme pressure.}

  \bitem{Key takeaway 3}{Overhead line contacts by cranes, telehandlers, concrete pump booms, and raised vehicle bodies continue to kill construction workers every year. HSE GS6 requires a hierarchy from diversion through goal-post restrictions to exclusion zones. The assessment must specify which controls apply to each piece of plant and confirm that they have been physically implemented and verified before any operations commence near overhead lines.}

  \bitem{Key takeaway 4}{The competence requirement under the Electricity at Work Regulations 1989 Regulation~16 must be operationalised through verifiable qualification checks, not accepted on the basis of an operative's uncorroborated self-declaration. JIB grading cards, SJIB registration, and formal qualifications in BS~7671 provide the evidential basis for competence verification that will withstand regulatory scrutiny.}

  \bitem{Key takeaway 5}{The commissioning phase --- when permanent high-voltage and low-voltage systems are first energised in a building that may still contain construction workers --- is the highest electrical-risk period of any construction project. A dedicated arc flash assessment, formal commissioning plan, zonal access restrictions, and arc-rated PPE must all be in place and verified before any energisation activity commences.}
\end{itemize}

\barquote{In Chapter~\ref{chap:Hot Works and Fire Risk Assessment}, the assessment methodology is applied to hot works and fire risk --- a hazard category that shares with electrical safety the capacity to initiate catastrophic events from apparently minor lapses in procedural control, and which demands an equally rigorous permit-based approach to managing ignition sources and combustible materials on construction sites.}
