\chapter{Traffic Management and Logistics Risk Assessment}
\label{chap:Traffic Management and Logistics Risk Assessment}

%---------------------------- SECTION ----------------------------
\section{Why this assessment matters in construction}

\paragraph{The interaction between moving vehicles and pedestrians on construction sites is one of the most persistently lethal hazard combinations in the UK construction industry. HSE fatal injury statistics over the last two decades consistently identify being struck by a moving vehicle as one of the most common causes of death on construction sites, and the pattern is remarkably uniform: a worker on foot is in the path of a reversing or turning vehicle whose driver cannot see them; the vehicle is an excavator, dumper truck, concrete mixer, or delivery lorry; the worker is in an area they have the right to be in; and there is no physical separation between the pedestrian route and the vehicle operating zone. The vehicle does not fail; the driver does not act recklessly; the operative does not do something obviously wrong. The accident happens because the system that should have physically prevented the interaction did not exist, or existed on paper but was not implemented or maintained.}

\paragraph{Construction sites are complex, dynamic logistics environments. Materials arrive by heavy goods vehicles at unpredictable intervals and must be unloaded, moved, and distributed across the site. Plant --- excavators, dumpers, telehandlers, piling rigs, concrete pumps, mobile cranes --- operates across the site in sequences that change day by day as the project progresses. Operatives of all trades move across the site on foot to reach their work areas, welfare facilities, and emergency exits. The concentration of large, slow-manoeuvring vehicles with significant dead zones in a working environment with multiple pedestrians creates a hazard profile that demands a structured, designed, and enforced traffic management scheme from the first day of site establishment. The Construction Phase Plan required by CDM 2015 must include a traffic management plan as a primary safety document --- not a reference to the principal contractor's generic procedure, but a site-specific, phased plan that evolves as the project progresses and site conditions change.}

\paragraph{The hazard is compounded by the diversity of vehicle types and the variability of driver competence. A major construction project may have on site simultaneously: 32-tonne articulated concrete mixers; 10-tonne rear-tip dumpers; 30-tonne tracked excavators; 8-wheel tipper lorries; forklift trucks; telehandlers; van deliveries; specialist plant including piling rigs and concrete pumps; and site cars and minibuses. Each vehicle type has a distinct operational envelope, different dead zones, different braking distances, and different turning circles. The risk assessment must address each vehicle type specifically, not rely on a generic ``vehicles and pedestrians must be kept separate'' statement. Banksman provision, required for reversing manoeuvres by large vehicles, is itself a safety-critical task that requires training and competence: a banksman who misunderstands the dead zones of the vehicle they are guiding, or who places themselves in the vehicle's travel path, compounds rather than reduces the risk.}

\barquote{``The separation of vehicles and pedestrians on a construction site is not a matter of preference or good practice --- it is the primary engineering control without which no amount of banksman provision, high-visibility clothing, or awareness training will reduce the risk to an acceptable level.''}

\example{Reversing Dumper Fatality}{A groundworks contractor is operating a 10-tonne articulated rear-tip dumper on a large housing development. The site traffic management plan designates a pedestrian walkway along the eastern side of the site separated from the vehicle routes by plastic barriers. During a busy delivery period, the barriers along a 30-metre stretch of the walkway have been removed by the civil contractor to allow direct access for a concrete pour, and have not been reinstated. A bricklaying operative, returning from the welfare cabin to the brick-laying area, crosses through the gap in the barrier into the vehicle zone. The dumper driver, reversing to a spoil tip, cannot see the operative in his mirrors due to the configuration of the tipping area, and strikes the operative fatally. Investigation reveals that the barrier removal had not been documented or authorised; no temporary alternative pedestrian route had been established; no banksman had been provided for the reversing manoeuvre; and the site induction had not addressed the mechanism by which barriers could be moved. The Traffic Management Plan had been produced but had not been reviewed following the changes in site layout that created the tipping area adjacent to the walkway.}

%---------------------------- SECTION ----------------------------
\section{Legal and professional framework}

\paragraph{The legal framework for traffic management on construction sites is anchored by the Construction (Design and Management) Regulations 2015, which require the principal contractor to plan, manage, and coordinate the construction phase in a way that controls vehicle and pedestrian interaction, and to produce and maintain a Construction Phase Plan that addresses traffic management. The Management of Health and Safety at Work Regulations 1999 require a suitable and sufficient risk assessment of all vehicle and pedestrian interaction hazards. The Workplace (Health, Safety and Welfare) Regulations 1992 apply to construction sites from the establishment of site facilities and require that pedestrian and vehicle traffic routes are organised so that pedestrians and vehicles can circulate in a safe manner. Key frameworks relevant to this chapter include: Health and Safety at Work etc.\ Act 1974; Management of Health and Safety at Work Regulations 1999; Construction (Design and Management) Regulations 2015; Workplace (Health, Safety and Welfare) Regulations 1992; PUWER 1998; Highways Act 1980; Road Traffic Act 1988; HSE guidance HSG136 (Workplace Transport Safety); and the industry guidance document ``Workplace Transport Safety in Construction'' (CIRIA/HSE).}

\begin{itemize}
  \bitem{Construction (Design and Management) Regulations 2015}{Place on the principal contractor the specific duty to plan and control the construction phase, including the management of vehicles and pedestrians. The Construction Phase Plan must include arrangements for site layout to ensure that vehicle and pedestrian routes are physically separated, that delivery and unloading areas are managed safely, and that the Traffic Management Plan is maintained and updated as the project progresses. CDM 2015 also places a duty on designers to consider traffic management in their designs, and on clients to provide pre-construction information about access constraints, highway restrictions, and third-party traffic interfaces.}
  \bitem{Workplace (Health, Safety and Welfare) Regulations 1992}{Regulation~17 requires that traffic routes are organised so that pedestrians and vehicles can circulate in a safe manner, and that sufficient separation is provided between pedestrian routes and vehicle routes. The regulations require that traffic routes are of sufficient width and headroom for the vehicles and pedestrians using them, that they are clearly marked and well-lit, and that sufficient passing places are provided on two-way routes. Construction sites are workplaces for the purposes of these regulations, and they apply from the moment the site is established.}
  \bitem{PUWER 1998}{Applies to all vehicles used at work on construction sites. PUWER requires that vehicles are suitable for their intended purpose, maintained in a safe condition, inspected at appropriate intervals, and operated only by trained and competent persons. PUWER Regulation~26 specifically addresses mobile work equipment, requiring that vehicles used on site are fitted with rollover protection, falling-object protection, and restraining systems where the risk assessment identifies the need.}
  \bitem{HSG136 --- Workplace Transport Safety (HSE)}{The primary HSE guidance document on workplace transport safety, applicable to construction sites. HSG136 provides a framework for assessing vehicle and pedestrian interaction risks, designing safe traffic routes, selecting appropriate vehicles, and managing reversing operations. The guidance emphasises that physical separation of vehicles and pedestrians is the primary control, and that reliance on high-visibility clothing alone is not an adequate risk control strategy.}
  \bitem{Road Traffic Act 1988 and Highways Act 1980}{Apply to vehicles using the public highway in connection with the construction project, including delivery vehicles, skip lorries, and plant movements. The principal contractor has responsibilities for the management of site access onto the public highway, including the provision of traffic management for highway works under the New Roads and Street Works Act 1991, the maintenance of the visibility splay at the site entrance, and the management of mud and debris tracking onto the highway.}
  \bitem{The Highway Code (Rule~32 for large vehicles and their blind spots)}{While not directly enforceable on private construction land, the Highway Code's guidance on large vehicle blind spots is relevant to the training of site operatives and the understanding of vehicle operating envelopes. The three-point mirror system on large construction vehicles, and the dead zones that remain even when this system is correctly adjusted, must be understood by banksmen and all site personnel.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Conducting the assessment using the five-step method}

\subsection{Step 1 --- Identify Hazards}
\paragraph{Traffic management hazards on construction sites fall into five categories: pedestrian/vehicle conflict; vehicle/vehicle conflict; uncontrolled delivery and unloading; poor site visibility and lighting; and interface with the public highway. Pedestrian/vehicle conflict arises wherever pedestrian routes and vehicle routes share the same space --- at site entrances, in delivery areas, adjacent to excavations where ground conditions force operatives onto vehicle routes, and where temporary access gaps are created through traffic management barriers without alternative pedestrian routes. Vehicle/vehicle conflict arises at crossing points, in narrow accesses where two-way vehicle movement is permitted without control, and in delivery turning areas where articulated vehicles manoeuvring into position conflict with site traffic passing through. Uncontrolled delivery and unloading hazards arise from: delivery drivers descending from cabs to unload materials without banksman attendance; forklift operations in areas where pedestrians are passing; crane-assisted unloading without adequate exclusion zones; and concrete mixer, pump, and tanker operations without driver briefing on site rules and restricted areas. Visibility hazards include: poor lighting in areas used by both plant and operatives; dust and rain reducing plant operator visibility; overhead obstructions limiting load height in unloading areas; and reflections and sun position reducing driver visibility at certain times of day.}

\subsection{Step 2 --- Decide Who or What Is Affected}
\paragraph{Persons at risk from site traffic include all site operatives, regardless of trade; delivery drivers from external hauliers who are unfamiliar with the site layout; visitors and clients who may not have received a full site induction; public authority inspectors and surveyors attending the site; and members of the public near the site entrance where delivery vehicles are entering and leaving. The client's representatives attending progress meetings are at particular risk if they are escorted directly into site without high-visibility PPE and without awareness of the traffic management arrangements. Sub-contractor operatives new to site are disproportionately represented in vehicle/pedestrian incidents, reflecting the importance of site induction in communicating the traffic management scheme.}

\subsection{Step 3 --- Evaluate Likelihood and Consequence}
\paragraph{The consequence of a struck-by incident involving a 10-tonne or heavier vehicle is virtually always fatal or life-changing, and the consequence rating is Catastrophic across all vehicle types from a telehandler upward. The likelihood of the interaction occurring, in the absence of physical separation, must be rated as High on any active construction site given the density of vehicle movements and pedestrian activity. The combination of Catastrophic consequence and High likelihood in the absence of controls places site traffic as one of the highest overall residual risks on any construction site, and the risk assessment must reflect this.}

\subsection{Step 4 --- Select and Implement Controls}
\paragraph{Controls for traffic management must begin with physical separation as the primary measure. A Traffic Management Plan must be produced for the site, reviewed and approved by the principal contractor and incorporated into the Construction Phase Plan. The plan must: identify all vehicle routes and pedestrian routes; specify the physical separation measures between them (Armco barriers, concrete blocks, Heras fencing, painted markings where physical barriers are impractical); identify all reversing manoeuvres and specify banksman requirements; designate delivery reception areas with exclusion zones during unloading; specify the site speed limit; identify areas where plant must be fitted with reversing alarms and camera systems; and specify the induction requirements for delivery drivers and visitors. The plan must be reviewed and updated as the site layout changes, and changes to the plan must be communicated to all site personnel before implementation.}

\subsection{Step 5 --- Record and Review}
\paragraph{The Traffic Management Plan must be a live document within the Construction Phase Plan. Traffic management inspections must be recorded as part of the weekly site safety inspection. Near-miss incidents involving vehicles must be reported, investigated, and actioned. Delivery driver induction records must be maintained. Vehicle pre-use inspection records must be retained. Any change to site layout that affects the Traffic Management Plan must be documented, reviewed, and approved before barriers are moved or routes altered.}

\example{Five-Step Application}{A principal contractor is managing a city-centre demolition and new-build project. The site has a single access point shared by all vehicles and pedestrians. The site is bounded by live highways on three sides. \textbf{Step~1 (Hazards):} Delivery vehicle/pedestrian conflict at site gate; reversing articulated vehicles in unloading zone; forklift operations in materials storage area; mud and debris tracking onto highway creating hazard to highway users. \textbf{Step~2 (Who Affected):} All site operatives, delivery drivers, site visitors, highway users. \textbf{Step~3 (Evaluation):} Vehicle/pedestrian interaction rated Catastrophic/High inherent risk. \textbf{Step~4 (Controls):} Dedicated pedestrian entrance with physical separation from vehicle entrance; banksman on gate to manage deliveries; concrete block barriers defining one-way vehicle circuit within site; banksman required for all reversing operations; delivery booking system to control vehicle arrival rate; vehicle wash at gate exit; speed limit 5 mph with speed humps; no delivery vehicles permitted when crane is lifting. \textbf{Step~5 (Record):} TMP in CPP reviewed monthly; gate banksman daily log; weekly traffic management inspection record; delivery driver induction log retained.}

%---------------------------- SECTION ----------------------------
\section{Applying the hierarchy of controls}

\paragraph{The hierarchy of controls for traffic management must prioritise physical separation over behavioural and PPE-based controls. High-visibility clothing is not a traffic management control --- it is the last line of defence after all other controls have been implemented.}

\begin{itemize}
  \bitem{Elimination}{Design the site layout to eliminate reversing manoeuvres wherever possible, using one-way circuits. Locate delivery reception areas, materials storage, and welfare facilities to minimise the number of pedestrian/vehicle crossing points. Where the project programme permits, schedule heavy plant operations and deliveries at times when the number of operatives on foot is minimised.}
  \bitem{Substitution}{Where reversing is unavoidable, substitute operator-only reversing with banksman-supervised reversing, or camera-assisted reversing systems. Consider pedestrian access routes that use overhead crossings or underpasses where vehicle density makes ground-level separation difficult.}
  \bitem{Engineering controls}{Physical barriers between vehicle and pedestrian routes are the primary engineering control. Concrete blocks, Armco barriers, and heavy-duty fencing provide physical protection that cannot be defeated by inattention or pressure of work. One-way traffic circuits eliminate head-on vehicle conflict. Speed humps and road narrowing enforce the site speed limit physically. Reversing cameras and proximity detection systems on large plant reduce dead-zone risk for reversing operations. Vehicle-activated warning systems at pedestrian crossing points alert pedestrians to approaching vehicles.}
  \bitem{Administrative controls}{Traffic Management Plan within the Construction Phase Plan, reviewed and maintained. Delivery booking system to control vehicle arrival. Banksman deployment at all reversing operations and site gate. Site induction for all personnel including delivery drivers. Prohibition of pedestrian access to vehicle-only zones. Tool-box talks on traffic management. Incident and near-miss reporting system for traffic events.}
  \bitem{PPE}{High-visibility clothing (minimum Class 2 vest or jacket) is mandatory for all persons on the construction site. High-visibility clothing does not constitute a traffic management control but forms part of the system. Appropriate footwear and hard hat at the site entrance before entering the vehicle zone.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Cadence of assessment and review triggers}

\paragraph{The Traffic Management Plan must be treated as a living document that is formally reviewed at the frequency specified in the Construction Phase Plan and updated following any change in site conditions.}

\begin{itemize}
  \bitem{At the start of each phase of works}{Each new phase of the project --- groundworks, structural, envelope, fit-out --- changes the pattern of vehicle movements and the positions of operatives on site. The Traffic Management Plan must be reviewed before each phase change and updated to reflect the new vehicle circulation pattern, new delivery requirements, and new pedestrian routes.}
  \bitem{Following any change to site layout}{Any change to the site layout that moves barriers, pedestrian routes, delivery areas, or plant operating zones must trigger a review and update of the Traffic Management Plan. Changes must be communicated to all site personnel before implementation.}
  \bitem{Following any vehicle/pedestrian near miss or incident}{Any near-miss between a vehicle and a pedestrian, any vehicle striking a barrier or infrastructure element, and any vehicle-related injury must trigger an immediate formal review of the Traffic Management Plan. The investigation must determine whether the controls in the plan were adequate, whether they were implemented, and what changes are required.}
  \bitem{Weekly traffic management inspection}{A formal documented traffic management inspection must be conducted weekly as part of the site safety inspection, checking that barriers are in position, pedestrian routes are clear and unobstructed, reversing alarm and camera systems are operational, and banksman deployment is being maintained.}
  \bitem{Following change in delivery volumes or vehicle types}{If the project's delivery programme changes significantly --- for example, a period of intensive concrete pours requiring a high frequency of mixer deliveries --- the Traffic Management Plan must be reviewed to confirm that the controls are adequate for the increased vehicle density.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Common failure modes}

\begin{itemize}
  \bitem{Traffic Management Plan produced but not maintained}{A Traffic Management Plan produced at project inception and never updated fails to address the changing site conditions across the project lifecycle. Site layout changes, new plant introductions, programme changes, and phase transitions each alter the risk profile; a static plan becomes progressively less relevant and less safe as the project evolves.}
  \bitem{Barriers removed without authorisation and not reinstated}{Physical separation barriers are removed to facilitate a specific operation --- a concrete pour, a crane lift, a delivery --- without authorisation, without a temporary alternative pedestrian route, and without a procedure for reinstatement. The removal creates an uncontrolled vehicle/pedestrian interaction zone, often for a period extending well beyond the original removal need.}
  \bitem{Banksman requirements not enforced for reversing operations}{The Traffic Management Plan specifies banksman attendance for reversing operations but supervision is insufficient to enforce compliance. Experienced drivers familiar with the site routinely reverse without banksman attendance, and the practice is accepted as normal. The event that causes a fatality is no different from any of the hundreds of unsupervised reversing operations that preceded it.}
  \bitem{Delivery drivers not inducted into the Traffic Management Plan}{Delivery drivers arriving at site for the first time, or drivers familiar with a previous site layout, are permitted to enter site without briefing on the current traffic management arrangements. Delivery drivers are disproportionately involved in construction site vehicle incidents, reflecting the unfamiliarity with site conditions.}
  \bitem{Reliance on high-visibility clothing as the primary traffic management control}{High-visibility clothing is the last line of defence in a layered control system, not the primary control. Sites that rely on high-visibility clothing to manage vehicle/pedestrian risk, without physical separation and route design, are operating without adequate controls. High-visibility clothing does not prevent a reversing vehicle whose driver cannot see the operative from striking them.}
  \bitem{No speed control measures on site haul roads}{Internal site haul roads with no physical speed control measures --- no speed humps, no narrowing, no signage enforced by supervision --- create conditions in which plant operators and delivery drivers travel at speeds that significantly extend their stopping distances and reduce the time available to avoid a pedestrian. The 5 mph site speed limit is the standard for UK construction sites; it must be enforced by physical measures, not signage alone.}
\end{itemize}

\example{Failure Pattern}{A fit-out contractor on a large commercial development creates an informal shortcut through the delivery vehicle holding area to reduce the walk time from the car park to the welfare cabin. The shortcut has been used informally for several weeks. The site manager is aware of it but has not formalised it or provided physical separation. During a period of intensive deliveries associated with a curtain wall installation, a 32-tonne flat-bed delivery lorry reverses into the holding area using a satellite navigation aid only, without a banksman. The driver cannot see an operative using the informal shortcut in his mirrors. The operative is struck and sustains serious leg injuries requiring amputation. Investigation reveals that the informal pedestrian route was known to site management; that the delivery area was not designated as a banksman-required reversing zone in the Traffic Management Plan; and that the volume of deliveries associated with the curtain wall installation had not triggered a review of the TMP. Both the absence of physical separation in the improvised route and the failure to review the TMP in response to programme change are identified as causal factors.}

%---------------------------- CHAPTER SUMMARY ----------------------------
\newpage
\section{Chapter Summary}

\begin{table}[H]
\centering
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm}X}
\toprule
\textbf{Assessment Element} & \textbf{Operational Requirement} \\
\midrule
Legal/Professional Basis & CDM 2015; Management Regulations 1999; Workplace (Health, Safety and Welfare) Regulations 1992; PUWER 1998; HSG136; Health and Safety at Work etc.\ Act 1974. \\
Method & Apply the full five-step process and document inherent/residual risk. Produce and maintain a site-specific Traffic Management Plan within the Construction Phase Plan. \\
Controls & Physical separation of vehicle and pedestrian routes as the primary control. One-way vehicle circuits to eliminate reversing where possible. Banksman for all unavoidable reversing operations. Delivery booking system and driver induction. Weekly traffic management inspection. \\
Cadence & Review at each phase change; following any layout change; following any near miss or incident; weekly inspection; following change in delivery volume or vehicle types. \\
Assurance & Traffic Management Plan within CPP; weekly inspection records; near-miss reporting; delivery driver induction log; plant pre-use inspection records. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{itemize}
  \bitem{Key takeaway 1}{Physical separation of vehicle and pedestrian routes --- by concrete blocks, Armco barriers, or heavy-duty fencing --- is the primary traffic management control. It cannot be substituted by high-visibility clothing, banksman provision, or awareness training.}
  \bitem{Key takeaway 2}{The Traffic Management Plan must be a live document that is reviewed and updated at each phase change and following any change to site layout. A static plan produced at project inception and never revised does not manage the evolving risk.}
  \bitem{Key takeaway 3}{All reversing operations by large plant and delivery vehicles must have banksman attendance. This requirement must be enforced by supervision, not assumed to be self-managing. The delivery holding area, the materials storage area, and all tipping zones are the highest-risk reversing locations.}
  \bitem{Key takeaway 4}{All delivery drivers arriving at site for the first time must receive a briefing on the current Traffic Management Plan, including the pedestrian routes, the one-way circuit, the speed limit, and the banksman protocol. This briefing must be recorded.}
  \bitem{Key takeaway 5}{Any near-miss involving a vehicle and a pedestrian must trigger an immediate formal review of the Traffic Management Plan, regardless of whether injury occurred. Near-misses are precursors to fatalities, and every unacted-upon near-miss increases the probability that the next event will be fatal.}
\end{itemize}

\barquote{In Chapter~\ref{chap:Demolition and Refurbishment Risk Assessment}, we move into Part~5 of this book, examining the risk assessment challenges specific to demolition and refurbishment operations --- a domain in which structural unpredictability, hazardous materials, and the legacy of previous construction activity combine to produce some of the highest-risk working environments encountered in the built environment sector.}
