\chapter{Plant, Vehicles and Site Transport Risk Assessment}
\label{chap:Plant, Vehicles and Site Transport Risk Assessment}

%---------------------------- SECTION ----------------------------
\section{Why this assessment matters in construction}

\paragraph{Mobile plant and site vehicles kill and seriously injure construction workers with a consistency that makes the pedestrian-plant interface one of the most reliably dangerous hazard domains in the built environment sector. Dumpers reversing without warning into workers who cannot hear the reversing alarm over ambient site noise; forklifts turning a blind corner with an obscured load that strikes a pedestrian who had no reason to expect the vehicle at that location; road sweepers entering a public highway from a site entrance without adequate traffic management, resulting in a collision with a cyclist who had no warning of the egress --- each of these represents not an exceptional accident but a predictable consequence of the failure to segregate plant and pedestrians effectively. The HSE's statistics on transport-related construction fatalities show a persistent pattern of struck-by, run-over, and crush injuries that the systematic application of segregation, site layout, and plant management controls can prevent.}

\paragraph{The breadth of the plant and vehicle category in construction is significant. Site plant includes excavators, dumpers, ride-on compaction rollers, tele-handlers, rough terrain forklifts, concrete pumps, concrete mixer trucks, road sweepers, vacuum excavation units, mobile cranes, and tower crane erection equipment. Each of these machines has a distinct visibility profile, a distinct blind-spot geometry, a distinct braking distance, and a distinct consequence of pedestrian contact. The risk assessment for plant and vehicles must not be generic; it must address the specific machines used on the project, the specific site layout, the specific pedestrian movements, and the specific interface with the public highway.}

\paragraph{The interface between site operations and the public road network presents a distinct and frequently underassessed risk category. When a loaded dumper exits a site onto a public road, it is subject to the Road Traffic Act 1988 and the requirements for road-worthy vehicles, licensed drivers, and compliance with highway law. When a road sweeper clears a site access road and then moves to the public highway, the driver transitions from a PUWER-governed site environment to a road traffic law environment in a single gate movement. The risk assessment must address both domains and must confirm that the vehicles used for both site and public-road operations comply with both the PUWER requirements applicable on site and the Road Traffic Act requirements applicable on the public highway.}

\paragraph{Operator visibility is the central technical challenge of pedestrian-plant segregation. A standard construction dumper has blind spots that extend behind the machine for approximately 10 metres at ground level; a tele-handler with an elevated load has forward visibility severely compromised by the load and the mast; a tracked excavator facing forward for travel has no visibility whatsoever of the ground behind its tracks. These are not design deficiencies that can be corrected by better operator training --- they are inherent physical characteristics of the machines. The risk assessment must address operator visibility limitations for each machine on site, and must implement segregation measures that ensure no pedestrian can occupy the blind-spot zones of any operating machine.}

\barquote{``Pedestrian-plant separation is not a nice-to-have site management aspiration --- it is the primary engineering control for a hazard that kills construction workers every year. The risk assessment must specify hard physical separation, not rely on the vigilance of workers or the caution of operators.''}

\example{Dumper Reverse-Over Incident}{A groundworks contractor is using a 6-tonne high-tip dumper to cart excavated material from a basement dig to a spoil storage area at the far end of a 120-metre-long site. The haul route is a shared corridor 4 metres wide that also serves as the pedestrian access from the site welfare facilities to the basement working area. The reversing alarm on the dumper is functioning but produces a sound level that is indistinguishable from ambient site noise at a distance of more than 4 metres. A groundworker walking back to the basement area from the welfare cabin uses the shared haul route, not having received a site induction that specified the welfare cabin exit route. The dumper operator, reversing to position for the tip, cannot see the pedestrian in his 10-metre rear blind spot. The pedestrian is struck and runs over by the dumper's rear axle. The investigation identifies the shared haul route and pedestrian access corridor as the primary causal factor and the absence of physical segregation as the critical control failure.}

%---------------------------- SECTION ----------------------------
\section{Legal and professional framework}

\paragraph{The legal framework for plant and vehicle management on construction sites draws on both occupational health and safety legislation and road traffic law, reflecting the dual operational environment of many site vehicles. The Construction (Design and Management) Regulations 2015 place the organisation of site traffic management and pedestrian-plant segregation within the scope of the Construction Phase Plan, making it a principal contractor responsibility to design and enforce the traffic management arrangements for the project. PUWER 1998 governs the suitability, inspection, and operation of all site plant and vehicles, including the requirement for operator training and competency. The Road Traffic Act 1988 extends the road traffic legal framework to vehicles operating on the public highway, regardless of whether those vehicles are normally used exclusively on construction sites.}

\paragraph{Key frameworks relevant to this chapter include: Provision and Use of Work Equipment Regulations 1998 (PUWER); Workplace (Health, Safety and Welfare) Regulations 1992; Construction (Design and Management) Regulations 2015; Road Traffic Act 1988; Health and Safety at Work etc.\ Act 1974; CLOCS (Construction Logistics and Community Safety) Standard; FORS (Fleet Operator Recognition Scheme) Standard; Management of Health and Safety at Work Regulations 1999.}

\begin{itemize}
  \bitem{Provision and Use of Work Equipment Regulations 1998 (PUWER)}{Requires that all plant and vehicles used on construction sites are suitable for their intended purpose, maintained in an efficient and safe condition, inspected at appropriate intervals, and operated only by trained and competent persons. PUWER Regulation 25 requires that mobile work equipment is stable during use, and Regulation 26 requires that self-propelled work equipment is fitted with devices to prevent an unauthorised start. PUWER Regulation 28 requires that vehicles carrying persons have restraining devices for all passengers.}
  \bitem{Workplace (Health, Safety and Welfare) Regulations 1992}{Regulation 17 requires that traffic routes on construction sites are suitable for the persons and vehicles using them, sufficient in number, in suitable positions, and of sufficient size. The Regulation requires that pedestrian routes and vehicle routes are separated where reasonably practicable, and that where they cannot be separated, adequate provision is made to protect pedestrians from vehicles. The Approved Code of Practice to the Workplace Regulations provides detailed guidance on traffic route design, visibility, speed limits, and segregation measures.}
  \bitem{Construction (Design and Management) Regulations 2015}{Requires the Construction Phase Plan to address arrangements for managing site traffic, including the organisation of deliveries, the positioning of loading and unloading areas, the management of site access and egress, and the segregation of pedestrians from plant and vehicles. The principal contractor must actively manage plant and vehicle movements and must take appropriate action when unsafe traffic management arrangements are identified.}
  \bitem{Road Traffic Act 1988}{Applies to all vehicles operating on public roads, including site dump trucks, road sweepers, and other construction vehicles that may operate both on site and on the public highway. The Act requires that vehicles are road-worthy, insured, driven by licensed drivers, and operated in compliance with all road traffic legislation. It is not a defence to a road traffic prosecution that the vehicle is principally a site vehicle.}
  \bitem{CLOCS Standard}{The Construction Logistics and Community Safety Standard is an industry-led framework requiring construction projects to manage the interactions between construction logistics vehicles and vulnerable road users, including pedestrians, cyclists, and motorcyclists. CLOCS applies to all vehicle movements to and from construction sites and requires the assessment of delivery routes, vehicle specifications (including direct vision standards and sensor packages), and driver training for vulnerability awareness.}
  \bitem{FORS Standard}{The Fleet Operator Recognition Scheme provides a voluntary framework for vehicle fleet operators to demonstrate safe and efficient management of their fleets. FORS Bronze accreditation represents a baseline of compliance with road traffic law and fleet management best practice; FORS Silver and Gold progressively require higher standards of driver training, vehicle specification, and management system maturity. Many principal contractors and project clients specify FORS Bronze as a minimum requirement for all HGVs delivering to or from their sites.}
  \bitem{Management of Health and Safety at Work Regulations 1999}{Requires the assessment of all plant and vehicle movement risks, the appointment of competent persons to manage those risks, and the implementation of control measures proportionate to the risk. Where multiple contractors operate plant and vehicles on the same site, the principal contractor must coordinate the traffic management arrangements to ensure that the controls of one contractor do not undermine the controls of another.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Conducting the assessment using the five-step method}

\subsection{Step 1 --- Identify Hazards}
\paragraph{Plant and vehicle hazards on construction sites can be organised into four primary categories: pedestrian-plant conflict; vehicle-vehicle conflict; vehicle-structure and overhead conflict; and public highway interface. Pedestrian-plant conflict hazards include: pedestrians struck by reversing vehicles in blind-spot zones; pedestrians struck by the swing or travel of excavator buckets; pedestrians crushed between moving vehicles and fixed structures at pinch points; pedestrians struck at site access and egress points; and pedestrians entering vehicle-only zones through inadequate or unenforced barriers. Vehicle-vehicle conflict hazards include: collisions at uncontrolled junctions on haul routes; collisions between plant operating in close proximity on a shared working area; vehicle overrun at the edge of raised platforms, slopes, or basement excavation lips; and loaded vehicles overturning on haul route gradients or cross-falls. Vehicle-structure and overhead conflict hazards include: overhead cables and bridge structures struck by elevated plant; buried services damaged by heavy vehicle trafficking; temporary structures, hoardings, and scaffold standards struck by manoeuvring vehicles; and property damage at the site access from oversized vehicles negotiating tight turns. Public highway interface hazards include: vehicles exiting the site without adequate visibility of highway traffic; mud and debris deposited on the highway creating a skidding hazard for road users; HGV delivery vehicles reversing from the public highway into the site; and construction traffic interacting with vulnerable road users at the site entrance.}

\subsection{Step 2 --- Decide Who or What Is Affected}
\paragraph{Persons at risk from plant and vehicle hazards include every person on or adjacent to the site: the plant operators themselves are at risk from machine overturning and falling objects; groundworkers and other tradespeople are at risk from vehicles operating near their work areas; delivery drivers are at risk during the loading and unloading interface with site plant; site managers and health and safety personnel are at risk when conducting site inspections on active haul routes; and the general public --- pedestrians, cyclists, motorists, and residents --- are at risk at the site boundary and at any point where construction traffic intersects with the public highway. Children attending schools adjacent to construction sites are a particularly vulnerable group and must be specifically addressed where the site is in proximity to a school or nursery.}

\subsection{Step 3 --- Evaluate Likelihood and Consequence}
\paragraph{The consequence of a pedestrian struck by a moving construction vehicle is typically severe: the mass and ground clearance of site plant means that a dumper or telehandler passing over a fallen person is almost invariably fatal. The likelihood of a pedestrian-plant conflict on an uncontrolled site is high: multiple machines operating simultaneously in a shared space with ad-hoc pedestrian routes creates frequent proximity conflicts. The inherent risk of uncontrolled plant and pedestrian interaction must be rated Very High. The effectiveness of physical segregation in reducing inherent risk is high: a site where pedestrian and plant routes are physically separated by permanent barriers, where all plant areas are signed and fenced as exclusion zones, and where a one-way traffic system eliminates vehicle reversing wherever possible can achieve a Low residual risk. The assessment must distinguish between the risk during normal planned operations and the risk during delivery, break, and end-of-shift periods when pedestrian movement is at its highest and plant activity may also be at its peak.}

\subsection{Step 4 --- Select and Implement Controls}
\paragraph{The primary and most effective control for pedestrian-plant conflict is physical separation: a layout that ensures no person on foot can reach the operating envelope of any moving plant without passing through a designated access control point. The site traffic management plan (STMP) --- a document that forms part of the Construction Phase Plan --- must specify: one-way traffic circulation wherever possible to eliminate reversing; designated pedestrian routes with physical barriers (jersey barriers, armco, hoarding panels) separating them from vehicle haul routes; a designated vehicle-free welfare, office, and storage zone; clearly marked pedestrian crossing points with adequate visibility sightlines; maximum site speed limits appropriate for the haul route geometry; the specification and PUWER compliance status of all plant to be used on site; operator competency requirements (CPCS card category for each machine type); a pre-start inspection checklist for all plant; a banksman deployment plan for any manoeuvre that requires reversing in an area accessible to pedestrians; and a delivery management plan that controls the arrival time, marshalling, and unloading of all HGV deliveries to prevent pedestrian-vehicle conflicts at the site entrance.}

\subsection{Step 5 --- Record and Review}
\paragraph{The site traffic management plan must be reviewed at every principal contractor site manager's weekly review meeting and updated whenever the site layout changes, new plant is introduced, or a new phase of work commences that alters the traffic flows on site. Daily pre-start plant checks must be recorded for every machine. Any near-miss between a vehicle and a pedestrian must immediately halt the relevant vehicle operation, trigger a review of the STMP, and require documented sign-off from the principal contractor's senior site management before the operation resumes. Near-misses are lagging indicators of inadequate segregation; they must be investigated and the control failure identified and corrected before the next incident occurs.}

\example{Five-Step Application}{A demolition and enabling works contractor is demolishing a four-storey concrete-frame office building on a 0.6-hectare urban site. Site plant includes a 20-tonne hydraulic excavator with a long-reach demolition arm, two 6-tonne articulated dumpers, a ride-on compaction roller, and a 4-tonne telehandler. Step 1 identifies: pedestrian-excavator conflict during demolition debris loading; dumper blind-spot pedestrian strikes on the haul route; telehandler load obscuring operator visibility at the site entrance. Step 2 identifies: demolition operatives, asbestos enclosure workers, site management, delivery drivers, adjacent office workers whose car park abuts the site boundary. Step 3 rates pedestrian-plant conflict as Very High. Step 4 implements: physical hoarding with locked gate between demolition area and all other site zones; one-way haul circuit for dumpers, eliminating all reversing on the shared haul route except at the tip point, protected by a banksman; telehandler fitted with a camera system covering the forward view over the load and the FOPS-protected cab; delivery management plan restricting HGV arrivals to one at a time with a dedicated lay-by and traffic marshal. Residual risk assessed as Low. Step 5: daily pre-start checks for all plant recorded; STMP reviewed and updated when asbestos enclosure dismantled and site layout changes.}

%---------------------------- SECTION ----------------------------
\section{Applying the hierarchy of controls}

\paragraph{The hierarchy of controls for plant and vehicle hazards in construction places physical separation of pedestrians and plant at the top of the practical control hierarchy, supplemented by engineering controls on individual vehicles, and supported by administrative and competency controls. PPE at the base of the hierarchy --- the high-visibility vest and hard hat of the construction worker --- provides no meaningful protection against a 6-tonne dumper and must never be presented as the primary control for pedestrian-plant conflict.}

\begin{itemize}
  \bitem{Elimination}{Eliminate vehicle movements from the pedestrian zone entirely: design the site layout so that all plant and vehicle operations are confined to zones that are physically inaccessible to persons on foot; specify off-site prefabrication and modular construction approaches that reduce the frequency and duration of plant operations; use remote-controlled demolition equipment to remove operatives from the operating envelope of demolition plant.}
  \bitem{Substitution}{Substitute higher-risk vehicle types and configurations with lower-risk alternatives where the task permits: use a controlled forward-tipping dumper in preference to a high-tip dumper on haul routes where height clearance is not required, improving operator visibility; specify telehandlers with full 360-degree CCTV camera systems as a condition of hire; use pedestrian-operated walk-behind compaction equipment rather than ride-on rollers in areas with constrained pedestrian access management.}
  \bitem{Engineering controls}{Physical separation of pedestrian and vehicle routes using jersey barriers, hoarding panels, or proprietary safety barrier systems; one-way traffic systems to eliminate reversing wherever possible; vehicle reversing cameras, radar proximity sensors, and acoustic-visual reversing alarms on all site plant; overhead protection structures at pedestrian crossings; speed humps, chicanes, and mandatory stop signs at site access points and junction approaches; tyre wash facilities at the site exit to prevent mud deposition on the public highway; proximity detection systems (radar or ultra-wideband) on high-traffic plant.}
  \bitem{Administrative controls}{Site traffic management plan reviewed and updated at each change of site layout or plant operations; CPCS card requirements for all plant operators, with verification before engagement; daily pre-start plant inspections with defect reporting and plant stand-down procedure; banksman deployment plan for all reversing operations near pedestrian areas; delivery management plan with timed delivery slots and a traffic marshal at the site entrance; site induction covering all traffic routes, pedestrian routes, and vehicle exclusion zones for every person entering the site.}
  \bitem{PPE}{High-visibility vest to Class 2 EN ISO 20471 minimum for all persons on site (not as a substitute for segregation, but as a supplementary visual indicator); safety boots with steel midsoles; hard hat; operator ROPS (Roll-Over Protective Structure) and FOPS (Falling Object Protective Structure) on all applicable plant as standard PUWER-compliant equipment; operator seat belt worn at all times in all applicable plant.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Cadence of assessment and review triggers}

\paragraph{The site traffic management plan is a living document that must track the constantly changing geometry of a construction site. As excavations progress, as structures rise, as site compounds are relocated, and as plant fleets change, the traffic management arrangements that were appropriate in week one of a project may be entirely inadequate in week twenty. The assessment cadence must ensure that the STMP is reviewed and updated proactively, not reactively after an incident has demonstrated its inadequacy.}

\begin{itemize}
  \bitem{At each change of site layout}{Whenever the site compound, welfare facilities, pedestrian routes, or haul routes are relocated due to construction progress, a formal review of the STMP must be conducted and the revised arrangements briefed to all site operatives before work recommences on the revised layout. It is not acceptable for site management to assume that operatives will identify and adapt to revised traffic routes without formal briefing.}
  \bitem{When new plant types are introduced}{Every new machine type introduced to the site brings a new visibility profile, a new operating envelope, and potentially new pedestrian conflict points. Before any machine not previously used on the site commences operations, the STMP must be reviewed to confirm that the existing segregation arrangements are adequate for the new machine's blind-spot geometry and operating radius.}
  \bitem{During peak pedestrian movement periods}{The periods of highest pedestrian-plant conflict risk on a construction site are the start of the working day (all operatives arriving simultaneously), break times (operatives crossing haul routes to and from welfare facilities), and end of the working day (operatives and plant both heading for the site entrance at the same time). The STMP must specifically address the management of these periods, with plant standing down or following a restricted route plan during peak pedestrian movement if segregation is not achievable.}
  \bitem{After any pedestrian-plant near-miss or incident}{Any event in which a person on foot comes within the operating envelope of a moving machine without being intentionally in that position, or any event involving a vehicle collision, near-miss, or property damage, must trigger an immediate review of the STMP and the control measures at the relevant location. The review must identify the specific failure that allowed the near-miss or incident to occur and must implement a physical control change, not merely a re-briefing of the existing rules.}
  \bitem{At each weekly site manager review}{The principal contractor's site manager must include a review of the STMP as a standing agenda item at the weekly site management meeting, confirming that the arrangements remain adequate for the current site activities and signing off any changes required for the following week.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Common failure modes}

\begin{itemize}
  \bitem{Shared pedestrian and vehicle routes due to inadequate site layout planning}{The most common single cause of pedestrian-plant conflict on construction sites is a site layout that does not provide adequate physical separation between pedestrian routes and vehicle haul routes. This failure typically originates in the pre-construction phase, when the site layout is established without adequate thought given to the geometry of pedestrian movement to and from welfare facilities, site entrances, and working areas.}
  \bitem{Banksman absent or inadequately positioned for reversing manoeuvres}{The use of a banksman to guide reversing plant is a standard administrative control, but its effectiveness depends on the banksman being in a position where they can see both the reversing vehicle and the full extent of the hazard zone, maintaining constant eye contact with the operator, and using a recognised signalling system. Banksmen positioned behind columns, on the blind side of the vehicle, or using non-standard gestures provide no meaningful risk reduction.}
  \bitem{Plant operators without current CPCS competency certificates}{Site plant operated by workers without the current Construction Plant Competence Scheme (CPCS) card for the relevant machine category is a common enforcement finding. The CPCS card system confirms that the operator has received formal training and passed a theory and practical assessment for the specific machine; the absence of a card means the operator's competence cannot be verified.}
  \bitem{Delivery vehicles reversing from the public highway without traffic management}{Large articulated delivery vehicles reversing into a site entrance from a public road create a pedestrian and road traffic hazard that requires specific traffic management: a traffic marshal on the footpath and road to stop pedestrians and road vehicles during the reversing manoeuvre, and a banksman within the site to guide the vehicle to its unloading position. This combination is frequently absent, with the delivery driver conducting the manoeuvre based on mirrors and instinct.}
  \bitem{Plant inspection defects not acted upon}{Pre-start plant inspection checklists that identify defects --- broken mirrors, non-functioning reversing alarms, damaged proximity sensors --- that are then not acted upon are a common failure in plant management. The inspection record that notes ``reversing alarm non-functional'' and then records the machine as returned to service on the same day without repair provides no risk reduction and creates evidence of management awareness of the defect.}
  \bitem{Speed limits not enforced on haul routes}{Site speed limits of 5--10\,mph are commonly specified in site traffic management plans and are commonly ignored by both plant operators and delivery drivers. Uncontrolled vehicle speeds on haul routes significantly increase braking distances and reaction time windows for pedestrian avoidance. Physical speed reduction measures --- speed humps, chicanes, mandatory stop points --- are more effective than posted limits without enforcement.}
\end{itemize}

\example{Failure Pattern}{A civils contractor is constructing a surface water attenuation tank beneath a new industrial estate access road. The site covers approximately 2 hectares, with a central excavation zone for the tank, a welfare compound at the north end, and a site entrance at the south end. The haul route from the excavation to the spoil storage area at the north end passes within 8 metres of the welfare compound entrance. No physical barrier separates the haul route from the welfare compound access path. During a morning break, six operatives walk from the welfare compound to the excavation area along the edge of the haul route. A 9-tonne articulated dumper travelling south on the haul route encounters two operatives walking abreast of the haul route edge. The dumper operator brakes and manoeuvres to avoid a collision; the machine's left front wheel drops into the unprotected excavation edge, causing the dumper to roll 30 degrees to the left. The operator is retained in the cab by the seatbelt and ROPS, sustaining minor bruising. Investigation finds that the welfare compound was sited in the original STMP layout plan during the pre-construction phase but that the haul route was subsequently repositioned during groundworks without a STMP review, creating the pedestrian-plant conflict corridor that caused the incident.}

%---------------------------- CHAPTER SUMMARY ----------------------------
\newpage
\section{Chapter Summary}

\begin{table}[H]
\centering
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm}X}
\toprule
\textbf{Assessment Element} & \textbf{Operational Requirement} \\
\midrule
Legal/Professional Basis & PUWER 1998; Workplace (Health, Safety and Welfare) Regulations 1992; CDM 2015; Road Traffic Act 1988; Health and Safety at Work etc.\ Act 1974; CLOCS Standard; FORS Standard. \\
Method & Apply the full five-step process and document inherent/residual risk. \\
Controls & Physical segregation of pedestrian and plant routes; site traffic management plan reviewed at each layout change; CPCS-competent operators for all plant; one-way traffic systems; banksman for all reversing operations near pedestrians; delivery management plan. \\
Cadence & At each change of site layout or plant fleet; during peak pedestrian movement periods; after any near-miss or incident; weekly review at site manager meeting. \\
Assurance & Use audit, incident learning, and governance reporting to verify effectiveness. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{itemize}
  \bitem{Key takeaway 1}{Physical separation of pedestrian and plant routes is the primary and most effective control for the pedestrian-plant conflict hazard; high-visibility PPE is a supplementary measure, not a substitute for segregation.}
  \bitem{Key takeaway 2}{The site traffic management plan must be a living document reviewed and updated at every change of site layout, plant fleet, or phase of work; a STMP that does not reflect current site conditions provides no meaningful risk reduction.}
  \bitem{Key takeaway 3}{All plant operators must hold a current CPCS card for the relevant machine category; operator competency must be verified before engagement and should be rechecked at site induction for each project.}
  \bitem{Key takeaway 4}{Near-misses between pedestrians and plant are leading indicators of inadequate segregation; every near-miss must trigger a physical control change, not merely a re-briefing of existing rules.}
\end{itemize}

\barquote{In Chapter~\ref{chap:Manual Handling and Musculoskeletal Risk Assessment}, we address the assessment of manual handling tasks and the prevention of musculoskeletal disorders.}
