\chapter{Excavation and Groundworks Risk Assessment}
\label{chap:Excavation and Groundworks Risk Assessment}

%---------------------------- SECTION ----------------------------
\section{Why this assessment matters in construction}

\paragraph{Excavation and groundworks represent the opening act of almost every construction project, and they carry a disproportionate burden of serious and fatal risk. Trench collapse, unplanned contact with buried services, ground instability, flooding, and the proximity of heavy plant to workers in confined ground-level spaces combine to make groundworks one of the most reliably dangerous phases of any build. The HSE's statistics consistently show that excavation-related fatalities --- operatives buried alive by collapsed trench sides, workers electrocuted by contact with undetected cables, pipe-layers crushed by unsupported ground --- recur with a regularity that demands the excavation risk assessment be treated as a primary document, not an administrative obligation.}

\paragraph{The deceptive familiarity of the activity is itself a hazard. Groundworkers who have opened hundreds of trenches without incident develop a working culture that normalises the absence of trench support, the absence of service scanning, and the absence of supervision. A trench that has stood open and unsupported for several days without any visible sign of movement provides a false assurance that it is safe; the mechanism of trench collapse is frequently sudden and without warning, and the mass of soil that enters an unsupported two-metre trench is sufficient to cause fatal compressive asphyxia within minutes. No operative buried to chest depth by a trench collapse has ever been saved by a colleague with a shovel --- the compressive force of the soil on the thorax prevents breathing before any manual excavation can be completed.}

\paragraph{The buried services environment on UK construction sites is highly complex. The land beneath a typical urban construction site may contain water mains, gas pipes at medium and high pressure, electricity cables at low and high voltage, telecommunications cables, district heating pipes, and historical drainage infrastructure that may not appear on any utility record. HSG47 (Avoiding Danger from Underground Services) provides the definitive framework for safe digging in the vicinity of buried services, and the New Roads and Street Works Act 1991 imposes additional duties in relation to street works and the protection of utilities. The failure to obtain current service records, conduct cable avoidance tool (CAT) surveys before excavation, and implement safe digging zones and hand-dig procedures around identified or estimated service locations remains one of the most common and most catastrophic failures in groundworks risk management.}

\paragraph{Water, gas, and electricity each carry distinct consequences when struck. An electricity cable strike at 11\,kV can deliver a fatal electric shock without physical contact, through arc flash. A gas main strike releases flammable or explosive gas into a confined excavation and adjacent buildings. A water main strike can flood a trench and drown an operative before pumping can be established. Each of these scenarios has been the subject of HSE prosecution and coroners' inquests; each is preventable by the systematic application of the service identification, CAT survey, safe digging zone, and hand-dig procedures described in HSG47. The excavation risk assessment must address all three service types specifically, and must describe the hierarchy of controls from desk study through CAT survey to hand digging, not merely require that services be ``identified before excavation commences.''}

\barquote{``An unsupported trench is not a temporary risk --- it is a loaded weapon that will discharge at the moment of least expectation. The excavation risk assessment exists to ensure the weapon is never loaded in the first place.''}

\example{Buried Service Strike}{A groundworks contractor is excavating a service trench across a commercial car park for the installation of new drainage. The existing utility records provided by the statutory undertakers show a 33\,kV electricity cable running approximately 1.2 metres to the west of the proposed trench centreline. A CAT survey is conducted the morning of excavation, and a linear signal is detected approximately 1.4 metres west of the centreline. The operative conducting the CAT survey marks the estimated service line and moves on to excavation using a tracked mini excavator. At a depth of 900\,mm, the excavator bucket contacts a second, unrecorded 11\,kV cable running obliquely across the excavation line. The cable sheath is breached; the arc flash burns the operative's forearm and damages the excavator boom. Investigation reveals that the unmarked cable was a former supply to a demolished building whose records had not been updated in the utility archive. The failure to conduct hand-digging within 500\,mm of the estimated service line --- as required by HSG47 --- is identified as the primary control failure.}

%---------------------------- SECTION ----------------------------
\section{Legal and professional framework}

\paragraph{The legal framework for excavation and groundworks is among the most multi-layered in construction health and safety. It draws on primary health and safety legislation, specific regulations for work equipment and electricity, the statutory utilities regime under the New Roads and Street Works Act 1991, and a body of HSE technical guidance that translates the broad regulatory duties into specific operational requirements. The Construction (Design and Management) Regulations 2015 place specific duties on principal designers and principal contractors in relation to the management of ground risks, the provision of pre-construction information (including utility records and ground investigation data), and the inclusion of excavation safety in the Construction Phase Plan.}

\paragraph{Key frameworks relevant to this chapter include: Health and Safety at Work etc.\ Act 1974; Management of Health and Safety at Work Regulations 1999; Construction (Design and Management) Regulations 2015; PUWER 1998; New Roads and Street Works Act 1991; HSG47; HSG185; Electricity at Work Regulations 1989; Gas Safety (Installation and Use) Regulations 1998; RIDDOR 2013.}

\begin{itemize}
  \bitem{Health and Safety at Work etc.\ Act 1974}{Places the overarching duty on employers to ensure, so far as is reasonably practicable, the health, safety and welfare of their employees, and the duty not to expose non-employees to risks to their health and safety. In the excavation context, this duty requires the implementation of suitable and sufficient risk controls for every phase of groundworks activity from setting out to final reinstatement.}
  \bitem{Management of Health and Safety at Work Regulations 1999}{Requires a suitable and sufficient risk assessment of all groundworks activities before work commences, and the appointment of competent persons to manage those risks. Regulation 3 specifically requires the assessment to identify all hazards and evaluate the risks they present to persons who may be harmed, including third parties such as members of the public adjacent to open excavations.}
  \bitem{Construction (Design and Management) Regulations 2015}{Requires that pre-construction information is gathered and communicated, including utility records, ground investigation data, and information about previous land use that may indicate contaminated ground or abandoned services. The Construction Phase Plan must address excavation safety arrangements, including trench support, service avoidance, and emergency procedures. The principal contractor must satisfy itself that groundworks contractors are competent to manage the excavation risks on the project.}
  \bitem{PUWER 1998}{Applies to all work equipment used in groundworks activities, including excavators, compactors, trench shoring systems, and pumping equipment. PUWER requires that equipment is suitable for its intended purpose, maintained in an efficient and safe condition, inspected at appropriate intervals, and operated only by trained and competent persons. The suitability of a trench box or hydraulic shoring system for the specific ground conditions and trench dimensions encountered must be confirmed by a competent person before the system is installed.}
  \bitem{New Roads and Street Works Act 1991 and Roads (Scotland) Act 1984}{Govern the legal framework for works in the public highway, including the requirement to obtain a street works licence, to notify the highway authority and relevant statutory undertakers, to attend a pre-commencement meeting where required, and to comply with the New Roads and Street Works Act 1991 Codes of Practice (Specification for the Reinstatement of Openings in Highways). The NRSWA framework is relevant to any groundworks activity in or adjacent to the public highway.}
  \bitem{HSG47 --- Avoiding Danger from Underground Services}{The primary HSE guidance document for the identification and avoidance of buried services. HSG47 prescribes the sequence of: obtaining current utility records from all relevant statutory undertakers; conducting a cable avoidance tool (CAT) and signal generator (Genny) survey of the excavation area; establishing safe digging zones of 500\,mm each side of the estimated service position within which only hand digging is permitted; and implementing safe excavation procedures including the use of vacuum excavation where buried service positions are confirmed.}
  \bitem{HSG185 --- Health and Safety in Excavations}{The primary HSE guidance document for excavation stability, trench support, and excavation management. HSG185 addresses soil classifications and the behaviour of different soil types under unsupported conditions; the design and installation of trench support systems including open-cut battering, close sheeting, and hydraulic trench boxes; inspection of excavations before each period of work and after any event that may have affected stability; and the management of ground water, adjacent loading from plant and spoil heaps, and vibration from nearby works.}
  \bitem{Electricity at Work Regulations 1989}{Requires that all electrical systems, including buried cables, are maintained in a safe condition and that work is not carried out on or near live conductors unless it is unreasonable in all the circumstances for the conductor to be made dead. In the groundworks context, this means that excavation in the vicinity of buried electrical cables must follow the service avoidance procedures in HSG47, that work near overhead lines requires specific controls including goal-post barriers and exclusion zones, and that any damaged cable must be treated as live and the relevant distribution network operator (DNO) notified immediately.}
  \bitem{Gas Safety (Installation and Use) Regulations 1998}{Require that gas installations, including buried mains and service pipes, are not put at risk by nearby works. Contact with a buried gas pipe during excavation must be treated as an emergency; the site must be evacuated, the gas emergency number called, and the area cordoned off. The risk assessment must include the specific emergency response procedure for a gas main strike.}
  \bitem{RIDDOR 2013}{Requires the reporting of deaths and specified injuries arising from excavation-related incidents, including fatal trench collapses, electric shock from buried cables, and other excavation-related fatalities, to the relevant enforcing authority within the prescribed notification periods.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Conducting the assessment using the five-step method}

\subsection{Step 1 --- Identify Hazards}
\paragraph{Excavation hazards fall into four broad categories: ground movement and collapse; buried and overhead services; flooding and groundwater ingress; and plant and vehicle proximity. Ground movement and collapse hazards include: sudden trench collapse due to unsupported or inadequately supported sides; progressive failure of trench sides in cohesionless soils such as sand and gravel; destabilisation of adjacent structures by excavation-induced settlement or vibration; and slope failure at the face of a bulk excavation. The risk of trench side collapse is affected by soil type (cohesive clays are more stable than granular soils but can fail suddenly when saturated), depth of excavation, proximity of surcharge loads such as plant, spoil heaps, or stored materials, groundwater conditions, and the presence of any pre-existing disturbance to the ground such as backfilled service trenches or made ground. Buried services hazards include: electric cables at all voltages from 240V domestic supply to 33\,kV high-voltage distribution; gas mains and service pipes at low, medium, and high pressure; water mains under pressure; telecommunications cables and ducting; and drainage infrastructure including combined and separate sewers. Flooding hazards include: tidal and seasonal groundwater ingress; burst water main flooding of an open trench; surface water runoff from adjacent ground and highways; and dewatering pump failure during a wet period. Plant proximity hazards include: excavator bucket contact with an operative in the trench; plant overrun of the trench edge causing collapse; dumper and tipper tipping into an open trench; and overhead power lines contacted by excavator or delivery crane boom.}

\subsection{Step 2 --- Decide Who or What Is Affected}
\paragraph{Persons at risk from excavation hazards include the groundworkers installing drainage, foundations, or services within the trench; the plant operators of excavators, dumpers, and compaction equipment working at the trench edge; banksmen and ground-level operatives working in the zone of influence of the excavator; operatives of other trades who may approach or cross an open excavation; and members of the public who may gain access to the site of excavation through inadequate hoarding or at night when the site is unoccupied. Adjacent property occupiers may be at risk from ground movement or vibration affecting their buildings, particularly in urban areas where excavations are close to existing foundations. Utility workers from statutory undertakers may be attending the site to identify or protect their assets. Emergency services personnel attending a trench collapse are themselves at serious risk, and the risk assessment must address the emergency response procedure that minimises the risk of secondary collapse during rescue operations.}

\subsection{Step 3 --- Evaluate Likelihood and Consequence}
\paragraph{The consequence of a trench collapse involving an operative is potentially fatal; the mass of soil that enters a two-metre-deep trench in cohesionless ground can exceed three tonnes per linear metre of trench. The inherent risk of unsupported excavation in any ground type must be rated Very High (typically the maximum on any risk matrix) before controls are applied. The likelihood of collapse in unsupported granular soils at depth is high; the Working Rule that no excavation exceeding 1.2 metres in depth should be worked in without an assessment of the need for support reflects the reality that this threshold marks the transition between depths from which a collapse might be survivable and depths from which it is almost certainly fatal. Electric cable strikes carry a high probability of fatal electrocution or severe burn injury; gas main strikes carry a high probability of explosion or asphyxiation in an enclosed trench. The evaluation must not allow the frequency of excavation activity without incident to reduce the likelihood rating below the technically correct assessment: the absence of a previously recorded collapse does not mean the ground will not collapse --- it means it has not collapsed yet.}

\subsection{Step 4 --- Select and Implement Controls}
\paragraph{Controls for excavation must address each category of hazard identified in Step 1. For ground stability: the method of excavation (open-cut battering, close sheeting with timber or steel, hydraulic trench box, or sheet piling) must be selected by a competent person based on soil conditions, depth, adjacent loading, and groundwater; the selected method must be capable of supporting the specific ground conditions encountered, not a generic ``trench box provided'' instruction; no operative should enter any excavation deeper than 1.2 metres until competent assessment of the support requirement has been completed and appropriate support installed; the trench box or shoring must be installed from outside the excavation or from a protected position within it, never by an operative standing in an unsupported trench. For buried services: the sequence of desk study (obtaining all available utility records), CAT and Genny survey, marking of estimated service positions, establishment of safe digging zones, hand-digging within 500\,mm of estimated service lines, and vacuum excavation for confirmed service positions must be specified step by step. For plant proximity: exclusion zones between excavator working radius and any person on the ground must be established and enforced; stop blocks must be positioned to prevent vehicle overrun of the trench edge; a banksman must be present whenever plant operates near an open excavation or within proximity of any person. For flooding: dewatering capacity must be pre-installed before excavation reaches groundwater level; a rescue procedure for a rapidly flooding trench must be established; the emergency stop procedure for a burst water main must identify the nearest stopcock and the responsible operative.}

\subsection{Step 5 --- Record and Review}
\paragraph{The excavation risk assessment must be reviewed at the start of each working day by the site supervisor or groundworks foreman against actual ground conditions, groundwater levels, and any changes to the excavation geometry or adjacent loading. Any change to the trench configuration, soil conditions encountered, or adjacent activity that may affect trench stability must trigger an immediate review. HSG185 and the Construction (Design and Management) Regulations 2015 require that excavations are inspected by a competent person before each period of work, after any event likely to have affected their stability, and at least once every working day. The inspection must be recorded in writing using a form that captures the date, time, location, condition observed, any defects identified, and the action taken. The competent person must have sufficient knowledge of ground behaviour and trench support to identify signs of incipient failure including tension cracks at the ground surface, seepage of water at the trench sides, or progressive bulging of the trench face.}

\example{Five-Step Application}{A drainage contractor is excavating a 2.4-metre-deep trench in made ground to install a new foul sewer on a brownfield housing development. Step 1 identifies: trench side collapse in made ground of variable composition; buried gas service pipe shown on utility records approximately 800\,mm south of trench centreline; potential groundwater ingress at 1.8 metres based on borehole data in the pre-construction information. Step 2 identifies: pipe-layer working in the trench; excavator operator; banksman; adjacent plot groundworkers within 15 metres. Step 3 rates inherent risk of collapse at Very High, buried gas strike at Very High. Step 4 implements: hydraulic trench box specified by a competent engineer for the made-ground conditions; CAT and Genny survey confirms gas pipe at 750\,mm from centreline; safe digging zone established; hand-dig approach within 600\,mm of gas pipe; dewatering pump installed before excavation reaches 1.6 metres; exclusion zone of 3 metres established around excavator working radius. Residual risk assessed as Medium following support installation and Low following service clearance. Step 5 specifies: daily competent person inspection before operatives enter trench; immediate review if groundwater breakthrough or cracking observed.}

%---------------------------- SECTION ----------------------------
\section{Applying the hierarchy of controls}

\paragraph{The hierarchy of controls in excavation and groundworks must be applied to each hazard category simultaneously, since a single excavation activity will typically present multiple independent risk pathways. The temptation to apply generic controls --- ``trench box provided, CAT survey conducted, banksman deployed'' --- without verifying that each control is adequate for the specific conditions encountered on the day is a recurring pattern in inadequate groundworks risk assessments. The hierarchy must be applied with specificity, and each control must be verified as installed and functional before any operative enters the excavation.}

\begin{itemize}
  \bitem{Elimination}{Eliminate the need for deep open excavations where possible: specify directional drilling (pipe jacking, micro-tunnelling) for buried service installation beneath roads and structures; specify pre-cast drainage systems that can be installed without deep open-cut excavation; design foundation systems that avoid deep basements in groundwater-affected sites; use ground investigation data to route excavations away from known service concentrations.}
  \bitem{Substitution}{Where open-cut excavation is unavoidable, substitute higher-risk trench configurations with lower-risk alternatives: use battering or benching of trench sides where site space permits rather than relying on trench box installation in unstable ground; use vacuum excavation (hydrovac or suction excavation) rather than mechanical excavation near confirmed buried service positions; specify pre-installation utility diversions for the highest-voltage cables and highest-pressure gas mains before groundworks commence, removing the proximity hazard entirely.}
  \bitem{Engineering controls}{Install proprietary hydraulic trench boxes or aluminium drag boxes of appropriate size and rated capacity for the trench depth and soil conditions; install close-sheeted timber support with hydraulic or acrow prop walings for irregular excavation geometries; install sheet piles for deep excavations or excavations adjacent to existing structures; specify and install adequate dewatering systems before groundwater level is reached; install stop blocks and edge restraints at trench lips to prevent plant overrun; install hoarding and fencing around all open excavations, with locked access gates when the site is unoccupied.}
  \bitem{Administrative controls}{Mandate the HSG47 five-stage service avoidance sequence for all excavation activity; require daily written inspection records for all excavations deeper than 1.2 metres; implement an excavation permit-to-work that must be signed by the site manager before any operative descends into an excavation; require that all operatives working in excavations hold a current CSCS card (Groundworker) and have received a site-specific induction that includes the emergency response procedure for trench collapse and service strikes; prohibit the use of mechanical excavation within 500\,mm of any identified service.}
  \bitem{PPE}{Hard hat, high-visibility vest, and safety boots with steel midsoles for all operatives in the excavation zone; rubber-soled safety footwear when working near identified electrical services; arc flash-rated PPE for any operations confirmed to be within proximity of high-voltage cables; gas detection equipment (combined CO, H\textsubscript{2}S, O\textsubscript{2}, and flammable gas monitor) for any excavation into made ground, near gas infrastructure, or in any confined or semi-confined situation; ear protection for operatives using compaction equipment or pneumatic tools in the trench.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Cadence of assessment and review triggers}

\paragraph{The excavation risk assessment is not a document produced once and filed. Ground conditions change hour by hour: rainfall saturates the soil and increases hydrostatic pressure on trench sides; overnight frost followed by morning thaw weakens previously stable faces; water main seepage changes the bearing capacity of the trench floor. The assessment review cadence must reflect this dynamic ground environment and must be embedded in the daily supervision routine of the groundworks foreman and site manager.}

\begin{itemize}
  \bitem{Pre-entry daily inspection}{Before any operative enters an excavation on any working day, a competent person must conduct a visual inspection of the trench sides and any installed support, confirm that no water ingress has occurred overnight, check for surface cracking at the trench perimeter, and confirm that no adjacent plant or materials have been placed within the surcharge exclusion zone. The inspection must be recorded in writing even if no defects are found.}
  \bitem{After rainfall or adverse weather}{Any significant rainfall event, overnight frost, or thaw must trigger an additional inspection of all open excavations before work resumes. Saturated ground behaves very differently from dry ground; previously stable trench sides may have lost their apparent cohesion and become susceptible to flow failure. The competent inspector must assess whether the installed support remains adequate for the changed ground conditions.}
  \bitem{When ground conditions change}{If the excavation encounters ground conditions different from those assumed in the original assessment --- unexpected made ground, buried obstructions, water-bearing gravel horizons, or soft alluvial deposits --- the excavation must be halted, the trench made safe from outside, and the assessment reviewed by a competent engineer before work resumes. No assumption should be made that installed support designed for one soil type will be adequate for another.}
  \bitem{After any adjacent plant or vehicle loading change}{Any change to the position or loading of plant, vehicles, or spoil heaps within the zone of influence of the excavation (typically 1.5 times the trench depth from the trench edge) must trigger a reassessment of the surcharge loading on the trench support system. A tipper truck discharging spoil at the trench edge can exert a surcharge pressure that causes trench box displacement or trench wall failure.}
  \bitem{After any service strike or near-miss}{Any contact with a buried service, near-miss involving plant and personnel in the excavation zone, or any incident that may have affected the integrity of the trench support or the safety of the excavation must immediately halt work, trigger a review of the full risk assessment, and require documented sign-off by a competent engineer before the excavation is re-entered. RIDDOR notification requirements must be considered immediately.}
\end{itemize}

%---------------------------- SECTION ----------------------------
\section{Common failure modes}

\begin{itemize}
  \bitem{Excavating without trench support below the safe depth threshold}{The most common single cause of fatal trench collapse incidents is the decision by experienced groundworkers to work in unsupported excavations at depths where support is required. The reasoning is typically that the trench will only be open for a few hours, that the ground feels stable, and that fitting the trench box will slow the work. HSG185 and the case law arising from fatal trench incidents are unambiguous: no excavation exceeding 1.2 metres in depth should be entered until the need for support has been assessed and, if required, support installed. Experienced groundworkers dying in unsupported trenches is not an indicator that support was unnecessary --- it is an indicator that a fatal event had not yet occurred.}
  \bitem{Failure to obtain current utility records before excavation}{Utility records are frequently outdated, missing services installed after the last survey, and inaccurate in their positional data. The failure to obtain records from all relevant statutory undertakers --- not just those the contractor assumes are present --- before any excavation commences is a primary control failure. On brownfield sites, former industrial services, private distribution networks, and decommissioned but not removed cables may be present without appearing in any current utility archive.}
  \bitem{CAT survey conducted but results ignored or misinterpreted}{The cable avoidance tool requires trained and competent operation to produce reliable results. A CAT survey conducted by an untrained operative, conducted without a signal generator, conducted without sweeping in both parallel and perpendicular directions across the excavation area, or conducted without marking and recording the results, provides no meaningful risk reduction. Even a correctly conducted CAT survey will not detect all services: deeply buried cables, cables in metal conduit that screens the signal, and plastic-encased services with no tracer wire may not be detected.}
  \bitem{Spoil heaps and plant positioned at the trench edge}{Placing excavated spoil immediately adjacent to the trench edge is a persistent failure that significantly increases the surcharge loading on the trench sides and reduces the ability to inspect the trench edge for surface cracking. Spoil heaps must be located at a safe distance from the trench edge, and this distance must be specified in the risk assessment based on the trench depth and soil type.}
  \bitem{No emergency response procedure for trench collapse}{Risk assessments that identify trench collapse as a hazard but provide no specific emergency response procedure --- identifying the emergency contact, the equipment to be used for rescue, and the instruction that no person should enter an unstable collapsed trench without specific shoring support --- leave the site completely unprepared for an event that can kill a rescuer as easily as the original victim.}
  \bitem{Excavation left open and unguarded at the end of the working day}{Open excavations on construction sites that are not adequately fenced, lighted, and secured at the end of the working day are a recurring source of falls by site operatives returning in the dark and, tragically, of falls by unauthorised persons including children who access the site overnight. The risk assessment must address the end-of-day securing procedure for every open excavation on the site.}
\end{itemize}

\example{Failure Pattern}{A groundworks gang is installing a storm drain across a commercial development site in sandy alluvial ground. The ground is dry and the trench sides appear stable at a depth of 1.8 metres. The gang foreman decides not to use the aluminium trench box that is available on site, on the grounds that it will be difficult to strip around the pre-cast manhole chambers being installed. Three operatives are in the trench backfilling haunching around a manhole when a 4-metre length of trench side fails, discharging approximately 6 tonnes of sand into the trench. Two operatives are buried from the shoulders up and one to the waist. Emergency services attend and conduct a rescue over 90 minutes, during which a secondary slip occurs, briefly burying one paramedic to the knees. One operative sustains fatal compression asphyxia; the other two survive with significant soft-tissue injuries. The investigation identifies that: no competent person had assessed the need for trench support; the risk assessment stated ``trench box to be used'' but was not enforced; and the foreman had made an informal operational decision not to use the box based on the perceived difficulty of installation. The management failure to verify compliance with the risk assessment controls is identified as a principal causal factor alongside the direct omission of trench support.}

%---------------------------- CHAPTER SUMMARY ----------------------------
\newpage
\section{Chapter Summary}

\begin{table}[H]
\centering
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm}X}
\toprule
\textbf{Assessment Element} & \textbf{Operational Requirement} \\
\midrule
Legal/Professional Basis & Health and Safety at Work etc.\ Act 1974; CDM 2015; PUWER 1998; HSG47; HSG185; Electricity at Work Regulations 1989; Gas Safety (Installation and Use) Regulations 1998; RIDDOR 2013. \\
Method & Apply the full five-step process and document inherent/residual risk. \\
Controls & Trench support designed for actual ground conditions; HSG47 five-stage service avoidance sequence; daily written inspection by competent person; exclusion zones for plant proximity; emergency response procedures for collapse and service strikes. \\
Cadence & Daily pre-entry inspection; immediate review after rainfall, ground condition change, adjacent loading change, or any incident; review at each change of excavation geometry or ground conditions. \\
Assurance & Use audit, incident learning, and governance reporting to verify effectiveness. \\
\bottomrule
\end{tabularx}
\end{table}

\begin{itemize}
  \bitem{Key takeaway 1}{Trench collapse is sudden, unpredictable, and almost invariably fatal at depths exceeding 1.2 metres --- the risk assessment must specify trench support appropriate for the actual ground conditions encountered, and support must be installed before any operative enters the excavation.}
  \bitem{Key takeaway 2}{The HSG47 five-stage service avoidance sequence --- desk study, CAT and Genny survey, safe digging zones, hand-dig, and vacuum excavation --- must be applied in full for every excavation; failure at any one stage invalidates the protection provided by the stages before it.}
  \bitem{Key takeaway 3}{Excavations must be inspected by a competent person before each period of work, after adverse weather, and after any event that may have affected stability; the inspection must be recorded in writing and the record retained on site.}
  \bitem{Key takeaway 4}{An emergency response procedure for trench collapse must be specific, pre-planned, and communicated to all site personnel; no second entry into a collapsed trench should be made without shoring support for the rescuers.}
  \bitem{Key takeaway 5}{Open excavations must be secured, fenced, and illuminated at the end of every working day; the securing procedure must form part of the risk assessment and must be verified by the site manager before the site is vacated.}
\end{itemize}

\barquote{In Chapter~\ref{chap:Lifting Operations and Lifting Equipment Risk Assessment}, we examine the planning and control of lifting operations using cranes, hoists, and MEWPs on construction sites.}
