\chapter*{Appendix C: Primary Regulators and Supervisory Bodies}
\label{chap:Appendix C: Primary Regulators and Supervisory Bodies}

\paragraph{This appendix provides a reference guide to the primary regulatory and supervisory bodies relevant to the risk assessment obligations discussed in this book. Organised by sector and jurisdiction.}

%---------------------------- SECTION ----------------------------
\section*{C.1 United Kingdom — Financial Services}
\paragraph{\textbf{Financial Conduct Authority (FCA)} - Website: \textit{www.fca.org.uk} The conduct regulator for financial services firms in the United Kingdom and the prudential regulator for firms not supervised by the PRA. The FCA's statutory objectives include protecting consumers, enhancing market integrity, and promoting competition. Responsible for the SM\&CR, Consumer Duty, AML supervision of many financial services firms, and market abuse enforcement}
\paragraph{\textbf{Prudential Regulation Authority (PRA)} - Website: \textit{www.bankofengland.co.uk/prudential-regulation} The prudential regulator for banks, building societies, credit unions, insurers, and major investment firms in the United Kingdom. A subsidiary of the Bank of England. Responsible for the prudential framework including capital adequacy requirements, stress testing, and operational resilience for the most systemically significant financial firms.}
\paragraph{\textbf{Financial Policy Committee (FPC)} - Website: \textit{www.bankofengland.co.uk/financial-policy} The Bank of England committee responsible for identifying, monitoring, and taking action to remove or reduce systemic risks — with the objective of protecting and enhancing the resilience of the UK financial system.}
\paragraph{\textbf{Payment Systems Regulator (PSR)} - Website: \textit{www.psr.org.uk} The economic regulator for payment systems in the United Kingdom — promoting competition and innovation in payment systems and ensuring that payment systems work in the interests of all users.}
\paragraph{\textbf{The Pensions Regulator (TPR)} - Website: \textit{www.thepensionsregulator.gov.uk} The regulator of work-based pension schemes in the United Kingdom — protecting members' benefits and promoting good administration of pension schemes.}

%---------------------------- SECTION ----------------------------
\section*{C.2 United Kingdom — Health, Safety, and Environment}
\paragraph{\textbf{Health and Safety Executive (HSE)} - Website: \textit{www.hse.gov.uk} The national regulator for workplace health and safety in Great Britain — responsible for preventing work-related death, injury, and ill health. Also responsible for regulating major hazard industries including nuclear (through the Office for Nuclear Regulation), offshore oil and gas, and COMAH sites.}
\paragraph{\textbf{Office for Nuclear Regulation (ONR)} - Website: \textit{www.onr.org.uk} The independent nuclear safety and security regulator for the United Kingdom — responsible for licensing nuclear sites, regulating nuclear security, and enforcing transport regulations for radioactive materials.}
\paragraph{\textbf{Environment Agency (EA)} - Website: \textit{www.gov.uk/environment-agency} The principal environmental regulator in England — responsible for regulating major industry and waste, treatment of contaminated land, water quality and resources, and fisheries. Issues environmental permits and enforces environmental legislation.}
\paragraph{\textbf{Natural England} - Website: \textit{www.gov.uk/government/organisations/natural-england} The government's adviser for the natural environment in England — responsible for protecting and improving the natural environment, including the regulation of activities affecting Sites of Special Scientific Interest and protected species.}
\paragraph{\textbf{Scottish Environment Protection Agency (SEPA)} - Website: \textit{www.sepa.org.uk} The environmental regulator for Scotland — responsible for protecting and improving the environment through regulation of industry, waste management, and water quality.}

%---------------------------- SECTION ----------------------------
\section*{C.3 United Kingdom — Healthcare}
\paragraph{\textbf{Care Quality Commission (CQC)} - Website: \textit{www.cqc.org.uk} The independent regulator of health and adult social care in England — responsible for registering, monitoring, and inspecting services to ensure they meet fundamental standards of quality and safety.}
\paragraph{\textbf{Medicines and Healthcare products Regulatory Agency (MHRA)} - Website: \textit{www.gov.uk/government/organisations/medicines-and-healthcare-products-regulatory-agency} The regulatory body responsible for ensuring that medicines, medical devices, and blood components for transfusion meet applicable standards of safety, quality, and efficacy.}
\paragraph{\textbf{General Medical Council (GMC)} - Website: \textit{www.gmc-uk.org} The regulator of doctors in the United Kingdom — responsible for maintaining the register of qualified doctors, setting standards for medical practice and education, and investigating concerns about doctors' fitness to practise.}
\paragraph{\textbf{Nursing and Midwifery Council (NMC)} - Website: \textit{www.nmc.org.uk} The regulator for nurses, midwives, and nursing associates in the United Kingdom — maintaining a register of those qualified to practise and setting standards for professional practice and education.}
\paragraph{\textbf{Health and Care Professions Council (HCPC)} - Website: \textit{www.hcpc-uk.org} The regulator for a range of health and care professions — including physiotherapists, paramedics, psychologists, and social workers — in the United Kingdom.}

%---------------------------- SECTION ----------------------------
\section*{C.4 United Kingdom — Legal Services}
\paragraph{\textbf{Solicitors Regulation Authority (SRA)} - Website: \textit{www.sra.org.uk} The independent regulatory body for solicitors and law firms in England and Wales — responsible for setting standards, authorising firms, and taking action against those who fall short of the required standards.}
\paragraph{\textbf{Bar Standards Board (BSB)} - Website: \textit{www.barstandardsboard.org.uk} The regulatory body for barristers in England and Wales — responsible for setting the education and training requirements, professional standards, and conduct requirements for the Bar.}
\paragraph{\textbf{Council for Licensed Conveyancers (CLC)} - Website: \textit{www.clc-uk.org} The specialist regulator for the property law sector — licensing and regulating licensed conveyancers and licensed conveyancing practitioners.}
\paragraph{\textbf{Legal Services Board (LSB)} - Website: \textit{www.legalservicesboard.org.uk} The oversight regulator for legal services in England and Wales — responsible for overseeing the approved regulators, promoting the regulatory objectives of the Legal Services Act 2007, and approving changes to regulatory arrangements.}

%---------------------------- SECTION ----------------------------
\section*{C.5 United Kingdom — Other Sectors}
\paragraph{\textbf{Information Commissioner's Office (ICO)} - Website: \textit{www.ico.org.uk} The UK's independent authority for upholding information rights — responsible for enforcing the UK GDPR, the Data Protection Act 2018, the Privacy and Electronic Communications Regulations, and other information rights legislation.}
\paragraph{\textbf{Civil Aviation Authority (CAA)} - Website: \textit{www.caa.co.uk} The UK's aviation regulator — responsible for the regulation of civil aviation including safety, airspace, consumer protection, and the economic regulation of airports.}
\paragraph{\textbf{Ofgem (Office of Gas and Electricity Markets)} - Website: \textit{www.ofgem.gov.uk} The regulator for gas and electricity markets in Great Britain — protecting consumers, promoting competition, and regulating network companies.}
\paragraph{\textbf{Ofwat (Water Services Regulation Authority)} - Website: \textit{www.ofwat.gov.uk} The economic regulator for the water sector in England and Wales — regulating water and wastewater companies and protecting the interests of consumers.}
\paragraph{\textbf{Competition and Markets Authority (CMA)} - Website: \textit{www.gov.uk/government/organisations/competition-and-markets-authority} The UK's primary competition and consumer protection authority — responsible for investigating mergers, markets, and anti-competitive behaviour.}
\paragraph{\textbf{Financial Reporting Council (FRC)} - Website: \textit{www.frc.org.uk} The regulator responsible for corporate governance and reporting in the United Kingdom — setting the UK Corporate Governance Code, the UK Stewardship Code, and accounting and auditing standards.}
\paragraph{\textbf{National Cyber Security Centre (NCSC)} - Website: \textit{www.ncsc.gov.uk} Part of GCHQ, the NCSC provides cybersecurity guidance, support, and incident response for UK organisations — producing authoritative guidance on cybersecurity risk management and threat intelligence.}

%---------------------------- SECTION ----------------------------
\section*{C.6 European Union}
\paragraph{\textbf{European Banking Authority (EBA)} - Website: \textit{www.eba.europa.eu} The EU authority responsible for the regulation and supervision of the European banking sector — developing technical standards, guidelines, and recommendations for the implementation of the Basel framework and other banking regulation across the EU.}
\paragraph{\textbf{European Insurance and Occupational Pensions Authority (EIOPA)} - Website: \textit{www.eiopa.europa.eu} The EU supervisory authority for insurance and occupational pensions — responsible for developing technical standards and guidelines for the implementation of Solvency II and other insurance regulation.}
\paragraph{\textbf{European Securities and Markets Authority (ESMA)} - Website: \textit{www.esma.europa.eu} The EU authority responsible for the regulation of securities markets — developing technical standards for MiFID II, EMIR, the Market Abuse Regulation, and other securities legislation.}
\paragraph{\textbf{European Data Protection Board (EDPB)} - Website: \textit{www.edpb.europa.eu} The independent European body responsible for ensuring the consistent application of the GDPR across the EU — issuing guidelines, recommendations, and best practices on the interpretation and application of the regulation.}
\paragraph{\textbf{European Central Bank (ECB) — Banking Supervision} - Website: \textit{www.bankingsupervision.europa.eu} The ECB's supervisory arm — directly supervising significant credit institutions in the euro area through the Single Supervisory Mechanism (SSM).}

%---------------------------- SECTION ----------------------------
\section*{C.7 International Bodies}
\paragraph{\textbf{Basel Committee on Banking Supervision (BCBS)} - Website: \textit{www.bis.org/bcbs} The primary global standard-setter for the prudential regulation of banks — operating under the auspices of the Bank for International Settlements. Responsible for the Basel Framework.}
\paragraph{\textbf{Financial Stability Board (FSB)} - Website: \textit{www.fsb.org} An international body that monitors and makes recommendations about the global financial system — coordinating national financial authorities and international standard-setting bodies to develop and promote the implementation of effective regulatory, supervisory, and other financial sector policies.}
\paragraph{\textbf{Financial Action Task Force (FATF)} - Website: \textit{www.fatf-gafi.org} The global money laundering and terrorist financing watchdog — setting international standards for anti-money laundering and counter-terrorist financing and assessing countries' compliance with those standards.}
\paragraph{\textbf{International Organization for Standardization (ISO)} - Website: \textit{www.iso.org} The international standard-setting body — responsible for ISO 31000 (risk management), ISO 27001 (information security), ISO 45001 (occupational health and safety), and many other standards relevant to risk management practice.}
\paragraph{\textbf{International Civil Aviation Organization (ICAO)} - Website: \textit{www.icao.int} The UN specialised agency for civil aviation — setting international standards and recommended practices for aviation safety, security, and environmental protection.}
\paragraph{\textbf{International Labour Organization (ILO)} - Website: \textit{www.ilo.org} The UN agency responsible for setting international labour standards — relevant to human rights and supply chain due diligence risk assessment.}
