\chapter*{Appendix M: Professional Bodies and Further Resources}
\label{chap:Appendix M: Professional Bodies and Further Resources}

\paragraph{A reference guide to the principal professional bodies, qualifications, and further learning resources relevant to risk assessment for compliance and legal professionals.}

\section*{M.1 Professional Bodies}
\paragraph{\textbf{Institute of Risk Management (IRM) Website}: \textit{www.theirm.org} The leading professional body for risk management globally — offering qualifications including the International Certificate in Enterprise Risk Management and the International Diploma in Risk Management. Publishes guidance, research, and professional standards on risk management practice.}
\paragraph{\textbf{Chartered Institute of Internal Auditors (CIIA)} Website: \textit{www.iia.org.uk} The professional body for internal auditors in the United Kingdom and Ireland — offering qualifications including the Chartered Internal Auditor designation and publishing the UK supplement to the IIA's International Standards for Internal Auditing.}
\paragraph{\textbf{International Compliance Association (ICA)} Website: \textit{www.int-comp.org} A global membership and examining body for compliance and financial crime professionals — offering qualifications at certificate, diploma, and advanced diploma level across compliance, AML, and financial crime prevention.}
\paragraph{\textbf{Chartered Institute of Compliance (CiC)} Website: \textit{www.compliance-institute.org} A professional body for compliance professionals — offering qualifications and professional development resources across a range of compliance disciplines.}
\paragraph{\textbf{Association of Certified Fraud Examiners (ACFE)} Website: \textit{www.acfe.com} The world's largest anti-fraud organisation — offering the Certified Fraud Examiner (CFE) qualification and publishing the Report to the Nations on occupational fraud and abuse.}
\paragraph{\textbf{Chartered Insurance Institute (CII)} Website: \textit{www.cii.co.uk} The professional body for insurance and financial planning professionals — offering qualifications across risk, insurance, and financial advice.}
\paragraph{\textbf{Chartered Banker Institute} Website: www\textit{.charteredbanker.com} The professional body for bankers — offering qualifications in banking, risk management, and compliance.}
\paragraph{\textbf{International Association of Privacy Professionals (IAPP)} Website: \textit{www.iapp.org} The world's largest global information privacy community — offering qualifications including the CIPP, CIPM, and CIPT designations.}

\section*{M.2 Key Publications and Further Reading}
\paragraph{\textbf{On Risk Management Foundations}.}
\begin{itemize}
  \item{Kaplan, R.S. and Mikes, A. (2012) 'Managing Risks: A New Framework', Harvard Business Review, May–June 2012.}
  \item{Power, M. (2007) Organized Uncertainty: Designing a World of Risk Management, Oxford University Press.}
  \item{Renn, O. (2008) Risk Governance: Coping with Uncertainty in a Complex World, Earthscan.}
\end{itemize}

\paragraph{\textbf{On Uncertainty and Decision-Making}.}
\begin{itemize}
  \item{Taleb, N.N. (2007) The Black Swan: The Impact of the Highly Improbable, Random House.}
  \item{Knight, F.H. (1921) Risk, Uncertainty and Profit, Hart, Schaffner \& Marx.}
  \item{Kahneman, D. (2011) Thinking, Fast and Slow, Farrar, Straus and Giroux.}
  \item{Thaler, R.H. and Sunstein, C.R. (2008) Nudge: Improving Decisions About Health, Wealth, and Happiness, Yale University Press.}
\end{itemize}

\paragraph{\textbf{On Human Factors and Organisational Failure}.}
\begin{itemize}
  \item{Reason, J. (1990) Human Error, Cambridge University Press.}
  \item{Reason, J. (1997) Managing the Risks of Organizational Accidents, Ashgate.}
  \item{Vaughan, D. (1996) The Challenger Launch Decision: Risky Technology, Culture, and Deviance at NASA, University of Chicago Press.}
  \item{Weick, K.E. and Sutcliffe, K.M. (2007) Managing the Unexpected: Resilient Performance in an Age of Uncertainty, Jossey-Bass.}
\end{itemize}

\paragraph{\textbf{On Culture and Governance}.}
\begin{itemize}
  \item{Edmondson, A.C. (2018) The Fearless Organization: Creating Psychological Safety in the Workplace for Learning, Innovation, and Growth, Wiley.}
  \item{Senge, P.M. (1990) The Fifth Discipline: The Art and Practice of the Learning Organization, Doubleday.}
  \item{Janis, I.L. (1982) Groupthink: Psychological Studies of Policy Decisions and Fiascoes, Houghton Mifflin.}
\end{itemize}

\paragraph{\textbf{On Financial Risk}.}
\begin{itemize}
  \item{Hull, J.C. (2018) Risk Management and Financial Institutions, 5th edition, Wiley.}
  \item{McNeil, A.J., Frey, R. and Embrechts, P. (2015) Quantitative Risk Management: Concepts, Techniques and Tools, revised edition, Princeton University Press.}
\end{itemize}

\paragraph{\textbf{On Cybersecurity}.}
\begin{itemize}
  \item{Anderson, R. (2020) Security Engineering: A Guide to Building Dependable Distributed Systems, 3rd edition, Wiley.}
  \item{NCSC Cyber Security Guidance — \textit{www.ncsc.gov.uk/section/advice-guidance}.}
\end{itemize}

\paragraph{\textbf{On Sustainability and Climate Risk}.}
\begin{itemize}
  \item{Network for Greening the Financial System (NGFS) — Climate Scenarios for Central Banks and Supervisors.}
  \item{Task Force on Climate-related Financial Disclosures (TCFD) — Final Recommendations Report (2017, updated 2021).}
  \item{Task Force on Nature-related Financial Disclosures (TNFD) — TNFD Framework (2023).}
\end{itemize}

\paragraph{\textbf{On Strategic Risk}.}
\begin{itemize}
  \item{Christensen, C.M. (1997) The Innovator's Dilemma, Harvard Business School Press.}
  \item{Rumelt, R. (2011) Good Strategy, Bad Strategy: The Difference and Why It Matters, Crown Business}
  \item{Walker, W.E. et al. (2013) 'Defining Uncertainty: A Conceptual Basis for Uncertainty Management in Model-Based Decision Support', Integrated Assessment, 4(1).}
\end{itemize}

\section*{M.3 Key Regulatory Guidance Documents}
\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.5}
  \begin{tabularx}{\textwidth}{ | >{\raggedright\arraybackslash}p{0.35\textwidth} 
								| >{\raggedright\arraybackslash}p{0.15\textwidth} 
                                | >{\raggedright\arraybackslash}X | }
    \hline
    \textbf{Document} & \textbf{Publisher} & \textbf{Topic} \\
    \hline
    \textbf{Financial Crime Guide} & FCA & Comprehensive guidance on financial crime risk assessment and controls for financial services firms\\
    \hline
    \textbf{JMLSG Guidance} & JMLSG & Sector-specific AML guidance for UK financial services\\
    \hline
    \textbf{ICAAP and ILAAP Guidance} & PRA & Expectations for banks' capital and liquidity adequacy assessments\\
    \hline
    \textbf{SS2/21 — Outsourcing and Third Party Risk Management} & PRA / FCA & Supervisory expectations for outsourcing in financial services\\
    \hline
    \textbf{Operational Resilience Policy Statement} & PRA / FCA / BoE & Framework for operational resilience in financial services\\
    \hline
    \textbf{Consumer Duty Guidance} & FCA & Implementation guidance for the Consumer Duty\\
    \hline
    \textbf{Five Steps to Risk Assessment} & HSE & Practical guide to workplace risk assessment\\
    \hline
    \textbf{Managing for Health and Safety} & HSE & Comprehensive guidance on health and safety management\\
    \hline
    \textbf{GDPR Guidelines} & EDPB / ICO & Guidance on GDPR compliance including DPIAs and security requirements\\
    \hline
    \textbf{TCFD Implementation Guidance} & TCFD & Practical guidance on implementing TCFD recommendations\\
    \hline
  \end{tabularx}
  \caption{Regulatory Guidance Documents}
\end{table}